### 2011-10-28 inst/doc/manual.Rnw
Added article and manual Rnw files

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Danny Arends
% The qcl mapping routine
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[12pt]{article}

\usepackage{times}
\usepackage{amsmath}
\usepackage{url}
\usepackage{color}

% revise margins
\setlength{\headheight}{0.0in}
\setlength{\topmargin}{-0.25in}
\setlength{\headsep}{0.0in}
\setlength{\textheight}{9.00in}
\setlength{\footskip}{0.5in}
\setlength{\oddsidemargin}{0in}
\setlength{\evensidemargin}{0in}
\setlength{\textwidth}{6.5in}

\setlength{\parskip}{6pt}
\setlength{\parindent}{0pt}

\newcommand{\code}{\texttt}
\newcommand{\lod}{\text{LOD}}

% make "Figure" within figure captions a small font
\renewcommand{\figurename}{\small Figure}

\begin{document}

\SweaveOpts{prefix.string=Figs/fig}
\setkeys{Gin}{width=\textwidth} %% <- change width of figures
% Try to get the R code from running into the margin
<<echo=FALSE>>=
options(width=87, digits=3, scipen=4)
set.seed(61777369)
@

% Change S input/output font size
\DefineVerbatimEnvironment{Sinput}{Verbatim}{fontsize=\footnotesize, baselinestretch=0.75, formatcom = {\color[rgb]{0, 0, 0.56}}}
\DefineVerbatimEnvironment{Soutput}{Verbatim}{fontsize=\footnotesize, baselinestretch=0.75, formatcom = {\color[rgb]{0.56, 0, 0}}}

\begin{center}
\vspace*{1in}
\textbf{\Large QCL mapping in R }

\bigskip \bigskip 
\bigskip \bigskip

{\large Danny Arends, and Ritsert C. Jansen}

\bigskip \bigskip

{\color{blue} \tt \bfseries University of Groningen \\
Groningen Bioinformatics centre \& GCC \\
Technical Report \# 214 \\[24pt]
XX Okt 2011}

\end{center}

\bigskip \bigskip

\textbf{\sffamily Abstract}: Introduction into quanitative correlation 
locus (QCL) mapping using the R package qcl
\vfill

\noindent \emph{email}: \verb|Danny.Arends@Gmail.com|
\newpage

\textbf{\sffamily Introduction}
Some introduction text

\bigskip

\textbf{\sffamily Format and load your data}
\nopagebreak

Loading your data

\bigskip

\textbf{\sffamily QCLscan and QCLscan.network}
\nopagebreak

Explains how to map QCL using the QCLscan and QCLscan.network functions.

\bigskip
\textbf{\sffamily Significance using QCLpermute}
\nopagebreak

Explains how to determine QCL significance using the QCLpermute function.

\bigskip
\textbf{\sffamily Advanced plots and QCL networks}
\nopagebreak

Some more advanced topics

\newpage
{\small
\noindent \textbf{\sffamily Package qcl function overview} \\
\noindent \begin{tabular}{ll} 
\hspace*{25mm} & \hspace*{103mm} \\ \hline
QCLscan & Genome wide scan for QCL \\
QCLscan.network & Scan for QCL using start gene(s) \\
QCLpermute & Significance thresholds for QCL by permutation \\
plot.QCL & Plot QCL profiles at increasing cut-offs \\
image.QCL & Image a multiple phenotype scan \\
\hline
\end{tabular}
}

\end{document}

### 2011-10-28 R/qcl.scan.R

@@ -41,8 +41,8 @@ QCLscan.cross <- function(cross, pheno.col, verbose = FALSE){
   if(missing(cross)) stop("cross is missing")
   if(.has_rqtl){
     require(qtl)
-    if(missing(pheno.col)) pheno.col <- 1:ncol(phenotypes)
     phenotypes <- apply(pull.pheno(cross),2,as.numeric)
+    if(missing(pheno.col)) pheno.col <- 1:ncol(phenotypes)
     genotypes <- pull.geno(cross)
     QCLscan(genotypes, phenotypes,pheno.col=pheno.col,verbose=verbose)
   }else{

### 2011-10-28 inst/doc/article.Rnw
Added article and manual Rnw files

--- 2011-10-28 inst/doc/manual.Rnw
@@
 \vspace*{1in}
-\textbf{\Large QCL mapping in R }
+\textbf{\Large QCL mapping in experimental crosses}
 
@@
 
-\textbf{\sffamily Abstract}: Introduction into quanitative correlation 
-locus (QCL) mapping using the R package qcl
+\textbf{\sffamily Abstract}: Quanitative correlation locus (QCL) mapping is 
+a novel approach to detect genetic regulation of phenotypes in recombinant 
+inbred line populations (RIL)
 \vfill
@@
 \textbf{\sffamily Introduction}
+
 Some introduction text
@@
 
-\textbf{\sffamily Format and load your data}
+\textbf{\sffamily Methodology}
 \nopagebreak
 
-Loading your data
+Some methodology text
 
@@
 
-\textbf{\sffamily QCLscan and QCLscan.network}
+\textbf{\sffamily Results}
 \nopagebreak
 
-Explains how to map QCL using the QCLscan and QCLscan.network functions.
+Some resulting text
 
 \bigskip
-\textbf{\sffamily Significance using QCLpermute}
+\textbf{\sffamily Discussion}
 \nopagebreak
 
-Explains how to determine QCL significance using the QCLpermute function.
-
-\bigskip
-\textbf{\sffamily Advanced plots and QCL networks}
-\nopagebreak
-
-Some more advanced topics
+Some discussion text
 

### 2011-10-28 inst/doc/article.Rnw
minor
Article updates

@@
 \vspace*{1in}
-\textbf{\Large QCL mapping in experimental crosses}
+\textbf{\Large QCL: the added value of correlation }
 
@@
 
-{\large Danny Arends, and Ritsert C. Jansen}
+{\large Danny Arends$^{1,2}$ and Ritsert C. Jansen$^{1}$}
 
@@
 
-{\color{blue} \tt \bfseries University of Groningen \\
-Groningen Bioinformatics centre \& GCC \\
-Technical Report \# 214 \\[24pt]
-XX Okt 2011}
+{\color{blue} \tt \bfseries $^{1}$ Groningen Bioinformatics Centre - University of Groningen \\
+$^{2}$ Genomics Coordination Centre \& University Medical Centre Groningen \\
+version: \# 0.0.1\\[24pt]
+received on XX XXX 201X\\
+accepted on XX XXX 201X\\
+}
 
@@
 
-\textbf{\sffamily Abstract}: Quanitative correlation locus (QCL) mapping is 
-a novel approach to detect genetic regulation of phenotypes in recombinant 
-inbred line populations (RIL)
+\textbf{\sffamily Abstract}: In this paper we present Quanitative Correlation Locus 
+(QCL) mapping a novel approach to detect genetic regulation of phenotypes in 
+recombinant inbred line populations (RIL) or other experimental crosses. QCL mapping 
+allows association of phenotype correlation differences to genetic markers, in almost 
+the same way as QTL mapping associates genetic markers with differences in expression.
+We use QCL mapping to reconstruct underlying genetic architecture, and provide insight 
+into regulation of (gene) expression. QCL information in combination with classical QTL 
+information leads to a more detailed insight into the genetic architecture underlying 
+complex traits. 
+In two example we reconstruct part of the genetic architecture in S. Cerevisiae grown 
+under ethanol conditions. The other example shows the reconstruction of the chain 
+elongation pathway in A. Thaliana. In both examples we map the observe differences 
+in correlation to genetic markers, and show the biological relevance of QCL peeks and 
+hotspots in regulatory network reconstruction.
+
 \vfill
 
-\noindent \emph{email}: \verb|Danny.Arends@Gmail.com|
+\noindent Corresponding Author: \verb|Danny.Arends@Gmail.com|
+\noindent Associate Editor: XXXX XXXX
 \newpage
@@
 
-Some introduction text
+Genetical genomics experiments have shown that gene expression variation can be mapped to 
+genotypic variation (Jansen and Nap, 2001). Quantitative Trait Locus (QTL) mapping of a 
+gene identifies regions in the genome for which different genotypes lead to differential 
+expression level of the gene. This QTL mapping can be done on all molecular levels including, 
+gene expression (eQTL), protein abundance (pQTL) and metabolites (mQTL).
+Recently different methods have been proposed that allow studying, in addition to the genes 
+differentially expressed between conditions, the genes that are differentially coexpressed 
+between conditions (Tesson et al. 2010, Kompass et al. 2011). Such differential co-expression 
+is taken as another level of information from genomic data. This level of information can for 
+example be used to expose transcriptional network rewiring (Langfelder et al. 2008). Differential 
+co-expression has been observed in studies of different diseases, including between healthy and 
+cancer tissue studies. In parallel, correlations between phenotypes are studied for various 
+reasons e.g. discovering and detecting batch effects, marker assisted breeding and causal 
+inference. The importance of identifying differences in correlation has been increasingly 
+recognized and this led to the development of this algorithm to perform QCL analysis in 
+experimental crosses.
+
+We proposed to map not differential co-expression but differential correlation to genetic 
+variation, i.e. to identify regions in the genome for which one genotype value leads to 
+correlated expression between a start gene and all other genes, exploiting the fact that any 
+cis-QTL will induce variation which will in turn lead to correlation differences among 
+individuals carrying the other genotype. We merge information comming from QCL and QTL to reveal 
+underlying genetic architecture, which would remain hidden when using only classical QTL information.
+
+We urge all QTL researchers to combine QCL analysis with existing QTL analysis to expose 
+transcriptional networks involved in the regulation of complex phenotypes. For this we 
+provide an R package with a reference implementation  of the QCL algorithm
 
@@
 
-Some methodology text
+We show two examples in which we reconstruct the genetic architecture using QCL mapping. 
+Both datasets are available inside the R package, as well as the code to analyse these 
+two datasets.
+
+\emph{\sffamily Dataset Example 1. A. Thaliana}
+
+Metabolite data from an A. Thaliana Landsberg erecta (Ler) and Cape Verde Islands (Cvi) 
+dataset from R/qtl, 24 characterized metabolites which are the products of chain elongation. 
+There are 162	recombinant inbred lines genotyped at 117 markers on 5 chromosomes. Additional 
+details about the experimental settings are available in the original paper (Keurentjes et al. 2006,
+Alonso-Blanco et al. 1998)
+
+\emph{\sffamily Dataset Example 2. S. Cerevisiae}
+
+S. Cerevisiae dataset from the gene expression omnibus (GEO) (GDS1115 / GDS1116). This 
+dataset contains expression profiles of parental strains and haploid progenies from a 
+cross of strain BY4716 and the wild wine strain RM11-1a. Data was collected for 4482 
+probes. Genotype data for the 109 progenies was available at 282 genetic markers 
+obtained by contacting the authors of the original paper. Expression of probes was 
+measured under growth conditions containing 1% ethanol. Additional details about the 
+experimental settings are available in the original paper (Brem et al. 2005).
+
+\emph{\sffamily Preprocessing}
+
+Probes showing little to no variance (< 0.01) were removed from the analysis. Batch 
+effects due to different hybridization dates were estimated and removed from the data 
+by using a mean centering approach (Group means were estimated and corrected).
+
+\textbf{\sffamily Algorithm: eQCL mapping}
+
+Using recombinant inbred lines we introduce natural variation into a population 
+detectable as cis-QTL in a genome wide QTL scan. The effect of any such genetic 
+variation is transfered downstream to interaction partners. When this leads to 
+changes in mean expression of one of the partners we can also detect this using 
+classical QTL methods (trans/distant QTL).
+
+However QTL mapping suffers from two major drawbacks (p1) the interaction partners 
+of genes are obscured by many phenotypes co-localizing at a single genetic marker, and 
+(p2) due to buffering the effect of any variation will 'fade' quickly and not be 
+detectable as the difference in mean phenotype in the population does not change 
+significantly.
+
+QCL mapping however uses this buffering to detect interaction partners. The hypothesis 
+is that when we split the population based on genotype at the cis-marker, related phenotypes 
+will loosing correlation with the gene under investigation. This allows us to not only 
+only identify its interaction partners. QCL mapping is thus not suffering from (p1) many 
+phenotypes co-localizing). Additionally it is more sensitive to detect variation caused by 
+regulators genes because variation of these genes causes many 'downstream' phenotypes 
+to show differential correlation with the inducing phenotype, because QCL relies on 
+correlation changes as opposed to differences in phenotype mean QCL is unaffected by (p2).
+
+\emph{\sffamily Step 1 Choosing a start/seed probe}
+
+Choose a phenotype to subject to QCL analysis, this can be your favourite gene, 
+protein, metabolite or classical phenotype.
+
+\emph{\sffamily Step 2 Single phenotype QCL}
+
+Using genotype data we can split individuals at each marker in two groups (A or B).
+At each marker calculate the correlation between our phenotype and all other 
+phenotypes in groups A and B. Collect all the result vectors into two matrices A 
+and B, and calculate the QCL matrix by taking the squared difference between matrix A 
+and B.
+
+\emph{\sffamily Step 2a Significance testing}
+
+The threshold for a significant difference in QCL is assessed by using permutations, in 
+each round the link between genotype and phenotype is broken, by redistributing at random 
+genotypes amongst individuals not allowing for duplicates. The QCL on these new 
+‘individuals’ allows for estimation of the underlying null-distribution of QCL scores.
+
+\emph{\sffamily Step 2b Create single trait QCL profiles}
+
+\emph{\sffamily Step 3 Network scan for QCL}
+
+Using the user defined seed gene we can reconstruct the interaction network, we 
+first do a single phenotype QCL scan (step 2), then using a threshold we select 
+phenotypes which show a significant change in QCL.
+
+Two things can happen: 1) No genes show a significant change in QCL, which means we 
+cannot find any interaction partners for this phenotype, or 2) This genes has an QCL 
+with other genes, then the algorithm adds these genes onto a list to scan and repeats 
+the single phenotype QCL scan (step 2) for these phenotypes (not allowing for duplicate 
+scans). This is repeated until no more significant QCL are found or a user defined 
+search depth is reached.
+
+\emph{\sffamily Step 3a Visualizing the Phenotypes x Marker relations}
+
+When plotting multiple QCL profiles as a heatmap, multiple QCL are seen mapping to the 
+same genetic location. At these locations, we expect to find one (or more) of the genes 
+to show a cis-QTL, this phenotype (or phenotypes) introduced genetic variation, which is 
+seen also (and more pronounced) in a QTL scan.
+
+\emph{\sffamily Step 3b Visualizing the Phenotypes x Phenotypes relations}
+
+\emph{\sffamily Step 4 Network reconstruction}
+
+So how do we build up gene interaction networks, This looks complicated but the information 
+is implicitely stored in the Phenotype x Marker and Phenotype x Phentype relations matrices. 
+By using network scanning we only identify part of the network in which our phenotype is 
+active and closely related interaction partners. To identify a pathway or main regulatory phenotypes 
+controlling our phenotype we need to combine this information.
 
 \bigskip
-
 \textbf{\sffamily Results}
@@
 
-Some discussion text
+Microarray experiments have allowed the identification of genes differentially 
+expressed between various conditions, or groups of genes coexpressed across 
+samples. Many genetical genomics screens have been carried out to map differential 
+expression to genetic variation. Screening the genome for genotype combinations 
+for which there is differential expression has been a successful strategy when 
+mapping Mendelian traits. Results on complex traits have been however less then 
+satisfactory with only small amount of heritability explained by individuals loci. 
+QCL mapping however 'borrows' information from each phenotype in the analysis set, 
+and enables easy extraction of phenotype marker phenotype networks. furthermore QCL 
+seem to 'transfer' between phenotypes in a more stabile fashion which gives greater 
+detail and confidence when reconstructing genetic architecture.
+
+\newpage
+\textbf{\sffamily Bibliography}
+
+Langfelder P, Horvath S (2008) WGCNA: an R package for weighted correlation network analysis. BMC Bioinformatics 2008, 9:559
+
+Jansen R. C, and Nap J. P. (2001) Genetical Genomics: the added value from seggregation. 
+\emph{Trends Genet}, 17(7), 388-391.
+
+Tesson, B. M., Breitling R. and Jansen, R. C. (2010). DiffCoEx: a simple and sensitive method 
+to find differentially coexpressed gene modules. \emph{BMC Bioinformatics}, 11,497-506.
+
+Kompass, K. and Witte, J. (2011) - Co-regulatory expression quantitative trait loci mapping: 
+method and application to endometrial cancer. \emph{BMC Medical Genomics}, 4, 6.
+
+Broman, K.W., Wu, H., Sen, S. and Churchill, G.A. (2003) - R/qtl: QTL mapping in experimental 
+crosses. \emph{Bioinformatics}, 19, 889-890.
+
+Arends, D., Prins, P., Jansen, R.C. and Broman, K.W. (2010) - R/qtl: high-throughput multiple 
+QTL mapping. \emph{Bioinformatics}, 26, 2990-2992.
+
+R Development Core Team (2010) - R: A Language and Environment for Statistical Computing. 
+http://www.R-project.org, ISBN: 3-900051-07-0
+
+Edgar R., Domrachev M., Lash A.E. (2002) - Gene Expression Omnibus: NCBI gene expression and 
+hybridization array data repository. \emph{Nucleic Acids Res.}, 30,207-210.
+
+Keurentjes, J. J., Fu, J., de Vos, C. H., Lommen, A., Hall, R. D., Bino, R. J., van der Plas, 
+L. H., Jansen, R. C., Vreugdenhil, D., Koornneef, M. (2006) - The genetics of plant metabolism. 
+\emph{Nature Genetics.}, 38(7), 842-849.
+
+Alonso-Blanco, C., Peeters, A. J., Koornneef, M., Lister, C., Dean, C., van den Bosch, N., Pot, 
+J., Kuiper, M. T. (1998) - Development of an AFLP based linkage map of Ler, Col and Cvi 
+Arabidopsis thaliana ecotypes and construction of a Ler/Cvi recombinant inbred line population. 
+\emph{Plant J.}, 14, 259-271.
+
+Brem RB, Storey JD, Whittle J, Kruglyak L. (2005) - Genetic interactions between polymorphisms 
+that affect gene expression in yeast. \emph{Nature}, 436(7051),701-703.
 

### 2011-11-03 R/correlations.R
R-2.14.0 updates

@@ -12,7 +12,7 @@ correlation <- function(x, y, verbose = FALSE){
   if(is.matrix(x) && dim(x)[2] > 1){
     return(matrix_correlation(x,verbose=verbose))
   }else{
-    if(!.has_d){
+    if(!get(".has_d", envir = .QclEnv)){
       if(verbose) cat(.has_d_warnmsg)
       return(cor(x,y))
     }else{
@@ -33,7 +33,7 @@ correlation <- function(x, y, verbose = FALSE){
 }
 
 matrix_correlation <- function(x, verbose = FALSE){
-  if(!.has_d){
+  if(!get(".has_d", envir = .QclEnv)){
     if(verbose) cat(.has_d_warnmsg)
     return(cor(x))
   }else{

### 2011-11-03 R/qcl.plot.R

@@ -8,7 +8,7 @@
 # Plotting routines for QCL analysis
 #
 
-plot.QCLscan <- function(x, pheno.col = 1, qcl.threshold =0.3, ...){
+plot.QCLscan <- function(x, pheno.col = 1, qcl.threshold =0.3, do.legend=TRUE, ...){
   npheno <- length(x)
   if(pheno.col > npheno) stop("No such phenotype")
   pname <- attr(x[[pheno.col]],"name") 
@@ -22,5 +22,7 @@ plot.QCLscan <- function(x, pheno.col = 1, qcl.threshold =0.3, ...){
     points(QCLprofiles(x,t)[pname,], lwd=2,col=rgb((1/max(x[[pheno.col]]))*t,0,0),type='l')
     colorz <- c(colorz,rgb((1/max(x[[pheno.col]]))*t,0,0))
   }
-  legend("topleft",paste("Threshold =",seq(qcl.threshold,max(x[[pheno.col]]),0.05)),lwd=2,col=colorz)
+  if(do.legend){
+    legend("topleft",paste("Threshold =",seq(qcl.threshold,max(x[[pheno.col]]),0.05)),lwd=2,col=colorz)
+  }
 }

### 2011-11-03 R/qcl.scan.network.R

@@ -46,13 +46,13 @@ QCLscan.network <- function(genotypes, phenotypes, pheno.col, qcl.threshold = 0.
 QCLscan.network.cross <- function(cross, pheno.col, qcl.threshold = 0.4, max.depth=10, verbose = TRUE){
   if(missing(cross)) stop("cross is missing")
   if(missing(pheno.col)) stop("pheno.col missing")
-  if(.has_rqtl){
+  if(get(".has_rqtl", envir = .QclEnv)){
     require(qtl)
-    phenotypes <- apply(pull.pheno(cross),2,as.numeric)
-    genotypes <- pull.geno(cross)  
+    phenotypes <- apply(qtl::pull.pheno(cross),2,as.numeric)
+    genotypes <- qtl::pull.geno(cross)  
     QCLscan.network(genotypes, phenotypes, pheno.col=pheno.col, 
                     qcl.threshold=qcl.threshold,max.depth=max.depth,verbose=verbose)
   }else{
-    stop(.has_rqtl_warnmsg)
+    warning(.has_rqtl_warnmsg)
   }
 }

### 2011-11-03 R/qcl.image.R
R-2.14.0 updates

@@ -14,9 +14,9 @@ image.QCLscan <- function(x, qcl.threshold = 0.35, against = c("markers","phenot
   if(!is.null(mymatrix)){ 
     image(1:ncol(mymatrix),1:nrow(mymatrix),t(mymatrix),
           main=paste("QCLs at",qcl.threshold,"phenotypes vs",against[1]),
-          yaxt="n",xaxt="n",ylab="", xlab="",col=colorrange)
-    axis(2,rownames(mymatrix),at=1:nrow(mymatrix),las=2,cex.axis=0.7)
-    axis(1,colnames(mymatrix),at=1:ncol(mymatrix),las=2,cex.axis=0.7)
+          yaxt="n",xaxt="n",ylab="", xlab="",col=colorrange,cex.main=0.7)
+    axis(2,rownames(mymatrix),at=1:nrow(mymatrix),las=2,cex.axis=0.5)
+    axis(1,colnames(mymatrix),at=1:ncol(mymatrix),las=2,cex.axis=0.3)
     if(do.grid){
       abline(h=seq(-0.5,nrow(mymatrix)+0.5,1),col=grid.col,lwd=1)
       abline(v=seq(-0.5,ncol(mymatrix)+0.5,1),col=grid.col,lwd=1)

### 2011-11-03 R/qcl.kegg.R

@@ -25,7 +25,7 @@ KEGG.pathway.parser <- function(HTMLdata, verbose = FALSE){
 }
 
 download.KEGG.annotation <- function(genenames, species = "sce", directory = "tmp_kegg", force.download=FALSE, verbose = FALSE){
-  if(.has_rcurl){
+  if(get(".has_rcurl", envir = .QclEnv)){
     require("RCurl")
     if(missing(genenames)) stop("No gene names to get annotation for supplied")
     cnt <- 1
@@ -40,7 +40,7 @@ download.KEGG.annotation <- function(genenames, species = "sce", directory = "tm
         if(verbose) cat("Loaded ",cnt,":",gene_name," from ",file_out_name,"\n",sep="")
       }else{
         url <- paste("http://www.kegg.jp/entry/",species,"/",gene_name,sep="")
-        HTMLdata <- getURL(url)
+        HTMLdata <- RCurl::getURL(url)
         cat(HTMLdata,file=paste(directory,"/kegg_",gene_name,".txt",sep=""))
         Sys.sleep(runif(1)*2)
         if(verbose) cat("Downloaded ",cnt,":",gene_name," from ",url,"\n",sep="")
@@ -62,6 +62,6 @@ download.KEGG.annotation <- function(genenames, species = "sce", directory = "tm
     cat("Pathways found for ",p_cnt,"/",cnt," genes\n",sep="")
     invisible(annotations)
   }else{
-    stop(.has_rcurl_warnmsg)
+    warning(.has_rcurl_warnmsg)
   }
 }

### 2011-11-03 R/qcl.permutation.R

@@ -61,12 +61,12 @@ significance.QCLHotSpot <- function(QCLpermute,significant=212){
 QCLpermute.cross <- function(cross, pheno.col, n.perm=10, directory="permutations", verbose=TRUE, ...){
   if(missing(cross)) stop("cross is missing")
   if(missing(pheno.col)) stop("pheno.col missing")
-  if(.has_rqtl){
+  if(get(".has_rqtl", envir = .QclEnv)){
     require(qtl)
-    phenotypes <- apply(pull.pheno(cross),2,as.numeric)
-    genotypes <- pull.geno(cross)
+    phenotypes <- apply(qtl::pull.pheno(cross),2,as.numeric)
+    genotypes <- qtl::pull.geno(cross)
     QCLpermute(genotypes, phenotypes,pheno.col=pheno.col,n.perm=n.perm,directory=directory,verbose=verbose)
   }else{
-    stop(.has_rqtl_warnmsg)
+    warning(.has_rqtl_warnmsg)
   }
 }

### 2011-11-03 R/qcl.scan.R

@@ -39,13 +39,13 @@ QCLscan <- function(genotypes, phenotypes, pheno.col = 1:ncol(phenotypes), verbo
 #-- R/qtl interface --#
 QCLscan.cross <- function(cross, pheno.col, verbose = FALSE){
   if(missing(cross)) stop("cross is missing")
-  if(.has_rqtl){
+  if(get(".has_rqtl", envir = .QclEnv)){
     require(qtl)
-    phenotypes <- apply(pull.pheno(cross),2,as.numeric)
+    phenotypes <- apply(qtl::pull.pheno(cross),2,as.numeric)
     if(missing(pheno.col)) pheno.col <- 1:ncol(phenotypes)
-    genotypes <- pull.geno(cross)
+    genotypes <- qtl::pull.geno(cross)
     QCLscan(genotypes, phenotypes,pheno.col=pheno.col,verbose=verbose)
   }else{
-    stop(.has_rqtl_warnmsg)
+    warning(.has_rqtl_warnmsg)
   }
 }

### 2011-11-03 R/qcl.utils.R

@@ -5,56 +5,67 @@
 # last modified Jun, 2011
 # first written nov, 2010
 # 
-#
+# nchr
 
 #Change any list of lodscores into a scanone object (only pre-req: length(lodscores)==sum(nmar(cross))
 lodscorestoscanone <- function(cross,lodscores,traitnames = NULL){
-  n <- unlist(lapply(FUN=names,pull.map(cross)))
-  chr <- NULL
-  if(!is.null(ncol(pull.map(cross)[[1]]))){
-    d <- as.numeric(unlist(lapply(pull.map(cross),FUN=function(x) {x[1,]})))
-    for(i in 1:nchr(cross)){
-      chr <- c(chr,rep(names(cross$geno)[i], ncol(pull.map(cross)[[i]])))
+  if(get(".has_rqtl", envir = .QclEnv)){
+    require(qtl)
+    mymap <- qtl::pull.map(cross)
+    n <- unlist(lapply(FUN=names,mymap))
+    chr <- NULL
+    if(!is.null(ncol(mymap[[1]]))){
+      d <- as.numeric(unlist(lapply(mymap,FUN=function(x) {x[1,]})))
+      for(i in 1:qtl::nchr(cross)){
+        chr <- c(chr,rep(names(cross$geno)[i], ncol(mymap[[i]])))
+      }
+    }else{
+      d <- as.numeric(unlist(mymap))
+      for(i in 1:qtl::nchr(cross)){
+        chr <- c(chr,rep(names(cross$geno)[i], length(mymap[[i]])))
+      }
     }
+    qtlprofile <- cbind(chr,d,lodscores)
+    qtlprofile <- as.data.frame(qtlprofile)
+    qtlprofile[,1] <- chr
+    qtlprofile[,2] <- as.numeric(d)
+    if(!is.null(ncol(lodscores))){
+      for(x in 1:ncol(lodscores)){
+        qtlprofile[,2+x] <- as.numeric(lodscores[,x])
+      }
+      traitnames = paste("lod",1:ncol(lodscores))
+    }else{
+       qtlprofile[,3] <- as.numeric(lodscores)
+       traitnames = "lod"
+    }
+    rownames(qtlprofile) <- n
+    colnames(qtlprofile) <- c("chr","cM",traitnames)
+    class(qtlprofile) <- c("scanone", "data.frame")
+    qtlprofile
   }else{
-    d <- as.numeric(unlist(pull.map(cross)))
-    for(i in 1:nchr(cross)){
[…]

### 2011-11-03 R/zzz.R

@@ -9,10 +9,12 @@
 #
 
 #Global package variables
-.has_d               <- TRUE
-.has_snow            <- FALSE
-.has_rcurl           <- FALSE
-.has_rqtl            <- FALSE
+.QclEnv <- new.env()
+assign(".has_d",     TRUE, envir = .QclEnv)
+assign(".has_snow",  FALSE, envir = .QclEnv)
+assign(".has_rcurl", FALSE, envir = .QclEnv)
+assign(".has_rqtl",  FALSE, envir = .QclEnv)
+
 
 .has_d_warnmsg       <- "- D not available, using standard R/C/C++ functionality\n"
 .has_snow_warnmsg    <- "- Package: SNOW not found, multi CPU support is not available/disabled\n"
@@ -20,22 +22,29 @@
 .has_rqtl_warnmsg    <- "- Package: qtl not found, R/qtl functionality not available/disabled\n"
 
 #Package loading
-.First.lib <- function(lib, pkg){
- cat("- Loading package qcl\n")
- library.dynam("qcl", pkg, lib)
- tryCatch(
-   library.dynam("Dcode", pkg, lib),
-   error = function(e){
+.onAttach <- function(lib, pkg){
+  packageStartupMessage("- Loading package qcl\n")
+  .has_d <- TRUE
+  .has_snow <- FALSE
+  .has_rcurl <- FALSE
+  .has_rqtl <- FALSE
+  library.dynam("qcl", pkg, lib)
+  tryCatch(
+    library.dynam("Dcode", pkg, lib),
+    error = function(e){
      .has_d <<- FALSE
    })
   .has_snow  <<- ("snow" %in% installed.packages()[,1])
   .has_rcurl <<- ("RCurl" %in% installed.packages()[,1])
   .has_rqtl  <<- ("qtl" %in% installed.packages()[,1])
-  
-  if(!.has_d)     cat(.has_d_warnmsg)
-  if(!.has_snow)  cat(.has_snow_warnmsg)
-  if(!.has_rcurl) cat(.has_rcurl_warnmsg)
-  if(!.has_rqtl)  cat(.has_rqtl_warnmsg)
+  assign(".has_d",     .has_d,     envir = .QclEnv)
+  assign(".has_snow",  .has_snow,  envir = .QclEnv)
+  assign(".has_rcurl", .has_rcurl, envir = .QclEnv)
[…]

### 2011-11-03 inst/doc/article.Rnw
R-2.14.0 updates

@@
 % The qcl mapping routine
+%\VignetteIndexEntry{Article}
 %
@@
 options(width=87, digits=3, scipen=4)
+@
+
+<<echo=FALSE,results=hide>>=
+  require(qcl)
   set.seed(61777369)
+  data(ath.metabolites) # Arabidopsis Thaliana dataset
+  qcl_result <- QCLscan(ath.metab$genotypes, ath.metab$phenotypes)
 @
@@
 genotypes amongst individuals not allowing for duplicates. The QCL on these new 
-‘individuals’ allows for estimation of the underlying null-distribution of QCL scores.
+?ndividuals?allows for estimation of the underlying null-distribution of QCL scores.
 
 \emph{\sffamily Step 2b Create single trait QCL profiles}
+
+Single trait QCL profiles can be obtained by setting a significance threshold.
+
+\begin{figure}
+\centering
+<<singletrait, fig=TRUE, echo=FALSE, height=3.5>>=
+plot(qcl_result, 2, do.legend=FALSE)
+@ 
+\caption{Single trait QCL profile (A. Thaliana).\label{fig:singletrait}}
+\end{figure}
 
@@
 
-Some resulting text
+Global overview of QCL scan after using a full scan for QCL
+<<echo=FALSE>>=
+  qcl_result
+@ 
+Single results from a QCL mapping, shows the pattern at the highest threshold.
+<<echo=FALSE>>=
+  qcl_result[[2]]
+@ 
+Phenotype x Marker image plot, this is 'comparable' to a QTL scan on mutliple traits.
+
+\begin{figure}[ht]
+\centering
+<<p2mmatrix, fig=TRUE, echo=FALSE, height=3.5>>=
+par(mar=c(3,8,1.1,1.1))
+image(qcl_result, 0.5, against="markers",do.grid=FALSE)
+@ 
+\caption{Phenotypes x Markers matrix (A. Thaliana).\label{fig:p2mmatrix}}
+\end{figure}
+
+Phenotype x Phenotype image plot, the new network information obtained by QCL mapping, 
+which can be used to reconstruct the underlying genetic architecture.
+\begin{figure}[ht]
+\centering
+<<p2pmatrix, fig=TRUE, echo=FALSE, height=3.5>>=
+par(mar=c(3,8,1.1,1.1))
+image(qcl_result, 0.5, against="phenotypes")
+@ 
+\caption{Phenotypes x Phenotypes matrix (A. Thaliana).\label{fig:p2pmatrix}}
+\end{figure}
 

### 2011-11-07 R/zzz.R
Article and internals

@@ -11,6 +11,7 @@
 #Global package variables
 .QclEnv <- new.env()
 
+get_QclEnv <- function(){.QclEnv}
 has_d <- function(){ get(".has_d", envir = .QclEnv) }
 has_snow <- function(){ get(".has_snow", envir = .QclEnv) }
 has_rcurl <- function(){ get(".has_rcurl", envir = .QclEnv) }
@@ -23,10 +24,6 @@ has_rqtl <- function(){ get(".has_rqtl", envir = .QclEnv) }
 
 #Package loading
 .onAttach <- function(lib, pkg){
-  assign(".has_d",     TRUE, envir = .QclEnv)
-  assign(".has_snow",  FALSE, envir = .QclEnv)
-  assign(".has_rcurl", FALSE, envir = .QclEnv)
-  assign(".has_rqtl",  FALSE, envir = .QclEnv)
   packageStartupMessage("- Loading package qcl\n", appendLF = FALSE)
   .has_d <- TRUE
   .has_snow <- FALSE
@@ -36,7 +33,7 @@ has_rqtl <- function(){ get(".has_rqtl", envir = .QclEnv) }
   tryCatch(
     library.dynam("Dcode", pkg, lib),
     error = function(e){
-     .has_d <- FALSE
+     .has_d <<- FALSE
    })
   .has_snow  <- ("snow" %in% installed.packages()[,1])
   .has_rcurl <- ("RCurl" %in% installed.packages()[,1])

### 2011-11-07 inst/doc/article.Rnw
Article and internals

@@
 
-\SweaveOpts{prefix.string=Figs/fig}
+\SweaveOpts{prefix.string=fig}
 \setkeys{Gin}{width=\textwidth} %% <- change width of figures
@@
 \vspace*{1in}
-\textbf{\Large QCL: the added value of correlation }
+\textbf{\Large QCL: Added value of differential correlation }
 
@@
 \textbf{\sffamily Abstract}: In this paper we present Quanitative Correlation Locus 
-(QCL) mapping a novel approach to detect genetic regulation of phenotypes in 
+(QCL) mapping, a novel approach to detect genetic regulation of phenotypes in 
 recombinant inbred line populations (RIL) or other experimental crosses. QCL mapping 
@@
 gene expression (eQTL), protein abundance (pQTL) and metabolites (mQTL).
+
 Recently different methods have been proposed that allow studying, in addition to the genes 
@@
 
-\textbf{\sffamily Algorithm: eQCL mapping}
+\textbf{\sffamily Algorithm: QCL mapping}
 
@@
 
-\emph{\sffamily Step 1 Choosing a start/seed probe}
+\emph{\sffamily Choosing a start/seed probe (Optional)}
 
@@
 
-\emph{\sffamily Step 2 Single phenotype QCL}
+\emph{\sffamily Single phenotype QCL}
 
@@
 
-\emph{\sffamily Step 2a Significance testing}
+\emph{\sffamily Significance testing}
 
@@
 genotypes amongst individuals not allowing for duplicates. The QCL on these new 
-?ndividuals?allows for estimation of the underlying null-distribution of QCL scores.
+individuals allows for estimation of the underlying null-distribution of QCL scores. 
+See figures XX, and XX showing the score distribution in permuted data.
 
-\emph{\sffamily Step 2b Create single trait QCL profiles}
+\emph{\sffamily Creating single trait QCL profiles}
 
-Single trait QCL profiles can be obtained by setting a significance threshold.
+After permutation single trait QCL profiles can be obtained by setting a qcl significance 
+threshold. This threshold defines when a change in correlation is significant, and at 
+each marker we can plot the count of number the other phenotypes that show a significant 
+difference in correlation. Overplotting chromosome, QTL information, and phenotype location 
+(if applicable) produces the single QCL plots seen in figures XX, and XX.
 
@@
 
-Using the user defined seed gene we can reconstruct the interaction network, we 
-first do a single phenotype QCL scan (step 2), then using a threshold we select 
-phenotypes which show a significant change in QCL.
+If we find any significant QCL information we can try to reconstruct the interaction 
+network of the phenotype. We first do a single phenotype QCL scan (step 2), then 
+using a permutation determin a qcl threshold, and select phenotypes which show a 
+significant change in QCL.
 
-Two things can happen: 1) No genes show a significant change in QCL, which means we 
-cannot find any interaction partners for this phenotype, or 2) This genes has an QCL 
-with other genes, then the algorithm adds these genes onto a list to scan and repeats 
-the single phenotype QCL scan (step 2) for these phenotypes (not allowing for duplicate 
-scans). This is repeated until no more significant QCL are found or a user defined 
-search depth is reached.
+Two things can happen: 
+1) No genes show a significant change in QCL, which means we 
+cannot find any interaction partners for this phenotype, or 
+2) This genes has an QCL with other genes, then the algorithm adds 
+these genes onto a list to scan and repeats the single phenotype 
+QCL scan (step 2) for these phenotypes (not allowing for duplicate 
+scans). This is repeated until no more significant QCL are found or 
+a user defined search depth is reached.
 
-\emph{\sffamily Step 3a Visualizing the Phenotypes x Marker relations}
+\emph{\sffamily Visualizing the Phenotypes x Marker relations}
 
@@
 same genetic location. At these locations, we expect to find one (or more) of the genes 
-to show a cis-QTL, this phenotype (or phenotypes) introduced genetic variation, which is 
-seen also (and more pronounced) in a QTL scan.
+to show a cis-QTL, this is because a QTL can induce a QCL, because variation of expression 
+in the phenotype is causing other phenotypes to become uncorrelated, when assigned to 
+groups determined by the QTL marker. This phenotype (or phenotypes) introduced genetic 
+variation, which is seen also (and more pronounced) in a QTL scan.
 
-\emph{\sffamily Step 3b Visualizing the Phenotypes x Phenotypes relations}
+\emph{\sffamily Visualizing the Phenotypes x Phenotypes relations}
 
-\emph{\sffamily Step 4 Network reconstruction}
+The additional information obtained by QCL mapping is summarized in the 
+Phenotypes x Phenotypes matrix. This shows the summed evidence of interaction 
+between the phenotypes showing QCL transmitted through all the markers and the phenotype scanned.
+
+\emph{\sffamily Network reconstruction}
 

### 2011-11-07 inst/doc/article.Rnw
Article

@@
 
-{\color{blue} \tt \bfseries $^{1}$ Groningen Bioinformatics Centre - University of Groningen \\
-$^{2}$ Genomics Coordination Centre \& University Medical Centre Groningen \\
+{\color{blue} \tt \bfseries $^{1}$ Groningen Bioinformatics Centre - University of Groningen\\
+$^{2}$ Genomics Coordination Centre - University Medical Centre Groningen\\
 version: \# 0.0.1\\[24pt]
@@
 complex traits. 
-In two example we reconstruct part of the genetic architecture in S. Cerevisiae grown 
-under ethanol conditions. The other example shows the reconstruction of the chain 
-elongation pathway in A. Thaliana. In both examples we map the observe differences 
-in correlation to genetic markers, and show the biological relevance of QCL peeks and 
-hotspots in regulatory network reconstruction.
+In two example we reconstruct part of the genetic architecture in S. Cerevisiae, another 
+example shows the reconstruction of the chain elongation pathway in A. Thaliana. In both 
+examples we are able to map the observe differences in correlation to genetic markers, 
+and show the biological relevance of QCL peeks and hotspots in regulatory network 
+reconstruction.
 
@@
 
-\noindent Corresponding Author: \verb|Danny.Arends@Gmail.com|
+\noindent Corresponding Author: Danny Arends \verb|Danny.Arends@Gmail.com|
 \noindent Associate Editor: XXXX XXXX
@@
 obtained by contacting the authors of the original paper. Expression of probes was 
-measured under growth conditions containing 1% ethanol. Additional details about the 
+measured under growth conditions containing 1\% ethanol. Additional details about the 
 experimental settings are available in the original paper (Brem et al. 2005).
@@
 @
+
+\emph{\sffamily The Phenotypes x Marker relations}
+
 Phenotype x Marker image plot, this is 'comparable' to a QTL scan on mutliple traits.
@@
 
-Phenotype x Phenotype image plot, the new network information obtained by QCL mapping, 
+\emph{\sffamily The Phenotypes x Phenotypes relations}
+
+Phenotype x Phenotype image plot, additional information obtained by QCL mapping, 
 which can be used to reconstruct the underlying genetic architecture.
@@
 expressed between various conditions, or groups of genes coexpressed across 
-samples. Many genetical genomics screens have been carried out to map differential 
+samples.
+
+Many genetical genomics screens have been carried out to map differential 
 expression to genetic variation. Screening the genome for genotype combinations 
 for which there is differential expression has been a successful strategy when 
-mapping Mendelian traits. Results on complex traits have been however less then 
-satisfactory with only small amount of heritability explained by individuals loci. 
+mapping Mendelian traits. However results on complex traits have been less 
+then satisfactory with only small amounts of heritability explained by 
+individuals loci. 
 QCL mapping however 'borrows' information from each phenotype in the analysis set, 
-and enables easy extraction of phenotype marker phenotype networks. furthermore QCL 
-seem to 'transfer' between phenotypes in a more stabile fashion which gives greater 
-detail and confidence when reconstructing genetic architecture.
+and enables extraction of phenotype to marker to phenotype networks. 
+Furthermore QCLs seem to 'transfer' from phenotype to phenotypes in a stabile 
+fashion, which gives greater detail and confidence when reconstructing genetic 
+architecture. 
+It can provides better insight in the missing of heritability of variation, we 
+hypothesize that changes in variation are affecting (by loss of) correlation in 
+the normally correlated phenotypes leading to an amplification of this variation, 
+by distrupting the network environment of which endo-phenotypes are part of.
 
@@
 
-Langfelder P, Horvath S (2008) WGCNA: an R package for weighted correlation network analysis. BMC Bioinformatics 2008, 9:559
+Langfelder P, Horvath S (2008) WGCNA: an R package for weighted correlation network 
+analysis. \emph{BMC Bioinformatics}, 9:559
 

### 2011-11-10 inst/doc/article.Rnw
Added the science style + updates to the paper

@@
 
+\usepackage{scicite}
 \usepackage{times}
@@
 \usepackage{url}
+\usepackage{graphicx}
 \usepackage{color}
@@
 \setlength{\headheight}{0.0in}
-\setlength{\topmargin}{-0.25in}
+\setlength{\topmargin}{0.0cm}
 \setlength{\headsep}{0.0in}
-\setlength{\textheight}{9.00in}
-\setlength{\footskip}{0.5in}
+\setlength{\textheight}{21.0cm}
+\setlength{\footskip}{1.0cm}
 \setlength{\oddsidemargin}{0in}
 \setlength{\evensidemargin}{0in}
-\setlength{\textwidth}{6.5in}
+\setlength{\textwidth}{16.0cm}
 
@@
 
+%The next command sets up an environment for the abstract to your paper.
+
+\newenvironment{sciabstract}{%
+\begin{quote} \bf}
+{\end{quote}}
+
+% If your reference list includes text notes as well as references,
+% include the following line; otherwise, comment it out.
+
+\renewcommand\refname{References and Notes}
+
+% The following lines set up an environment for the last note in the
+% reference list, which commonly includes acknowledgments of funding,
+% help, etc.  It's intended for users of BibTeX or the {thebibliography}
+% environment.  Users who are hand-coding their references at the end
+% using a list environment such as {enumerate} can simply add another
+% item at the end, and it will be numbered automatically.
+
+\newcounter{lastnote}
+\newenvironment{scilastnote}{%
+\setcounter{lastnote}{\value{enumiv}}%
+\addtocounter{lastnote}{+1}%
+\begin{list}%
+{\arabic{lastnote}.}
+{\setlength{\leftmargin}{.22in}}
+{\setlength{\labelsep}{.5em}}}
+{\end{list}}
+
+\title{QCL: Added value of differential correlation}
+
+% Place the author information here.  Please hand-code the contact
+% information and notecalls; do *not* use \footnote commands.  Let the
+% author contact information appear immediately below the author names
+% as shown.  We would also prefer that you don't change the type-size
+% settings shown here.
+
+\author
+{Danny Arends,$^{1\ast,2}$ Ritsert C. Jansen,$^{1}$\\
+\\
+\normalsize{$^{1}$Groningen Bioinformatics Centre - University of Groningen}\\
+\normalsize{Nijenborgh 7, 9747 AG Groningen, The Netherlands}\\
+\normalsize{$^{2}$Genomics Coordination Centre - University Medical Centre Groningen}\\
+\normalsize{Hanzeplein 1, 9700 RB Groningen, The Netherlands}\\
+\\
+\normalsize{$^\ast$To whom correspondence should be addressed; E-mail:  Danny.Arends@Gmail.com.}
+}
+
+% Include the date command, but leave its argument blank.
+
+\date{}
+
+%%%%%%%%%%%%%%%%% END OF PREAMBLE %%%%%%%%%%%%%%%%
+
 \begin{document}
+
+% Double-space the manuscript.
+
+\baselineskip24pt
+
+% Make the title.
+
+\maketitle
 
@@
 
-\begin{center}
-\vspace*{1in}
-\textbf{\Large QCL: Added value of differential correlation }
-
-\bigskip \bigskip 
-\bigskip \bigskip
-
-{\large Danny Arends$^{1,2}$ and Ritsert C. Jansen$^{1}$}
-
-\bigskip \bigskip
-
-{\color{blue} \tt \bfseries $^{1}$ Groningen Bioinformatics Centre - University of Groningen\\
-$^{2}$ Genomics Coordination Centre - University Medical Centre Groningen\\
-version: \# 0.0.1\\[24pt]
-received on XX XXX 201X\\
-accepted on XX XXX 201X\\
-}
-
-\end{center}
-
-\bigskip \bigskip
-
-\textbf{\sffamily Abstract}: In this paper we present Quanitative Correlation Locus 
-(QCL) mapping, a novel approach to detect genetic regulation of phenotypes in 
-recombinant inbred line populations (RIL) or other experimental crosses. QCL mapping 
-allows association of phenotype correlation differences to genetic markers, in almost 
-the same way as QTL mapping associates genetic markers with differences in expression.
+\begin{sciabstract}
+In this paper we present Quanitative Correlation Locus (QCL) mapping, a novel approach 
+to detect genetic regulation of phenotypes in recombinant inbred line populations (RIL) 
+or other experimental crosses. QCL mapping allows association of phenotype correlation 
+differences to genetic markers, in almost the same way as QTL mapping associates genetic 
+markers with differences in expression.
 We use QCL mapping to reconstruct underlying genetic architecture, and provide insight 
@@
 information leads to a more detailed insight into the genetic architecture underlying 
-complex traits. 
-In two example we reconstruct part of the genetic architecture in S. Cerevisiae, another 
-example shows the reconstruction of the chain elongation pathway in A. Thaliana. In both 
-examples we are able to map the observe differences in correlation to genetic markers, 
-and show the biological relevance of QCL peeks and hotspots in regulatory network 
-reconstruction.
+complex traits. The method also provides a vital link in regulatory network reconstruction.
+\end{sciabstract}
 
-\vfill
-
-\noindent Corresponding Author: Danny Arends \verb|Danny.Arends@Gmail.com|
-\noindent Associate Editor: XXXX XXXX
-\newpage
-
-\textbf{\sffamily Introduction}
+\section*{Introduction}
 
 Genetical genomics experiments have shown that gene expression variation can be mapped to 
-genotypic variation (Jansen and Nap, 2001). Quantitative Trait Locus (QTL) mapping of a 
+genotypic variation \cite{Jansen:2001}. Quantitative Trait Locus (QTL) mapping of a 
 gene identifies regions in the genome for which different genotypes lead to differential 
@@
 differentially expressed between conditions, the genes that are differentially coexpressed 
-between conditions (Tesson et al. 2010, Kompass et al. 2011). Such differential co-expression 
+between conditions \cite{Tesson:2010,Kompass:2011}. Such differential co-expression 
 is taken as another level of information from genomic data. This level of information can for 
-example be used to expose transcriptional network rewiring (Langfelder et al. 2008). Differential 
+example be used to expose transcriptional network rewiring \cite{Langfelder:2008}. Differential 
 co-expression has been observed in studies of different diseases, including between healthy and 
@@
 experimental crosses.
-
-We proposed to map not differential co-expression but differential correlation to genetic 
-variation, i.e. to identify regions in the genome for which one genotype value leads to 
-correlated expression between a start gene and all other genes, exploiting the fact that any 
-cis-QTL will induce variation which will in turn lead to correlation differences among 
-individuals carrying the other genotype. We merge information comming from QCL and QTL to reveal 
-underlying genetic architecture, which would remain hidden when using only classical QTL information.
-
-We urge all QTL researchers to combine QCL analysis with existing QTL analysis to expose 
-transcriptional networks involved in the regulation of complex phenotypes. For this we 
-provide an R package with a reference implementation  of the QCL algorithm
-
-\bigskip
-
-\textbf{\sffamily Methodology}
-\nopagebreak
-
-We show two examples in which we reconstruct the genetic architecture using QCL mapping. 
-Both datasets are available inside the R package, as well as the code to analyse these 
-two datasets.
-
-\emph{\sffamily Dataset Example 1. A. Thaliana}
-
-Metabolite data from an A. Thaliana Landsberg erecta (Ler) and Cape Verde Islands (Cvi) 
-dataset from R/qtl, 24 characterized metabolites which are the products of chain elongation. 
-There are 162	recombinant inbred lines genotyped at 117 markers on 5 chromosomes. Additional 
-details about the experimental settings are available in the original paper (Keurentjes et al. 2006,
-Alonso-Blanco et al. 1998)
-
-\emph{\sffamily Dataset Example 2. S. Cerevisiae}
-
-S. Cerevisiae dataset from the gene expression omnibus (GEO) (GDS1115 / GDS1116). This 
-dataset contains expression profiles of parental strains and haploid progenies from a 
-cross of strain BY4716 and the wild wine strain RM11-1a. Data was collected for 4482 
-probes. Genotype data for the 109 progenies was available at 282 genetic markers 
-obtained by contacting the authors of the original paper. Expression of probes was 
-measured under growth conditions containing 1\% ethanol. Additional details about the 
-experimental settings are available in the original paper (Brem et al. 2005).
-
-\emph{\sffamily Preprocessing}
-
-Probes showing little to no variance (< 0.01) were removed from the analysis. Batch 
-effects due to different hybridization dates were estimated and removed from the data 
-by using a mean centering approach (Group means were estimated and corrected).
-
-\textbf{\sffamily Algorithm: QCL mapping}
 
@@
 significantly.
+
+We proposed to map not differential co-expression but differential correlation to genetic 
+variation, i.e. to identify regions in the genome for which one genotype value leads to 
+correlated expression between a start gene and all other genes, exploiting the fact that any 
+cis-QTL will induce variation which will in turn lead to correlation differences among 
+individuals carrying the other genotype. We merge information coming from QCL and QTL to reveal 
+underlying genetic architecture, which would remain hidden when using only classical QTL information.
+
+We urge all QTL researchers to combine QCL analysis with existing QTL analysis to expose 
+transcriptional networks involved in the regulation of complex phenotypes. For this we 
+provide an R package with a reference implementation  of the QCL algorithm
+
+\bigskip
+
+\section*{Methodology}
+\nopagebreak
+
+\emph{\sffamily Preprocessing}
+
+Probes showing little to no variance (< 0.01) were removed from the analysis. Batch 
+effects due to different hybridization dates were estimated and removed from the data 
+by using a mean centering approach (Group means were estimated and corrected).
+
+\section*{Algorithm: QCL mapping}
 
@@
 
-\begin{figure}
-\centering
-<<singletrait, fig=TRUE, echo=FALSE, height=3.5>>=
-plot(qcl_result, 2, do.legend=FALSE)
-@ 
-\caption{Single trait QCL profile (A. Thaliana).\label{fig:singletrait}}
-\end{figure}
-
 \emph{\sffamily Step 3 Network scan for QCL}
@@
 
-Two things can happen: 
+Two things can occur: 
 1) No genes show a significant change in QCL, which means we 
@@
 
+We show two examples in which we reconstruct the genetic architecture using QCL mapping. 
+Both datasets are available inside the R package, as well as the code to analyse these 
+two datasets.
+
+\subsection*{Materials}
+\emph{\sffamily Dataset Example 1. A. Thaliana}
+
+Metabolite data from an A. Thaliana Landsberg erecta (Ler) and Cape Verde Islands (Cvi) 
+dataset from R/qtl, 24 characterized metabolites which are the products of chain elongation. 
+There are 162	recombinant inbred lines genotyped at 117 markers on 5 chromosomes. Additional 
+details about the experimental settings are available in the original paper \cite{Keurentjes:2006,AlonsoBlanco:1998}
+
+\emph{\sffamily Dataset Example 2. S. Cerevisiae}
+
+S. Cerevisiae dataset from the gene expression omnibus (GEO) \cite{Edgar:2002} (GEO 
+accession GDS1115, and GDS1116, \cite{Brem:2005}. This dataset contains expression 
+profiles of parental strains and haploid progenies from a cross of strain BY4716 and 
+the wild wine strain RM11-1a. Data was collected for 4482 probes. Genotype data 
+for the 109 progenies was available at 282 genetic markers obtained by contacting 
+the authors of the original paper. 
+Expression of probes was measured under growth conditions containing 1\% ethanol. 
+Additional details about the experimental settings are available in the original 
+paper \cite{Brem:2005}.
+
 \bigskip
-\textbf{\sffamily Results}
+\section*{Results}
 \nopagebreak
 
-Global overview of QCL scan after using a full scan for QCL
-<<echo=FALSE>>=
-  qcl_result
-@ 
-Single results from a QCL mapping, shows the pattern at the highest threshold.
-<<echo=FALSE>>=
-  qcl_result[[2]]
-@
+\emph{\sffamily Example 1. A. Thaliana}
 
-\emph{\sffamily The Phenotypes x Marker relations}
+QCL mapping was performed on 24 characterized metabolities measured in Arabidopsis 
+Thaliana. A significance threshold for QCL mapping was determined by using 1000 
+permutations, the outcomes can be seen in table XX. We opted for the stringent 
+threshold at P < 0.001 to use as 'significant' QCL score for further analysis.
 
-Phenotype x Marker image plot, this is 'comparable' to a QTL scan on mutliple traits.
+Overview plots of the results of the QCL mapping in Arabidopsis can be found in 
+Figure XX Phenotype to Marker interaction, which is comparable to the QTL mapping 
+(Fig. XX). The additional information obtained by QCL mapping is shown in a heatmap 
+of the Phenotype to Phenotype interaction matrix (figure XX) and the same information 
+but now represented as an interaction network is seen in figure XX
+\begin{table}[ht]
+\begin{tabular}{ l | c c }
+P-value & QCL threshold Arabidopsis & QCL threshold Saccharomyces\\
+\hline
+5 \% & 0.220 & 0.435\\
+1 \% & 0.294 & 0.495\\
+0.1 \% & 0.375 & 0.562\\
+\end{tabular}
+\caption{QCL thresholds obtained after 1000 permutations}
+\label{permutations}
+\end{table}
 
-\begin{figure}[ht]
-\centering
-<<p2mmatrix, fig=TRUE, echo=FALSE, height=3.5>>=
-par(mar=c(3,8,1.1,1.1))
-image(qcl_result, 0.5, against="markers",do.grid=FALSE)
-@ 
-\caption{Phenotypes x Markers matrix (A. Thaliana).\label{fig:p2mmatrix}}
-\end{figure}
+\emph{\sffamily Example 1. S. Cerevisiae}
 
-\emph{\sffamily The Phenotypes x Phenotypes relations}
-
-Phenotype x Phenotype image plot, additional information obtained by QCL mapping, 
-which can be used to reconstruct the underlying genetic architecture.
-\begin{figure}[ht]
-\centering
-<<p2pmatrix, fig=TRUE, echo=FALSE, height=3.5>>=
-par(mar=c(3,8,1.1,1.1))
-image(qcl_result, 0.5, against="phenotypes")
-@ 
-\caption{Phenotypes x Phenotypes matrix (A. Thaliana).\label{fig:p2pmatrix}}
-\end{figure}
+QCL mapping was performed on
 
 \bigskip
-\textbf{\sffamily Discussion}
+\section*{Discussion}
 \nopagebreak
@@
 
-\newpage
-\textbf{\sffamily Bibliography}
+\section*{Acknowledgements}
+This work was supported by the Centre for BioSystems Genomics (CBSG) and the 
+Netherlands Consortium of Systems Biology (NCSB), both of which are part of the Netherlands 
+Genomics Initiative / Netherlands Organisation for Scientific Research [to DA], and 
+by the EU 7th Framework Programme under the Research Project PANACEA, [222936 to RJ].
 
-Langfelder P, Horvath S (2008) WGCNA: an R package for weighted correlation network 
-analysis. \emph{BMC Bioinformatics}, 9:559
+Also we'd like to mention Joeri van der Velde and Konrad Zych for supporting me with some very 
+valuable feedback on figures presented here.
 
-Jansen R. C, and Nap J. P. (2001) Genetical Genomics: the added value from seggregation. 
-\emph{Trends Genet}, 17(7), 388-391.
-
-Tesson, B. M., Breitling R. and Jansen, R. C. (2010). DiffCoEx: a simple and sensitive method 
-to find differentially coexpressed gene modules. \emph{BMC Bioinformatics}, 11,497-506.
-
-Kompass, K. and Witte, J. (2011) - Co-regulatory expression quantitative trait loci mapping: 
-method and application to endometrial cancer. \emph{BMC Medical Genomics}, 4, 6.
-
-Broman, K.W., Wu, H., Sen, S. and Churchill, G.A. (2003) - R/qtl: QTL mapping in experimental 
-crosses. \emph{Bioinformatics}, 19, 889-890.
-
-Arends, D., Prins, P., Jansen, R.C. and Broman, K.W. (2010) - R/qtl: high-throughput multiple 
-QTL mapping. \emph{Bioinformatics}, 26, 2990-2992.
-
-R Development Core Team (2010) - R: A Language and Environment for Statistical Computing. 
-http://www.R-project.org, ISBN: 3-900051-07-0
-
-Edgar R., Domrachev M., Lash A.E. (2002) - Gene Expression Omnibus: NCBI gene expression and 
-hybridization array data repository. \emph{Nucleic Acids Res.}, 30,207-210.
-
-Keurentjes, J. J., Fu, J., de Vos, C. H., Lommen, A., Hall, R. D., Bino, R. J., van der Plas, 
-L. H., Jansen, R. C., Vreugdenhil, D., Koornneef, M. (2006) - The genetics of plant metabolism. 
-\emph{Nature Genetics.}, 38(7), 842-849.
-
-Alonso-Blanco, C., Peeters, A. J., Koornneef, M., Lister, C., Dean, C., van den Bosch, N., Pot, 
-J., Kuiper, M. T. (1998) - Development of an AFLP based linkage map of Ler, Col and Cvi 
-Arabidopsis thaliana ecotypes and construction of a Ler/Cvi recombinant inbred line population. 
-\emph{Plant J.}, 14, 259-271.
-
-Brem RB, Storey JD, Whittle J, Kruglyak L. (2005) - Genetic interactions between polymorphisms 
-that affect gene expression in yeast. \emph{Nature}, 436(7051),701-703.
+We'd like to thank readers for their attention, and invite them to 
+address any questions to Danny Arends, at Danny.Arends@gmail.com.
 
 \newpage
-{\small
-\noindent \textbf{\sffamily Package qcl function overview} \\
-\noindent \begin{tabular}{ll} 
-\hspace*{25mm} & \hspace*{103mm} \\ \hline
-QCLscan & Genome wide scan for QCL \\
-QCLscan.network & Scan for QCL using start gene(s) \\
-QCLpermute & Significance thresholds for QCL by permutation \\
-plot.QCL & Plot QCL profiles at increasing cut-offs \\
-image.QCL & Image a multiple phenotype scan \\
-\hline
-\end{tabular}
-}
+\bibliography{scibib}
+\bibliographystyle{Science}
+
+\newpage
+
+\section*{Figures}
+
+\begin{figure}[ht]
+  \centerline{\includegraphics{QCL.pdf}}
+  \caption{QCL mapping: The underlying hypothesis of QCL mapping. Seen here is a network of 3 phenotypes, in 
+  which expression of P1, P2 and P3 is measured, and we want to infer the hidden relation between P1 
+  and P2. In our model a genetic pertubation is causing P1 to show a QTL at a certain marker. We expect variation 
+  induced by P1 to transfer to P2, However (like in most biological systems) expression of P2 is tightly controlled, 
+  and variation seen in phenotype 1 wil not induce a QTL in phenotype 2. However when we partition individuals based 
+  on the genotype at the QTL marker of P1, we can make use of the correlation structure between P1 and P2. Allowing 
+  us to infer the relationship between phenotype 1 and 2, when we see difference in correlation between P1A and P2A 
+  and P1B and P2B. We don't observe this difference in correlation between P3 and P1, meaning there is no relation 
+  between P1 and P3}
+  \label{QCL_img}
+\end{figure}
+
+\begin{figure}
+  \centering
+<<singletrait, fig=TRUE, echo=FALSE, height=3.5>>=
+plot(qcl_result, 2, do.legend=FALSE)
+@ 
+  \caption{Single trait QCL profile (A. Thaliana) shows the number of phenotypes that have a significant 
+  QCL with the scanned phenotype at different QCL significance thresholds.\label{fig:singletrait}}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+<<p2mmatrix, fig=TRUE, echo=FALSE, height=3.5>>=
+par(mar=c(3,8,1.1,1.1))
+image(qcl_result, 0.5, against="markers",do.grid=FALSE)
+@ 
+  \caption{Phenotypes x Markers matrix (A. Thaliana) this is 'comparable' to a QTL scan on multiple traits.\label{fig:p2mmatrix}}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+<<p2pmatrix, fig=TRUE, echo=FALSE, height=3.5>>=
+par(mar=c(3,8,1.1,1.1))
+image(qcl_result, 0.5, against="phenotypes")
+@ 
+  \caption{Phenotypes x Phenotypes matrix (A. Thaliana) additional information obtained by QCL mapping, 
+  which can be used to reconstruct the underlying genetic architecture.\label{fig:p2pmatrix}}
+\end{figure}
+
+\begin{figure}[ht]
+  \centerline{\includegraphics{AraNet.pdf}}
+  \caption{Reconstructed network for chain elongation in Arabidopsis Thaliana}
+  \label{AraNet}
+\end{figure}
+
+\begin{figure}[ht]
+  \centerline{\includegraphics{AraPerm.pdf}}
+  \caption{Histogram of scores obtained durting permutation of the QCL mapping methodology}
+  \label{AraPerm}
+\end{figure}
 

### 2011-11-23 inst/doc/article.Rnw
Updates to the paper
Added the stackedHistogram plot, to show the pheno -> marker -> phenotype relations

@@
 
-\emph{\sffamily Preprocessing}
+\emph{\sffamily QCL mapping}
 
-Probes showing little to no variance (< 0.01) were removed from the analysis. Batch 
-effects due to different hybridization dates were estimated and removed from the data 
-by using a mean centering approach (Group means were estimated and corrected).
-
-\section*{Algorithm: QCL mapping}
-
-QCL mapping however uses this buffering to detect interaction partners. The hypothesis 
+QCL mapping uses this buffering to detect interaction partners. The hypothesis 
 is that when we split the population based on genotype at the cis-marker, related phenotypes 
 will loosing correlation with the gene under investigation. This allows us to not only 
-only identify its interaction partners. QCL mapping is thus not suffering from (p1) many 
-phenotypes co-localizing). Additionally it is more sensitive to detect variation caused by 
+only identify its interaction partners, but also QCL mapping does not suffering from (p1) many 
+phenotypes co-localizing. Additionally it is more sensitive to detect variation caused by 
 regulators genes because variation of these genes causes many 'downstream' phenotypes 
@@
 \section*{Results}
+\emph{\sffamily Preprocessing}
+
+Probes showing little to no variance (< 0.01) were removed from the analysis. Batch 
+effects due to different hybridization dates were estimated and removed from the data 
+by using a mean centering approach (Group means were estimated and corrected).
+
 \nopagebreak
@@
 \hline
-5 \% & 0.220 & 0.435\\
-1 \% & 0.294 & 0.495\\
-0.1 \% & 0.375 & 0.562\\
+5 \% & 0.220 & 0.436\\
+1 \% & 0.294 & 0.491\\
+0.1 \% & 0.375 & 0.556\\
 \end{tabular}

### 2011-11-23 R/qcl.permutation.R

@@ -10,37 +10,69 @@
 #
 
 #-- QCLpermute main function --#
-QCLpermute <- function(genotypes, phenotypes, pheno.col, n.perm=10, directory="permutations", verbose=FALSE, ...){
+QCLpermuteMC <- function(genotypes, phenotypes, pheno.col, n.perm=10, n.cores=2, directory="permutations", verbose=FALSE, ...){
+  if(!has_snow()) stop("SNOW is not installed (or not yet loaded)")
   if(!file.exists(directory)) dir.create(directory)
   if(missing(pheno.col)) pheno.col <- 1:ncol(phenotypes)
   QCLperm <- vector("list",length(pheno.col))
   idx <- 1
+  cl <- makeCluster(rep("localhost",n.cores))
   for(p in pheno.col){
     ss <- proc.time()
-    cat("Starting permutation for phenotype",p,"\n")
-    for(x in 1:n.perm){
-      sl <- proc.time()
-      if(verbose) cat("- Starting permutation",x,"/",n.perm,"\n")
-		  phenotypes <- phenotypes[sample(nrow(phenotypes)),]
-      write.table(QCLscan(genotypes, phenotypes, pheno.col=p, ...),file=paste(directory,"/Permutation_",x,".txt",sep=""))
-      el <- proc.time()
-      if(verbose) cat("- Permutation",x,"took:",as.numeric(el[3]-sl[3]),"Seconds.\n")
-    }
-    cat("-",x,"Permutations took:",as.numeric(el[3]-ss[3]),"Seconds.\n")
-    QCLperm[[idx]] <- read.QCLpermute(directory,n.perm,verbose)
+    cat("Starting multi core permutation for phenotype",p,"\n")
+    parLapply(cl,as.list(1:n.perm),get("QCLpermute.internal"), genotypes=genotypes, phenotypes=phenotypes, pheno.col=p,directory=directory,verbose=verbose)
+    QCLperm[[idx]] <- read.QCLpermute(directory, p, n.perm, verbose)
     idx <- idx+1
+    el <- proc.time()
+    cat("-",n.perm,"permutations took:",as.numeric(el[3]-ss[3]),"seconds.\n")
   }
+  stopCluster(cl)
   invisible(QCLperm)
 }
 
-read.QCLpermute <- function(directory="permutations", n.perm, verbose){
[…]

### 2011-11-23 R/qcl.statistics.R

@@ -64,11 +64,15 @@ extrapolateBeyondRange <- function(permvalues, value = 0.6){
 }
 
 QCLtoLOD <- function(QCLscan, permutations, pheno.col = 1){
-  apply(-log10(QCLtoPvalue(QCLscan, permutations, pheno.col)),1,sum)
+  -log10(QCLtoPvalue(QCLscan, permutations, pheno.col))
+}
+
+QCLtoLODvector <- function(QCLscan, permutations, pheno.col = 1){
+  apply(QCLtoLOD(QCLscan, permutations, pheno.col),1,sum)
 }
 
 QCLscantoScanone <- function(cross, QCLscan, permutations, pheno.col=1){
-  scores <- QCLtoLOD(QCLscan, permutations, pheno.col)
+  scores <- QCLtoLODvector(QCLscan, permutations, pheno.col)
   scores[which(!is.finite(scores))] <- NA
   lodscorestoscanone(cross, scores)
 }

### 2011-11-23 tests/test_qcl.R

@@ -6,7 +6,7 @@ if(has_rqtl()){
   cross <- read.cross("csvr",file="brem_cross.csvr",geno=c("AA","AB"))
   cross <- convert2riself(cross)
 
-  QCL <- QCLscan.cross(cross,1:3, verbose=T)
+  QCL <- QCLscan.cross(cross,1, verbose=T)
   image(QCL)
   plot(QCL)
 

### 2011-11-23 tests/test_qcl_long.R

@@ -3,10 +3,12 @@ data(ath.metabolites)
 library(qtl)
 data(multitrait)
 qcl_result <- QCLscan.cross(multitrait, 1:24)
-qcl_perms <- QCLpermute(ath.metab$genotypes, ath.metab$phenotypes,1:24,30)
-
+qcl_perms <- QCLpermute(ath.metab$genotypes, ath.metab$phenotypes,1:24,300)
 
 for(x in 1:24){
-  plot(QCLscantoScanone(multitrait, qcl_result, qcl_perms, pheno.col=x),scanone(multitrait,pheno.col=x))
-  Sys.sleep(1)
+  jpeg(paste("phenotype",x,".jpg",sep=""),w=1024,h=768)
+  op <- par(mfrow=c(2,1))
+  plot(QCLscantoScanone(multitrait, qcl_result, qcl_perms, pheno.col=x),scanone(multitrait,pheno.col=x),main="QCL versus QTL")
+  plotAsStackedHist(qcl_result, qcl_perms, pheno.col=x, main="Phenotypes contributing to QCL")
+  dev.off()
 }

### 2011-11-23 R/qcl.plot.R
Added the stackedHistogram plot, to show the pheno -> marker -> phenotype relations

@@ -34,3 +34,21 @@ plotAsLOD <- function(x, permutations, pheno.col = 1, qcl.threshold =0.3, do.leg
   pname <- attr(x[[pheno.col]],"name") 
   plot(QCLtoLOD(x,permutations=permutations),type='l')
 }
+
+plotAsStackedHist <- function(qcl_result,qcl_perms,pheno.col=1, ...){
+  summarized <- QCLtoLODvector(qcl_result, qcl_perms, pheno.col=pheno.col)
+  plot(summarized,type='l',...)
+  p <- rep(0,ncol(qcl_result[[pheno.col]]))
+  i <- 1;
+  mycolors <- rainbow(nrow(qcl_result[[pheno.col]]))
+  apply(QCLtoLOD(qcl_result,qcl_perms,pheno.col),2,
+    function(d){
+     for(idx in 1:length(d)){
+        rect(idx-0.5,p[idx],idx+0.5,p[idx]+d[idx],type='s',col=mycolors[i])
+      }
+      p <<- p + d
+      i <<- i + 1
+    }
+  )
+  points(summarized,type='l',lwd=2)
+}

### 2011-11-24 R/qcl.statistics.R
Minor code updates, mostly to plotting

@@ -33,12 +33,20 @@ QCLtoPvalue.internal <- function(QCLscore, sorted, l){
   res <- NULL
   warn <- TRUE
   for(y in QCLscore){
-    index_l <- min(which(sorted > y))
-    index_r <- max(which(sorted < y))
+    if(!is.na(which(sorted > y)&&1)){
+      index_l <- min(which(sorted > y))
+    }else{
+      index_l <- Inf
+    }
+    if(!is.na(which(sorted < y)&&1)){
+      index_r <- max(which(sorted < y))
+    }else{
+      index_r <- Inf
+    }
     if(!is.finite(index_l)){
       res <- c(res,extrapolateBeyondRange(sorted, y)) #res <- c(res,1-((index_r-1)/l))
       if(warn){
-        cat("Warning: scores out of permutation range, please do more permutations\n")
+        #cat("Warning: scores out of permutation range, please do more permutations\n")
         warn <- FALSE  
       }
     }
@@ -60,6 +68,9 @@ extrapolateBeyondRange <- function(permvalues, value = 0.6){
   scale <- mle$scale
   loc <- mle$threshold[1]
   dens <- function(x) dgpd(x, loc, scale, shape)
+  while(as.numeric(dens(value))==0){
+    value <- value - 0.01
+  }
   dens(value)
 }
 

### 2011-11-24 R/qcl.image.R
Minor code updates, mostly to plotting

@@ -8,12 +8,24 @@
 # Image plot routines for QCL analysis
 #
 
-image.QCLscan <- function(x, qcl.threshold = 0.35, against = c("markers","phenotypes"), do.grid=TRUE, grid.col = "black", ...){
-  colorrange <- c("white",gray.colors(100)[100:1])
-  mymatrix <- QCLprofiles(x, qcl.threshold=qcl.threshold, against=against)
+image.QCLscan <- function(x, qcl_perms, qcl.threshold = 0.35, against = c("markers","phenotypes"), do.grid=TRUE, grid.col = "black", verbose = FALSE, ...){
+  colorrange <- c("white",gray.colors(10)[10:1])
+  if(missing(qcl_perms) || against=="phenotypes"){
+    if(verbose) cat("No permutations, using a cut-off leading to sub-optimal results\n")
+    mymatrix <- QCLprofiles(x, qcl.threshold=qcl.threshold, against=against)
+    mainlabel <- paste("QCLs at",qcl.threshold,"phenotypes vs",against[1])
+  }else{
+    mymatrix <- NULL
+    for(p in 1:length(x)){
+      if(verbose) cat("Processing:",p,"from QCL to LOD\n")
+      mymatrix <- rbind(mymatrix,QCLtoLODvector(x,qcl_perms,p))
+      rownames(mymatrix) <- unlist(lapply(x,attr,"name"))
+    }
+    mainlabel <- paste("QCL phenotypes vs",against[1])
+  }
   if(!is.null(mymatrix)){ 
     image(1:ncol(mymatrix),1:nrow(mymatrix),t(mymatrix),
-          main=paste("QCLs at",qcl.threshold,"phenotypes vs",against[1]),
+          main=mainlabel,
           yaxt="n",xaxt="n",ylab="", xlab="",col=colorrange,cex.main=0.7)
     axis(2,rownames(mymatrix),at=1:nrow(mymatrix),las=2,cex.axis=0.5)
     axis(1,colnames(mymatrix),at=1:ncol(mymatrix),las=2,cex.axis=0.3)

### 2011-11-24 R/qcl.plot.R

@@ -28,27 +28,33 @@ plot.QCLscan <- function(x, pheno.col = 1, qcl.threshold =0.3, do.legend=TRUE, .
   }
 }
 
-plotAsLOD <- function(x, permutations, pheno.col = 1, qcl.threshold =0.3, do.legend=TRUE, ...){
+plotAsLOD <- function(x, permutations, pheno.col = 1, addQTL=TRUE, ...){
   npheno <- length(x)
   if(pheno.col > npheno) stop("No such phenotype")
-  pname <- attr(x[[pheno.col]],"name") 
-  plot(QCLtoLOD(x,permutations=permutations),type='l')
+  pname <- paste("Comparison QCL, QTL",attr(x[[pheno.col]],"name"))
+  QCLscores <- QCLtoLODvector(x, permutations, pheno.col=pheno.col)
+  QTLscores <- as.numeric(QTLscan(ath.metab$genotypes, ath.metab$phenotypes,1))
+  plot(c(0,length(QCLscores)),c(0,max(c(QCLscores,QTLscores))), main=pname, ylab="LOD",xlab="Marker")
+  points(QCLscores,type='l',col="black",lwd=3)
+  points(QTLscores,type='l',col="red",lwd=2,lty=3)
+  legend("topleft",c("QCL","QTL"),col=c("black","red"),lty=c(1,3),lwd=c(3,2))
 }
 
-plotAsStackedHist <- function(qcl_result,qcl_perms,pheno.col=1, ...){
+plotAsStackedHist <- function(qcl_result, qcl_perms, pheno.col=1, do.legend=TRUE, ...){
   summarized <- QCLtoLODvector(qcl_result, qcl_perms, pheno.col=pheno.col)
-  plot(summarized,type='l',...)
+  plot(summarized, type='l',main=attr(qcl_result[[pheno.col]],"name"),...)
   p <- rep(0,ncol(qcl_result[[pheno.col]]))
   i <- 1;
   mycolors <- rainbow(nrow(qcl_result[[pheno.col]]))
   apply(QCLtoLOD(qcl_result,qcl_perms,pheno.col),2,
     function(d){
      for(idx in 1:length(d)){
-        rect(idx-0.5,p[idx],idx+0.5,p[idx]+d[idx],type='s',col=mycolors[i])
+        rect(idx-0.5,p[idx],idx+0.5,p[idx]+d[idx],col=mycolors[i])
       }
       p <<- p + d
[…]

### 2011-11-24 inst/doc/article.tex
Major updates to the article

--- 2011-11-23 inst/doc/article.Rnw
@@
 \usepackage{times}
-\usepackage{amsmath}
-\usepackage{url}
-\usepackage{graphicx}
 \usepackage{color}
+\usepackage{Sweave}
 
 % revise margins
-\setlength{\headheight}{0.0in}
-\setlength{\topmargin}{0.0cm}
-\setlength{\headsep}{0.0in}
-\setlength{\textheight}{21.0cm}
-\setlength{\footskip}{1.0cm}
-\setlength{\oddsidemargin}{0in}
-\setlength{\evensidemargin}{0in}
-\setlength{\textwidth}{16.0cm}
-
-\setlength{\parskip}{6pt}
-\setlength{\parindent}{0pt}
-
-\newcommand{\code}{\texttt}
-\newcommand{\lod}{\text{LOD}}
-
-% make "Figure" within figure captions a small font
-\renewcommand{\figurename}{\small Figure}
+\topmargin 0.0cm
+\oddsidemargin 0.2cm
+\textwidth 16cm 
+\textheight 21cm
+\footskip 1.0cm
 
@@
 \author
-{Danny Arends,$^{1\ast,2}$ Ritsert C. Jansen,$^{1}$\\
+{Danny Arends,$^{1,2,\ast}$ Ritsert C. Jansen,$^{1}$\\
 \\
@@
 
-\SweaveOpts{prefix.string=fig}
 \setkeys{Gin}{width=\textwidth} %% <- change width of figures
 % Try to get the R code from running into the margin
-<<echo=FALSE>>=
-options(width=87, digits=3, scipen=4)
-@
-
-<<echo=FALSE,results=hide>>=
-  require(qcl)
-  set.seed(61777369)
-  data(ath.metabolites) # Arabidopsis Thaliana dataset
-  qcl_result <- QCLscan(ath.metab$genotypes, ath.metab$phenotypes)
-@
 
@@
   to detect genetic regulation of phenotypes in recombinant inbred line populations (RIL) 
-or other experimental crosses. QCL mapping allows association of phenotype correlation 
+  or SNP genotypes populations. QCL mapping allows association of phenotype correlation 
   differences to genetic markers, in almost the same way as QTL mapping associates genetic 
@@
   information leads to a more detailed insight into the genetic architecture underlying 
-complex traits. The method also provides a vital link in regulatory network reconstruction.
+  complex traits. The method also provides a missing link in regulatory network reconstruction.
 \end{sciabstract}
@@
 
-We proposed to map not differential co-expression but differential correlation to genetic 
+  This inspired us to re-think parts of the basic assumptions in genetics:
+  Imagine phenotype A shows a QTL, we suppose that near this marker something (e.g. DNA variation, 
+  methylation, promotor sequences, etc) introduces expression variation in A. If we now assume there is a 
+  phenotype B (downstream to A) variation of expression in phenotype A is causing 'variation' 
+  for phenotype B. However B will in most cases not show variation because it's expression 
+  depends (but is not dependant) on A, a phenomena generally refered to as phenotype buffering. 
+  This is what QCL mapping exploits, the hypothesis is that although B doesn't show a QTL, 
+  it will become uncorrelated when we partition phenotype B into the groups based on the genotype at 
+  the QTL marker of phenotype A.
+
+  QCL thus maps not differential co-expression but differential correlation to genetic 
   variation, i.e. to identify regions in the genome for which one genotype value leads to 
-correlated expression between a start gene and all other genes, exploiting the fact that any 
-cis-QTL will induce variation which will in turn lead to correlation differences among 
-individuals carrying the other genotype. We merge information coming from QCL and QTL to reveal 
-underlying genetic architecture, which would remain hidden when using only classical QTL information.
+  correlated expression between a phenotype and all other phenotypes, exploiting the fact 
+  that any QTL will be the result of either direct induced variation or indirect variation, 
+  which will be visible as correlation differences among individuals carrying different genotypes. 
+  We merge information coming from QCL and QTL to reveal underlying genetic architecture, 
+  which would remain hidden when using only classical QTL information.
 
-We urge all QTL researchers to combine QCL analysis with existing QTL analysis to expose 
+  We urge all QTL researchers to combine the new QCL method with existing QTL analysis to expose 
   transcriptional networks involved in the regulation of complex phenotypes. For this we 
-provide an R package with a reference implementation  of the QCL algorithm
-
+  provide an R package with a reference implementation of the QCL algorithm.
 \bigskip
@@
 
-\emph{\sffamily QCL mapping}
-
-QCL mapping uses this buffering to detect interaction partners. The hypothesis 
-is that when we split the population based on genotype at the cis-marker, related phenotypes 
+\emph{\sffamily QCL mapping}\\
+  As mentioned in the introduction QCL mapping uses buffering to detect interaction partners. 
+  When we split the population based on genotype at a QTL marker, only related phenotypes 
   will loosing correlation with the gene under investigation. This allows us to not only 
-only identify its interaction partners, but also QCL mapping does not suffering from (p1) many 
-phenotypes co-localizing. Additionally it is more sensitive to detect variation caused by 
-regulators genes because variation of these genes causes many 'downstream' phenotypes 
-to show differential correlation with the inducing phenotype, because QCL relies on 
+  only identify its interaction partners, but also means QCL mapping does not suffering from 
+  (p1) many phenotypes co-localizing. Additionally it is more sensitive to detect variation 
+  caused by regulatory phenotypes because variation of these regulators causes many 'downstream' 
+  phenotypes to show differential correlation with the inducing phenotype, because QCL relies on 
   correlation changes as opposed to differences in phenotype mean QCL is unaffected by (p2).
 
-\emph{\sffamily Choosing a start/seed probe (Optional)}
-
+\emph{\sffamily Choosing a start/seed probe (Optional)}\\
   Choose a phenotype to subject to QCL analysis, this can be your favourite gene, 
@@
 
-\emph{\sffamily Single phenotype QCL}
-
+\emph{\sffamily Single phenotype QCL mapping}\\
   Using genotype data we can split individuals at each marker in two groups (A or B).
@@
 
-\emph{\sffamily Significance testing}
+\emph{\sffamily Significance testing}\\
+  To compare to QTL mapping we need to convert the QCLscores to LOD scores. Significance 
+  of QCL were assessed by using permutations, in each round the link between genotype and 
+  phenotype is broken, by redistributing at random genotypes amongst individuals not 
+  allowing for duplicates. The QCL on these new individuals allows for estimation of 
+  the underlying null-distribution of QCL scores. We frequently observe QCL scores in 
+  real data which is higher than any QCL score obtained during permutation. When this 
+  happens we use a generalized Pareto distribution (GPD) to estimate the extreme tail 
+  of the distribution \cite{Knijnenburg:2009}, to also get likelyhood estimates for the 
+  extreme scores observed. See figures XX, and XX showing the score distribution in 
+  permuted data, Fig XX also shows a tail estimation using the GPD method.
 
-The threshold for a significant difference in QCL is assessed by using permutations, in 
-each round the link between genotype and phenotype is broken, by redistributing at random 
-genotypes amongst individuals not allowing for duplicates. The QCL on these new 
-individuals allows for estimation of the underlying null-distribution of QCL scores. 
-See figures XX, and XX showing the score distribution in permuted data.
+\emph{\sffamily Creating single trait QCL profiles}\\
+  After permutation single trait QCL profiles can be obtained by transforming the QCL scores 
+  into LOD scores. For this we use a look-up table of scores obtained during permutation, we 
+  assess the likelyhood and take the $-log10(P.value)$ of every QCL score see Fig. 
+  \ref{fig:singletraitLOD}. This is done 1) to compare QTL profiles with summarized QCL profiles 
+  (an example in Fig. \ref{fig:singletrait}), and 2) Show the contribution of each phenotype 
+  to the total likelyhood.
 
-\emph{\sffamily Creating single trait QCL profiles}
+\emph{\sffamily Visualizing the Phenotypes x Phenotypes QCL relations}\\
+  In Fig. \ref{fig:singletraitLOD} we see the additional information obtained by QCL mapping summarized 
+  graphically.  We can also calculate the full Phenotypes x Phenotypes matrix. This shows the summed 
+  evidence of all paiwise interaction between multiple phenotypes.
 
-After permutation single trait QCL profiles can be obtained by setting a qcl significance 
-threshold. This threshold defines when a change in correlation is significant, and at 
-each marker we can plot the count of number the other phenotypes that show a significant 
-difference in correlation. Overplotting chromosome, QTL information, and phenotype location 
-(if applicable) produces the single QCL plots seen in figures XX, and XX.
+\emph{\sffamily Visualizing the Phenotypes x Marker relations}\\
+  When plotting multiple QCL profiles as a heatmap, QCLs are seen mapping to the 
+  same genetic location. At these locations, regulation is taking place. We expect 
+  to find one (or more) of the genes to show a cis-QTL, this is because a QCL can 
+  only exist if there is a QTL inducing it.
 
-\emph{\sffamily Step 3 Network scan for QCL}
-
-If we find any significant QCL information we can try to reconstruct the interaction 
-network of the phenotype. We first do a single phenotype QCL scan (step 2), then 
-using a permutation determin a qcl threshold, and select phenotypes which show a 
-significant change in QCL.
-
-Two things can occur: 
-1) No genes show a significant change in QCL, which means we 
-cannot find any interaction partners for this phenotype, or 
-2) This genes has an QCL with other genes, then the algorithm adds 
-these genes onto a list to scan and repeats the single phenotype 
-QCL scan (step 2) for these phenotypes (not allowing for duplicate 
-scans). This is repeated until no more significant QCL are found or 
-a user defined search depth is reached.
-
-\emph{\sffamily Visualizing the Phenotypes x Marker relations}
-
-When plotting multiple QCL profiles as a heatmap, multiple QCL are seen mapping to the 
-same genetic location. At these locations, we expect to find one (or more) of the genes 
-to show a cis-QTL, this is because a QTL can induce a QCL, because variation of expression 
-in the phenotype is causing other phenotypes to become uncorrelated, when assigned to 
-groups determined by the QTL marker. This phenotype (or phenotypes) introduced genetic 
-variation, which is seen also (and more pronounced) in a QTL scan.
-
-\emph{\sffamily Visualizing the Phenotypes x Phenotypes relations}
-
-The additional information obtained by QCL mapping is summarized in the 
-Phenotypes x Phenotypes matrix. This shows the summed evidence of interaction 
-between the phenotypes showing QCL transmitted through all the markers and the phenotype scanned.
-
-\emph{\sffamily Network reconstruction}
-
+\emph{\sffamily Network reconstruction}\\
   So how do we build up gene interaction networks, This looks complicated but the information 
@@
 \subsection*{Materials}
-\emph{\sffamily Dataset Example 1. A. Thaliana}
-
+\emph{\sffamily Dataset Example 1. A. Thaliana}\\
   Metabolite data from an A. Thaliana Landsberg erecta (Ler) and Cape Verde Islands (Cvi) 
@@
 
-\emph{\sffamily Dataset Example 2. S. Cerevisiae}
-
+\emph{\sffamily Dataset Example 2. S. Cerevisiae}\\
   S. Cerevisiae dataset from the gene expression omnibus (GEO) \cite{Edgar:2002} (GEO 
@@
 
-\emph{\sffamily Example 1. A. Thaliana}
+  \subsection{Example 1. A. Thaliana}
+  QCL mapping was performed on 24 characterized metabolities measured in Arabidopsis 
+  Thaliana. A significance threshold for QCL mapping was determined for each trait 
+  by using 1000 permutations, the outcomes can be seen in Fig \ref{fig:histPermAra}. 
+  We used a very stringent QCL threshold at P < 0.001 to use as 'significant' QCL 
+  score for further analysis.
 
-QCL mapping was performed on 24 characterized metabolities measured in Arabidopsis 
-Thaliana. A significance threshold for QCL mapping was determined by using 1000 
-permutations, the outcomes can be seen in table XX. We opted for the stringent 
-threshold at P < 0.001 to use as 'significant' QCL score for further analysis.
+  \emph{\sffamily Significance}\\
+  The major observations is that we cannot use a 'flat' cut-off for all traits, each 
+  trait shows a distinct NULL-distribution. A squared correlation difference of 0.5 is 
+  significant for X3.Hydroxypropyl (green). But for X4.Methylsulfinylbutyl we need a 
+  difference in squared correlation larger then 0.7 (yellow).
 
-Overview plots of the results of the QCL mapping in Arabidopsis can be found in 
+  \emph{\sffamily Single trait QCL}\\
+  The 24 QCL profiles were compared to the QTL profiles previously determined. We observed 3 cases:
+  1) Additional QCL found, where no QTL was present as seen in the QCL profile X3.Hydroxypropyl (Fig. \ref{fig:singletraitLOD}), 
+  These are the cases in which the loss of variation in the genes underneath the QCL is not causing 
+  variation differences in the observed phenotype, but correlation differences.
+  2) QCL co-localized with QTL, (Also QCL gives a detailed view of which phenotypes are underneath its peek, we 
+  interpret this as variation transfered from the phenotypes underneath the QCL to the phenotype investigated.
+  3) QTL with no QCL, we interpret this as the source of the variation.
+  Analysis in this way of all 24 Single trait QCL profiles combined with their QTL profiles 
+  allows us to reconstruct the source of natural variation and follow it down the network.
+
+  \emph{\sffamily QCL combined with QTL}\\
+  Plots of the results of the QCL mapping in Arabidopsis can be found in 
   Figure XX Phenotype to Marker interaction, which is comparable to the QTL mapping 
@@
   of the Phenotype to Phenotype interaction matrix (figure XX) and the same information 
-but now represented as an interaction network is seen in figure XX
-\begin{table}[ht]
-\begin{tabular}{ l | c c }
-P-value & QCL threshold Arabidopsis & QCL threshold Saccharomyces\\
-\hline
-5 \% & 0.220 & 0.436\\
-1 \% & 0.294 & 0.491\\
-0.1 \% & 0.375 & 0.556\\
-\end{tabular}
-\caption{QCL thresholds obtained after 1000 permutations}
-\label{permutations}
-\end{table}
+  but now represented as an interaction network is seen in figure XX and in the QCL plot
 
-\emph{\sffamily Example 1. S. Cerevisiae}
+  \emph{\sffamily Reconstructed network}\\
+  Using the derived data we can begin to reconstruct the network. We'll start with a metabolite 
+  which shows mappable variation changes, but no QCL colocatizing (Inducer).
 
-QCL mapping was performed on
+  \subsection{Example 2. S. Cerevisiae}
+  QCL mapping is computationaly expensive (We need 1000+ permutations for each phenotype), 
+  and for the S. Cerevisiae we pre-selected a background of 300 probes which showed the 
+  highest variation, and had been previously been identified as having QTLs. Significance 
+  thresholds for QCL mapping were determined for each trait by using 1000 permutations 
+  ({\emph Data not shown}).
+
+  \emph{\sffamily Single trait QCL}\\
 
@@
 
-Microarray experiments have allowed the identification of genes differentially 
-expressed between various conditions, or groups of genes coexpressed across 
-samples.
+  Microarray experiments have identified genes differentially expressed between various 
+  conditions, or groups of genes coexpressed across samples. Genetical genomics screens 
+  have been carried out to map differential expression to genetic variation. Screening 
+  the genome for genotype combinations for which there is differential expression has 
+  been a successful strategy when mapping Mendelian traits.
 
-Many genetical genomics screens have been carried out to map differential 
-expression to genetic variation. Screening the genome for genotype combinations 
-for which there is differential expression has been a successful strategy when 
-mapping Mendelian traits. However results on complex traits have been less 
-then satisfactory with only small amounts of heritability explained by 
-individuals loci. 
-QCL mapping however 'borrows' information from each phenotype in the analysis set, 
-and enables extraction of phenotype to marker to phenotype networks. 
+  However results on complex traits have been less then satisfactory with only small 
+  amounts of heritability explained by individuals loci. QCL mapping however 'borrows' 
+  information from each phenotype in the analysis set, and enables extraction of 
+  phenotype to marker to phenotype networks, the flow of variation in the network. 
   Furthermore QCLs seem to 'transfer' from phenotype to phenotypes in a stabile 
   fashion, which gives greater detail and confidence when reconstructing genetic 
-architecture. 
-It can provides better insight in the missing of heritability of variation, we 
+  architecture. Also QCL allows us to better identify gene transcript introducing 
+  variation, by their absense under QTL peeks which introduce variation, compared 
+  to being present underneath QTL which are affected by upstream variation.
+
+  It provides better insight in the missing heritability observed in phenotypes by 
+  allowing it to be bufferen and thus be hiden for classical QTL mapping approaches, we 
   hypothesize that changes in variation are affecting (by loss of) correlation in 
@@
   which expression of P1, P2 and P3 is measured, and we want to infer the hidden relation between P1 
-  and P2. In our model a genetic pertubation is causing P1 to show a QTL at a certain marker. We expect variation 
+  and P2. In our model a genetic pertubation is causing P1 to change in expression showing a QTL at a marker. We expect variation 
   induced by P1 to transfer to P2, However (like in most biological systems) expression of P2 is tightly controlled, 
@@
 
+\begin{figure}[ht]
+  \centering
+\includegraphics{fig-permutation_scores}
+  \caption{QCL scores obtained during permutations (A. Thaliana) additional information obtained by QCL mapping, 
+  which can be used to reconstruct the underlying genetic architecture.\label{fig:histPermAra}}
+\end{figure}
+
 \begin{figure}
   \centering
-<<singletrait, fig=TRUE, echo=FALSE, height=3.5>>=
-plot(qcl_result, 2, do.legend=FALSE)
-@ 
-  \caption{Single trait QCL profile (A. Thaliana) shows the number of phenotypes that have a significant 
-  QCL with the scanned phenotype at different QCL significance thresholds.\label{fig:singletrait}}
+\includegraphics{fig-singletrait}
+  \caption{QCL profile (A. Thaliana, X3.Hydroxypropyl) transformed to LOD scores compared to the QTL profile of the same phenotype, We see an 
+  additional significant peek for this phenotype (Between Marker 50 to 60) \label{fig:singletrait}}
+\end{figure}
+
+\begin{figure}
+  \centering
+\includegraphics{fig-singletraitLOD}
+  \caption{QCL profile (A. Thaliana, X3.Hydroxypropyl) transformed to LOD scores. Shown is the likelyhood contribution of each of 
+  the 24 metabolites to the QCL profile. \label{fig:singletraitLOD}}
 \end{figure}
@@
   \centering
-<<p2mmatrix, fig=TRUE, echo=FALSE, height=3.5>>=
-par(mar=c(3,8,1.1,1.1))
-image(qcl_result, 0.5, against="markers",do.grid=FALSE)
-@ 
+\includegraphics{fig-p2mmatrix}
   \caption{Phenotypes x Markers matrix (A. Thaliana) this is 'comparable' to a QTL scan on multiple traits.\label{fig:p2mmatrix}}
@@
   \centering
-<<p2pmatrix, fig=TRUE, echo=FALSE, height=3.5>>=
-par(mar=c(3,8,1.1,1.1))
-image(qcl_result, 0.5, against="phenotypes")
-@ 
+\includegraphics{fig-QTLmatrix}
+  \caption{Heatmap showing QTL scores for all 24 metabolities in A. Thaliana.\label{fig:QTLmatrix}}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+\includegraphics{fig-p2pmatrix}
   \caption{Phenotypes x Phenotypes matrix (A. Thaliana) additional information obtained by QCL mapping, 
@@
 
-\begin{figure}[ht]
-  \centerline{\includegraphics{AraPerm.pdf}}
-  \caption{Histogram of scores obtained durting permutation of the QCL mapping methodology}
-  \label{AraPerm}
-\end{figure}
-
 \end{document}

### 2011-11-24 inst/doc/article.Rnw
Uploaded the tex file by accident

@@
   require(qcl)
-  set.seed(61777369)
+  set.seed(200)
   data(ath.metabolites) # Arabidopsis Thaliana dataset
+  cat("QCL scan\n")
   qcl_result <- QCLscan(ath.metab$genotypes, ath.metab$phenotypes)
+  cat("QTL scan\n")
+  qtl_result <- QTLscan(ath.metab$genotypes, ath.metab$phenotypes)
+  cat("QCL permutation\n")
+  qcl_perms <- QCLpermute(ath.metab$genotypes, ath.metab$phenotypes,1:24,20, n.cores = 3)
+  #qcl_perms <- vector("list",24)
+  #idx <- 1
+  #for(p in 1:24){
+  #  qcl_perms[[idx]] <- read.QCLpermute("permutations", p, 300,TRUE)
+  #  idx <- idx+1
+  #}
 @
@@
 
-\emph{\sffamily Example 1. A. Thaliana}
+  \subsection{Example 1. A. Thaliana}
+  QCL mapping was performed on 24 characterized metabolities measured in Arabidopsis 
+  Thaliana. A significance threshold for QCL mapping was determined for each trait 
+  by using 1000 permutations, the outcomes can be seen in Fig \ref{fig:histPermAra}. 
+  We used a very stringent QCL threshold at P < 0.001 to use as 'significant' QCL 
+  score for further analysis.
 
-QCL mapping was performed on 24 characterized metabolities measured in Arabidopsis 
-Thaliana. A significance threshold for QCL mapping was determined by using 1000 
-permutations, the outcomes can be seen in table XX. We opted for the stringent 
-threshold at P < 0.001 to use as 'significant' QCL score for further analysis.
+  \emph{\sffamily Significance}\\
+  The major observations is that we cannot use a 'flat' cut-off for all traits, each 
+  trait shows a distinct NULL-distribution. A squared correlation difference of 0.5 is 
+  significant for X3.Hydroxypropyl (green). But for X4.Methylsulfinylbutyl we need a 
+  difference in squared correlation larger then 0.7 (yellow).
 
-Overview plots of the results of the QCL mapping in Arabidopsis can be found in 
+  \emph{\sffamily Single trait QCL}\\
+  The 24 QCL profiles were compared to the QTL profiles previously determined. We observed 3 cases:\\
+  1) Additional QCL found, where no QTL was present as seen in the QCL profile X3.Hydroxypropyl 
+  (Fig. \ref{fig:singletraitLOD}), these are the cases in which the loss of variation in the genes 
+  underneath the QCL is not causing variation differences in the observed phenotype, but still 
+  we're able to detect the correlation differences.\\
+  2) QCL co-localized with QTL, in this situation QCL gives a detailed view of which phenotypes 
+  are underneath its peek, we interpret this as variation transfered from the phenotypes underneath 
+  the QCL to the phenotype investigated.\\
+  3) QTL with no QCL, we interpret this as the source of the variation. Analysis in this way of 
+  all 24 Single trait QCL profiles combined with their QTL profiles allows us to reconstruct the 
+  source of natural variation and follow it down the network.
+
+  \emph{\sffamily QCL combined with QTL}\\
+  Plots of the results of the QCL mapping in Arabidopsis can be found in 
   Figure XX Phenotype to Marker interaction, which is comparable to the QTL mapping 
@@
 
+\begin{figure}[ht]
+  \centering
+<<permutation_scores, fig=TRUE, echo=FALSE, height=3.5>>=
+op <- par(cex=0.5)
+plot(c(0,0.8),c(0,400),type='n',main="QCL scores permutations",ylab="Frequency",xlab="Difference in correlation^2")
+hist(qcl_perms[[3]],add=T, col="yellow")
+hist(qcl_perms[[2]],add=T, col="red")
+hist(qcl_perms[[1]],add=T, col="green")
+legend("topright",colnames(ath.metab$phenotypes)[1:3],col=c("yellow","red","green"),lwd=6)
+op <- par(cex=1)
+@ 
+  \caption{QCL scores obtained during permutations (A. Thaliana) additional information obtained by QCL mapping, 
+  which can be used to reconstruct the underlying genetic architecture.\label{fig:histPermAra}}
+\end{figure}
+
 \begin{figure}
   \centering
-<<singletrait, fig=TRUE, echo=FALSE, height=3.5>>=
-plot(qcl_result, 2, do.legend=FALSE)
+<<singletrait, fig=TRUE, echo=FALSE, height=3.0>>=
+op <- par(cex=0.5)
+plotAsLOD(qcl_result, qcl_perms, pheno.col=1)
+op <- par(cex=1)
 @
-  \caption{Single trait QCL profile (A. Thaliana) shows the number of phenotypes that have a significant 
-  QCL with the scanned phenotype at different QCL significance thresholds.\label{fig:singletrait}}
+  \caption{QCL profile (A. Thaliana, X3.Hydroxypropyl) transformed to LOD scores compared to the QTL profile of the same phenotype, We see an 
+  additional significant peek for this phenotype (Between Marker 50 to 60) \label{fig:singletrait}}
+\end{figure}
+
+\begin{figure}
+  \centering
+<<singletraitLOD, fig=TRUE, echo=FALSE, height=3.0>>=
+op <- par(cex=0.5)
+plotAsStackedHist(qcl_result, qcl_perms, pheno.col=1)
+op <- par(cex=1)
+@ 
+  \caption{QCL profile (A. Thaliana, X3.Hydroxypropyl) transformed to LOD scores. Shown is the likelyhood contribution of each of 
+  the 24 metabolites to the QCL profile. \label{fig:singletraitLOD}}
 \end{figure}
@@
   \centering
-<<p2mmatrix, fig=TRUE, echo=FALSE, height=3.5>>=
+<<p2mmatrix, fig=TRUE, echo=FALSE, height=3.0>>=
 par(mar=c(3,8,1.1,1.1))
-image(qcl_result, 0.5, against="markers",do.grid=FALSE)
+image(qcl_result, qcl_perms, against="markers",do.grid=FALSE)
 @ 
   \caption{Phenotypes x Markers matrix (A. Thaliana) this is 'comparable' to a QTL scan on multiple traits.\label{fig:p2mmatrix}}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+<<QTLmatrix, fig=TRUE, echo=FALSE, height=3.0>>=
+par(mar=c(3,8,1.1,1.1))
+image(1:ncol(qtl_result),1:nrow(qtl_result),t(qtl_result),col=c("white",gray.colors(100)[100:1]), yaxt="n", xaxt="n", ylab="", xlab="")
+axis(2,rownames(qtl_result),at=1:nrow(qtl_result),las=2,cex.axis=0.5)
+axis(1,colnames(qtl_result),at=1:ncol(qtl_result),las=2,cex.axis=0.3)
+@ 
+  \caption{Heatmap showing QTL scores for all 24 metabolities in A. Thaliana.\label{fig:QTLmatrix}}
 \end{figure}

### 2011-11-24 R/qcl.image.R

@@ -19,8 +19,8 @@ image.QCLscan <- function(x, qcl_perms, qcl.threshold = 0.35, against = c("marke
     for(p in 1:length(x)){
       if(verbose) cat("Processing:",p,"from QCL to LOD\n")
       mymatrix <- rbind(mymatrix,QCLtoLODvector(x,qcl_perms,p))
-      rownames(mymatrix) <- unlist(lapply(x,attr,"name"))
     }
+    rownames(mymatrix) <- unlist(lapply(x,attr,"name"))
     mainlabel <- paste("QCL phenotypes vs",against[1])
   }
   if(!is.null(mymatrix)){ 

### 2011-11-25 R/qcl.permutation.R
Updates to the R code to plot more easily

@@ -20,7 +20,11 @@ QCLpermuteMC <- function(genotypes, phenotypes, pheno.col, n.perm=10, n.cores=2,
   for(p in pheno.col){
     ss <- proc.time()
     cat("Starting multi core permutation for phenotype",p,"\n")
-    parLapply(cl,as.list(1:n.perm),get("QCLpermute.internal"), genotypes=genotypes, phenotypes=phenotypes, pheno.col=p,directory=directory,verbose=verbose)
+    rvm <- NULL
+    for(x in 1:n.perm){
+      rvm <- rbind(rvm,sample(nrow(phenotypes)))
+    }
+    parLapply(cl,as.list(1:n.perm),get("QCLpermute.internal"), genotypes=genotypes, phenotypes=phenotypes, pheno.col=p, rvm, directory=directory,verbose=verbose)
     QCLperm[[idx]] <- read.QCLpermute(directory, p, n.perm, verbose)
     idx <- idx+1
     el <- proc.time()
@@ -30,11 +34,11 @@ QCLpermuteMC <- function(genotypes, phenotypes, pheno.col, n.perm=10, n.cores=2,
   invisible(QCLperm)
 }
 
-QCLpermute.internal <- function(perm, genotypes, phenotypes, pheno.col, directory="permutations", verbose=FALSE, ...){
+QCLpermute.internal <- function(perm, genotypes, phenotypes, pheno.col, rvm, directory="permutations", verbose=FALSE, ...){
   require(qcl)
   sl <- proc.time()
   if(verbose) cat("- Starting permutation",perm,"\n")
-  phenotypes <- phenotypes[sample(nrow(phenotypes)),]
+  phenotypes <- phenotypes[rvm[perm,],]
   write.table(QCLscan(genotypes, phenotypes, pheno.col, ...),file=paste(directory,"/Permutation_",pheno.col,"_",perm,".txt",sep=""))
   el <- proc.time()
   if(verbose) cat("- permutation",perm,"took:",as.numeric(el[3]-sl[3]),"seconds.\n")
@@ -55,7 +59,10 @@ QCLpermute <- function(genotypes, phenotypes, pheno.col, n.perm=10, n.cores=2, d
       ss <- proc.time()
       cat("Starting permutation for phenotype",p,"\n")
       for(x in 1:n.perm){
-        QCLpermute.internal(x,genotypes,phenotypes,p,directory,verbose)
[…]

### 2011-11-25 R/qcl.plot.R

@@ -16,7 +16,7 @@ plot.QCLscan <- function(x, pheno.col = 1, qcl.threshold =0.3, do.legend=TRUE, .
   totpheno <- dim(x[[pheno.col]])[1]
   totmarkers <- dim(x[[pheno.col]])[2]
   y_range <- c(0,1.25*max(QCLprofiles(x,qcl.threshold)[pname,]))
-  plot(c(0,totmarkers),y_range,type="n",main=paste("QCL mapping of ",attr(x[[pheno.col]],"name")),ylab="# of significant QCL", xlab="Genetic marker")
+  plot(c(0,totmarkers),y_range,type="n",main=paste("QCL mapping of",attr(x[[pheno.col]],"name")),ylab="# of significant QCL", xlab="Genetic marker")
   colorz <- NULL
   max_value <- max(abs(x[[pheno.col]]))
   for(t in seq(qcl.threshold,max_value,0.05)){
@@ -31,10 +31,10 @@ plot.QCLscan <- function(x, pheno.col = 1, qcl.threshold =0.3, do.legend=TRUE, .
 plotAsLOD <- function(x, permutations, pheno.col = 1, addQTL=TRUE, ...){
   npheno <- length(x)
   if(pheno.col > npheno) stop("No such phenotype")
-  pname <- paste("Comparison QCL, QTL",attr(x[[pheno.col]],"name"))
+  pname <- paste("Comparison QCL:QTL of",attr(x[[pheno.col]],"name"))
   QCLscores <- QCLtoLODvector(x, permutations, pheno.col=pheno.col)
   QTLscores <- as.numeric(QTLscan(ath.metab$genotypes, ath.metab$phenotypes,1))
-  plot(c(0,length(QCLscores)),c(0,max(c(QCLscores,QTLscores))), main=pname, ylab="LOD",xlab="Marker")
+  plot(c(0,length(QCLscores)),c(0,max(c(QCLscores,QTLscores))),type='n', main=pname, ylab="LOD",xlab="Marker")
   points(QCLscores,type='l',col="black",lwd=3)
   points(QTLscores,type='l',col="red",lwd=2,lty=3)
   legend("topleft",c("QCL","QTL"),col=c("black","red"),lty=c(1,3),lwd=c(3,2))
@@ -42,7 +42,7 @@ plotAsLOD <- function(x, permutations, pheno.col = 1, addQTL=TRUE, ...){
 
[…]

### 2011-11-25 R/qcl.statistics.R

@@ -46,7 +46,7 @@ QCLtoPvalue.internal <- function(QCLscore, sorted, l){
     if(!is.finite(index_l)){
       res <- c(res,extrapolateBeyondRange(sorted, y)) #res <- c(res,1-((index_r-1)/l))
       if(warn){
-        #cat("Warning: scores out of permutation range, please do more permutations\n")
+        cat("Warning: scores out of permutation range, please do more permutations\n")
         warn <- FALSE  
       }
     }
@@ -68,8 +68,14 @@ extrapolateBeyondRange <- function(permvalues, value = 0.6){
   scale <- mle$scale
   loc <- mle$threshold[1]
   dens <- function(x) dgpd(x, loc, scale, shape)
+  warn <- FALSE
+  prev.value <- value
   while(as.numeric(dens(value))==0){
+    warn <- TRUE
     value <- value - 0.01
+  }
+  if(warn){
+    cat("Warning: scores out of permutation range, unable to estimate correctly",value,"/",prev.value,"\n")
   }
   dens(value)
 }

### 2011-11-25 inst/doc/article.Rnw
Updates to the article

@@
   cat("QCL permutation\n")
-  qcl_perms <- QCLpermute(ath.metab$genotypes, ath.metab$phenotypes,1:24,20, n.cores = 3)
+  qcl_perms1 <- QCLpermute(ath.metab$genotypes, ath.metab$phenotypes, 1:3, 200, n.cores = 3)
+  qcl_perms2 <- QCLpermute(ath.metab$genotypes, ath.metab$phenotypes, 4:24, 50, n.cores = 3)
+  qcl_perms <- c(qcl_perms1,qcl_perms2)
   #qcl_perms <- vector("list",24)
@@
   When we split the population based on genotype at a QTL marker, only related phenotypes 
-  will loosing correlation with the gene under investigation. This allows us to not only 
-  only identify its interaction partners, but also means QCL mapping does not suffering from 
-  (p1) many phenotypes co-localizing. Additionally it is more sensitive to detect variation 
+  will loose correlation with the gene under investigation. This allows us to not only 
+  only identify its interaction partners, while not suffering from (p1) many phenotypes 
+  co-localizing. Additionally it is more sensitive to detect variation 
   caused by regulatory phenotypes because variation of these regulators causes many 'downstream' 
@@
   of the distribution \cite{Knijnenburg:2009}, to also get likelyhood estimates for the 
-  extreme scores observed. See figures XX, and XX showing the score distribution in 
-  permuted data, Fig XX also shows a tail estimation using the GPD method.
+  extreme scores observed. See Fig. \ref{fig:permutation_scores} showing the score 
+  distribution in permuted data of 3 traits in A. Thaliana, Fig XX also shows a tail 
+  estimation using the GPD method.
 
@@
   \centering
-<<permutation_scores, fig=TRUE, echo=FALSE, height=3.5>>=
+<<permutation_scores, fig=TRUE, echo=FALSE, height=3.0>>=
 op <- par(cex=0.5)
@@
   \centering
-<<p2pmatrix, fig=TRUE, echo=FALSE, height=3.5>>=
+<<p2pmatrix, fig=TRUE, echo=FALSE, height=3.0>>=
 par(mar=c(3,8,1.1,1.1))

### 2011-11-25 inst/doc/article.Rnw
Final commit before Ritsert his comments

@@
 op <- par(cex=0.5)
-plot(c(0,0.8),c(0,400),type='n',main="QCL scores permutations",ylab="Frequency",xlab="Difference in correlation^2")
+plot(c(0,0.8),c(0,1000),type='n',main="QCL scores permutations",ylab="Frequency",xlab="Difference in correlation^2")
 hist(qcl_perms[[3]],add=T, col="yellow")

### 2011-12-16 R/qcl.image.R
R cmd check based fixes

@@ -46,7 +46,7 @@ internal.image <- function(mymatrix, colorrange, mainlabel, do.grid, grid.col){
   invisible(mymatrix)
 }
 
-imageQTLfromQCLobject <- function(x, onlySignificant = FALSE, significance = 0.05, do.grid=TRUE, grid.col = "black", verbose = FALSE, ...){
+QTLimage <- function(x, onlySignificant = FALSE, significance = 0.05, do.grid=TRUE, grid.col = "black", verbose = FALSE, ...){
   colorrange <- c("white",gray.colors(10)[10:1])
   mymatrix <- NULL
   mynames <- NULL

### 2011-12-16 inst/doc/article.Rnw
Reasonable time for testing

@@
   cat("QCL scan\n")
-  qcl_result <- QCLscan(ath.metab$genotypes, ath.metab$phenotypes)
-  cat("QTL scan\n")
-  qtl_result <- QTLscan(ath.metab$genotypes, ath.metab$phenotypes)
-  cat("QCL permutation\n")
-  qcl_perms1 <- QCLpermute(ath.metab$genotypes, ath.metab$phenotypes, 1:3, 200, n.cores = 3)
-  qcl_perms2 <- QCLpermute(ath.metab$genotypes, ath.metab$phenotypes, 4:24, 50, n.cores = 3)
-  qcl_perms <- c(qcl_perms1,qcl_perms2)
-  #qcl_perms <- vector("list",24)
-  #idx <- 1
-  #for(p in 1:24){
-  #  qcl_perms[[idx]] <- read.QCLpermute("permutations", p, 300,TRUE)
-  #  idx <- idx+1
-  #}
+  qcl_result <- QCLscan(ath.metab$genotypes, ath.metab$phenotypes, 1:24, n.perm=250)
 @

### 2011-12-16 tests/test_qcl_long.R

@@ -1,14 +1,17 @@
 library(qcl)
 data(ath.metabolites)
-library(qtl)
-data(multitrait)
-qcl_result <- QCLscan.cross(multitrait, 1:24)
-qcl_perms <- QCLpermute(ath.metab$genotypes, ath.metab$phenotypes,1:24,300)
+qcl_result <- QCLscan(ath.metab$genotypes, ath.metab$phenotypes, n.perm=100)
+
+jpeg("pxmmatrix.jpg",w=1024,h=768)
+image(qcl_result,grid.col="white")
+dev.off()
+
+jpeg("pxpmatrix.jpg",w=1024,h=768)
+image(qcl_result,grid.col="white",against="phenotypes")
+dev.off()
 
 for(x in 1:24){
   jpeg(paste("phenotype",x,".jpg",sep=""),w=1024,h=768)
-  op <- par(mfrow=c(2,1))
-  plot(QCLscantoScanone(multitrait, qcl_result, qcl_perms, pheno.col=x),scanone(multitrait,pheno.col=x),main="QCL versus QTL")
-  plotAsStackedHist(qcl_result, qcl_perms, pheno.col=x, main="Phenotypes contributing to QCL")
+  plot(qcl_result[[1]])  
   dev.off()
 }

### 2011-12-16 inst/doc/article.Rnw
R CMD check should be warning free

@@
 op <- par(cex=0.5)
-plot(c(0,0.8),c(0,1000),type='n',main="QCL scores permutations",ylab="Frequency",xlab="Difference in correlation^2")
-hist(qcl_perms[[3]],add=T, col="yellow")
-hist(qcl_perms[[2]],add=T, col="red")
-hist(qcl_perms[[1]],add=T, col="green")
-legend("topright",colnames(ath.metab$phenotypes)[1:3],col=c("yellow","red","green"),lwd=6)
+hist(qcl_result,c(3,2,1))
 op <- par(cex=1)
@@
 op <- par(cex=0.5)
-plotAsLOD(qcl_result, qcl_perms, pheno.col=1)
+plot(qcl_result[[1]])
 op <- par(cex=1)
@@
 op <- par(cex=0.5)
-plotAsStackedHist(qcl_result, qcl_perms, pheno.col=1)
+QCLasLOD(qcl_result[[1]], qcl_result[[1]]$qtl)
 op <- par(cex=1)
@@
 par(mar=c(3,8,1.1,1.1))
-image(qcl_result, qcl_perms, against="markers",do.grid=FALSE)
+plot(qcl_result, against="markers", do.grid=TRUE, grid.col="white")
 @ 
@@
 par(mar=c(3,8,1.1,1.1))
-image(1:ncol(qtl_result),1:nrow(qtl_result),t(qtl_result),col=c("white",gray.colors(100)[100:1]), yaxt="n", xaxt="n", ylab="", xlab="")
-axis(2,rownames(qtl_result),at=1:nrow(qtl_result),las=2,cex.axis=0.5)
-axis(1,colnames(qtl_result),at=1:ncol(qtl_result),las=2,cex.axis=0.3)
+QTLimage(qcl_result, do.grid=TRUE, grid.col="white")
 @
@@
 par(mar=c(3,8,1.1,1.1))
-image(qcl_result, 0.5, against="phenotypes")
+plot(qcl_result, against="phenotypes", do.grid=TRUE, grid.col="white")
 @ 

### 2011-12-16 R/qcl.hist.R

@@ -1,0 +1,22 @@
+#
+# qcl.hist.R
+#
+# copyright (c) 2010-2011 Danny Arends and Ritsert C. Jansen
+# last modified Dec, 2011
+# first written Dec, 2011
+# 
+# Histogram routines for QCL analysis
+#
+
+hist.QCLobject <- function(x, pheno.col=1, ...){
+  if(missing(x)) stop("argument 'x' which expects a 'QCLobject' object is missing, with no default")
+  namez <- NULL
+  plot(c(0,0.8),c(0,length(unlist(qcl_result[[pheno.col[1]]]$p))/5),type='n',main="QCL scores permutations",ylab="Frequency",xlab="Difference in correlation^2")
+  for(pheno in pheno.col){
+    if(is.null(x[[pheno]]$p)) stop(paste("Permutations not found for pheno.col=",pheno))
+    sorted <- sort(unlist(x[[pheno]]$p))
+    hist(sorted, breaks=seq(0,0.8,0.01), add=TRUE, col=pheno, ...)
+    namez <- c(namez,attr(x[[pheno]]$qcl,"name"))
+  }
+  legend("topright", namez, col=pheno.col,lwd=6)
+}

### 2011-12-16 R/qcl.statistics.R

@@ -143,12 +143,3 @@ plot.QCLpermute <- function(x, ...){
     idx <- idx+1
   }
 }
-
-hist.QCLpermute <- function(x, ...){
-  if(missing(x)) stop("argument 'x' which expects a 'QCLpermute' object is missing, with no default")
-  maximums <- lapply(x, function(v){
-    apply(abs(v), 2, max)
-  })
-  sorted <- sort(unlist(maximums))
-  hist(sorted,breaks=100,main="QCL scores during permutation",...)
-}

### 2011-12-16 R/qcl.print.R
R CMD check should be warning free

@@ -1,11 +1,11 @@
 #
 # qcl.print.R
 #
-# copyright (c) 2010 Danny Arends and Ritsert C. Jansen
-# last modified Oct, 2011
+# copyright (c) 2010-2011 Danny Arends and Ritsert C. Jansen
+# last modified Dec, 2011
 # first written Oct, 2011
 # 
-# Plotting routines for QCL analysis
+# Print routines for QCL analysis
 #
 
 print.QCLscan <- function(x, ...){

### 2011-12-16 R/qcl.plot.R
Minor

@@ -17,6 +17,7 @@ QCLasLOD <- function(QCLscan, QTLscores, main, do.legend=TRUE){
   plot.QCLscan(QCLscan,do.legend=do.legend)
   QCLscores <- QCLtoLODvector(QCLscan)
   if(!missing(QTLscores)){
+    QTLscores <- as.numeric(unlist(QTLscores))
     plot(c(0,length(QCLscores)),c(0,max(c(QCLscores,QTLscores))),type='n', main=main, ylab="LOD",xlab="Marker")
     points(QCLscores,type='l',col="black",lwd=3)
     points(QTLscores,type='l',col="red",lwd=2,lty=1)

### 2011-12-16 R/qcl.plot.R
pheno.col to plot.QCLobject

@@ -28,12 +28,14 @@ QCLasLOD <- function(QCLscan, QTLscores, main, do.legend=TRUE){
   }
 }
 
-plot.QCLobject <- function(x, ...){
+plot.QCLobject <- function(x, pheno.col=1:length(x) ...){
   if(length(x) == 1){
-    plot.QCLscan(x[[1]],...)
-  }else{
-    image.QCLobject(x,...)
+    return(plot.QCLscan(x[[1]],...))
   }
+  if(length(pheno.col) == 1){
+    return(plot.QCLscan(x[[pheno.col]],...))
+  }
+  return(image.QCLobject(x,...))
 }
 
 plot.QCLscan <- function(x, onlySignificant = TRUE, qcl.threshold =0.6, do.legend=TRUE, ...){

### 2011-12-16 R/qcl.plot.R
Moved a plot to QCL.plot.R also add legends

@@ -79,3 +79,17 @@ plot.QCLscan <- function(x, onlySignificant = TRUE, qcl.threshold =0.6, do.legen
   rownames(QCLmatrix) <- rownames(x$qcl)[mysign]
   invisible(QCLmatrix)
 }
+
+plot.QCLpermute <- function(x, type="s", ...){
+  if(missing(x)) stop("argument 'x' which expects a 'QCLpermute' object is missing, with no default")
+  plot(seq(0,0.9,0.01),QCLscoretoPvalue(seq(0,0.9,0.01),x),main="QCL to P.value",xlab="QCL",ylab="Pvalue", type=type,...)
+  significant <- print.QCLpermute(x)
+  mycolors <- c("red","orange","green")
+  idx <- 1
+  for(y in c(.05,.01,.001)){
+    lines(rbind(c(-1,y),c(significant[idx],y)),lty=2,col=mycolors[idx])
+    lines(rbind(c(significant[idx],-1),c(significant[idx],y)),lty=2,col=mycolors[idx])
+    idx <- idx+1
+  }
+  legend("topright",c("QCL-FDR: 5%","QCL-FDR: 1%","QCL-FDR: 0.1%"), col=mycolors, lty=2, lwd=1, cex=0.7)
+}

### 2011-12-16 R/qcl.statistics.R

@@ -130,16 +130,3 @@ QCLscantoScanone <- function(cross, QCLscan){
   scores[which(!is.finite(scores))] <- NA
   lodscorestoscanone(cross, scores)
 }
-
-plot.QCLpermute <- function(x, ...){
-  if(missing(x)) stop("argument 'x' which expects a 'QCLpermute' object is missing, with no default")
-  plot(seq(0,0.9,0.01),QCLscoretoPvalue(seq(0,0.9,0.01),x),main="QCL to P.value",xlab="QCL",ylab="Pvalue")
-  significant <- print.QCLpermute(x)
-  mycolors <- c("red","orange","green")
-  idx <- 1
-  for(y in c(.05,.01,.001)){
-    lines(rbind(c(-1,y),c(significant[idx],y)),lty=2,col=mycolors[idx])
-    lines(rbind(c(significant[idx],-1),c(significant[idx],y)),lty=2,col=mycolors[idx])
-    idx <- idx+1
-  }
-}

### 2012-01-05 R/qcl.hist.R
Minor bugin hist plotting

@@ -11,11 +11,11 @@
 hist.QCLobject <- function(x, pheno.col=1, ...){
   if(missing(x)) stop("argument 'x' which expects a 'QCLobject' object is missing, with no default")
   namez <- NULL
-  plot(c(0,0.8),c(0,length(unlist(x[[pheno.col[1]]]$p))/5),type='n',main="QCL scores permutations",ylab="Frequency",xlab="Difference in correlation^2")
+  plot(c(0,1.0),c(0,length(unlist(x[[pheno.col[1]]]$p))/5),type='n',main="QCL scores permutations",ylab="Frequency",xlab="Difference in correlation^2")
   for(pheno in pheno.col){
     if(is.null(x[[pheno]]$p)) stop(paste("Permutations not found for pheno.col=",pheno))
     sorted <- sort(unlist(x[[pheno]]$p))
-    hist(sorted, breaks=seq(0,0.8,0.01), add=TRUE, col=pheno, ...)
+    hist(sorted, breaks=seq(0,1.0,0.01), add=TRUE, col=pheno, ...)
     namez <- c(namez,attr(x[[pheno]]$qcl,"name"))
   }
   legend("topright", namez, col=pheno.col,lwd=6)

### 2012-01-05 inst/doc/article.Rnw
New dataset and resaved data

@@
   set.seed(200)
-  data(ath.metabolites) # Arabidopsis Thaliana dataset
-  cat("QCL scan\n")
-  qcl_result <- QCLscan(ath.metab$genotypes, ath.metab$phenotypes, 1:24, n.perm=250)
+  data(ath.result)
 @
@@
 op <- par(cex=0.5)
-hist(qcl_result,c(3,2,1))
+hist(ath.result,c(3,2,1))
 op <- par(cex=1)
@@
 op <- par(cex=0.5)
-plot(qcl_result[[1]])
+plot(ath.result[[1]])
 op <- par(cex=1)
@@
   \centering
-<<singletraitLOD, fig=TRUE, echo=FALSE, height=3.0>>=
+<<singletraitLOD, fig=TRUE, echo=FALSE, height=6.0>>=
 op <- par(cex=0.5)
-QCLasLOD(qcl_result[[1]], qcl_result[[1]]$qtl)
+QCLasLOD(ath.result[[1]], ath.result[[1]]$qtl)
 op <- par(cex=1)
@@
 par(mar=c(3,8,1.1,1.1))
-plot(qcl_result, against="markers", do.grid=TRUE, grid.col="white")
+plot(ath.result, against="markers", do.grid=TRUE, grid.col="white")
 @ 
@@
 par(mar=c(3,8,1.1,1.1))
-QTLimage(qcl_result, do.grid=TRUE, grid.col="white")
+QTLimage(ath.result, do.grid=TRUE, grid.col="white")
 @
@@
 par(mar=c(3,8,1.1,1.1))
-plot(qcl_result, against="phenotypes", do.grid=TRUE, grid.col="white")
+plot(ath.result, against="phenotypes", do.grid=TRUE, grid.col="white")
 @ 

### 2012-01-10 R/qcl.kegg.R
Code beautification

@@ -2,12 +2,11 @@
 # qcl.kegg.R
 #
 # copyright (c) 2010 Danny Arends
-# last modified Oct, 2011
+# last modified Jan, 2012
 # first written Oct, 2011
 # 
 # Download gene annotation from KEGG for QCL analysis
 #
-
 
 KEGG.pathway.parser <- function(HTMLdata, verbose = FALSE){
   possible <- regexpr("<nobr><a href=\"/kegg-bin/show_pathway(.+?)</tr>", HTMLdata,useBytes = TRUE)
@@ -65,3 +64,5 @@ download.KEGG.annotation <- function(genenames, species = "sce", directory = "tm
     warning(.has_rcurl_warnmsg)
   }
 }
+
+# end of qcl.kegg.R

### 2012-01-10 R/qcl.network.R

@@ -2,7 +2,7 @@
 # qcl.network.R
 #
 # copyright (c) 2010 Danny Arends and Bruno Tesson
-# last modified Oct, 2011
+# last modified Jan, 2012
 # first written Oct, 2011
 # 
 # Network routines for QCL analysis
@@ -120,3 +120,5 @@ write.nodeAttributeFile <- function(nodenames, spotAnnotation){
   }
   cat("Wrote annotation for",length(nodenames),"probes to node.attributes\n")
 }
+
+# end of qcl.network.R

### 2012-01-10 R/qcl.statistics.R

@@ -2,7 +2,7 @@
 # qcl.statistics.R
 #
 # copyright (c) 2010 Danny Arends, Bruno Tesson and Ritsert C. Jansen
-# last modified Nov, 2011
+# last modified Jan, 2012
 # first written Nov, 2011
 # 
 # R functions to do transform QCL mapping scores to Pvalues and LOD
@@ -130,3 +130,5 @@ QCLscantoScanone <- function(cross, QCLscan){
   scores[which(!is.finite(scores))] <- NA
   lodscorestoscanone(cross, scores)
 }
+
+# end of qcl.statistics.R

### 2012-01-10 R/qcl.permutation.R
Code beautification

@@ -2,8 +2,8 @@
 # qcl.permutations.R
 #
 # copyright (c) 2010 Danny Arends, Bruno Tesson and Ritsert C. Jansen
-# last modified Oct, 2011
-# first written nov, 2010
+# last modified Jan, 2012
+# first written Nov, 2010
 # 
 # R functions to do permutation on QCL mapping
 # Example data C. Elegans and available at request ( Danny.Arends@gmail.com )
@@ -84,3 +84,5 @@ QCLpermute.cross <- function(cross, pheno.col, method = c("pearson", "kendall", 
     warning(.has_rqtl_warnmsg)
   }
 }
+
+# end of qcl.permutations.R

### 2012-01-10 R/qcl.plot.R

@@ -2,7 +2,7 @@
 # qcl.plot.R
 #
 # copyright (c) 2010 Danny Arends and Bruno Tesson
-# last modified Oct, 2011
+# last modified Jan, 2012
 # first written Nov, 2010
 # 
 # Plotting routines for QCL analysis
@@ -93,3 +93,5 @@ plot.QCLpermute <- function(x, type="s", ...){
   }
   legend("topright",c("QCL-FDR: 5%","QCL-FDR: 1%","QCL-FDR: 0.1%"), col=mycolors, lty=2, lwd=1, cex=0.7)
 }
+
+# end of qcl.plot.R

### 2012-01-10 R/qcl.print.R

@@ -2,7 +2,7 @@
 # qcl.print.R
 #
 # copyright (c) 2010-2011 Danny Arends and Ritsert C. Jansen
-# last modified Dec, 2011
+# last modified Jan, 2012
 # first written Oct, 2011
 # 
 # Print routines for QCL analysis
@@ -41,3 +41,5 @@ getPermuteThresholds <-function(x, ..., verbose = FALSE){
 print.QCLpermute <- function(x, ...){
   getPermuteThresholds(x, ..., verbose=TRUE)
 }
+
+# end of qcl.print.R

### 2012-01-10 R/qcl.profiles.R

@@ -2,7 +2,7 @@
 # qcl.profiles.R
 #
 # copyright (c) 2011 Danny Arends and Ritsert C. Jansen
-# last modified Oct, 2011
+# last modified Jan, 2012
 # first written Oct, 2010
 # 
 #
@@ -52,3 +52,5 @@ QCLprofile <- function(QCLobject, pheno.col=1, significance = 0.05, verbose = TR
     return(result)
   }
 }
+
+# end of qcl.profiles.R

### 2012-01-10 R/qcl.scan.R

@@ -2,7 +2,7 @@
 # qcl.R
 #
 # copyright (c) 2010 Danny Arends and Ritsert C. Jansen
-# last modified Oct, 2011
+# last modified Jan, 2012
 # first written Jan, 2011
 # 
 # R functions to do QCL mapping
@@ -14,7 +14,7 @@ QCLscan <- function(genotypes, phenotypes, pheno.col = 1:ncol(phenotypes), metho
   if(missing(phenotypes)) stop("argument 'phenotypes' is missing, with no default")
   
   cat("Stage 0: Checking data\n")
-  toremove <- check.genotypes(genotypes,genotype.values)
+  toremove <- check.genotypes(genotypes, genotype.values, verbose)
   if(!is.null(toremove)) genotypes <- genotypes[,-toremove]
 
   results <- vector("list",length(pheno.col))
@@ -30,11 +30,11 @@ QCLscan <- function(genotypes, phenotypes, pheno.col = 1:ncol(phenotypes), metho
       cat("Stage 3: Permutation\n")
       results[[idx]]$p <- QCLpermute(genotypes, phenotypes, p, method=method, n.perm, n.cores, genotype.values, directory, saveFiles, verbose)
       
-      cat("Stage 4: Transformation into LOD\n")
+      if(verbose)cat("Stage 4: Transformation into LOD\n")
       results[[idx]]$l <- toLod(results[[idx]], FALSE, FALSE)
     }else{
       cat("Stage 3: Skipping permutation\n")
-      cat("Stage 4: Skipping transformation into LOD\n")
+      if(verbose)cat("Stage 4: Skipping transformation into LOD\n")
     }
     class(results[[idx]]) <- c(class(results[[idx]]),"QCLscan")
     idx <- idx + 1
@@ -81,3 +81,5 @@ QCLscan.cross <- function(cross, pheno.col, method = c("pearson", "kendall", "sp
     warning(.has_rqtl_warnmsg)
   }
 }
+
+# end of qcl.scan.R

### 2012-01-10 R/qcl.utils.R

@@ -2,12 +2,12 @@
 # qcl.utils.R
 #
 # copyright (c) 2010 Danny Arends and Ritsert C. Jansen
-# last modified Jun, 2011
+# last modified Jan, 2012
 # first written nov, 2010
 # 
 # check.genotypes, getRVM, lodscorestoscanone, getCorrelatedPhenotypes, gcLoop
 
-check.genotypes <- function(genotypes,genotype.values=c(1,2), verbose=FALSE){
+check.genotypes <- function(genotypes, genotype.values=c(1,2), verbose=FALSE){
   cat(" - Genotypes, ind=",nrow(genotypes),", markers=",ncol(genotypes),"\n",sep="")
   if(length(genotype.values)!=2) stop("argument 'genotype.values' length is incorrect, provide two genotype.values")
   toremove <- NULL
@@ -81,3 +81,5 @@ gcLoop <- function(verbose = FALSE){
     if(verbose) cat("GCloop ",n_usage," ",p_usage,"\n")
   }
 }
+
+# end of qcl.utils.R

### 2012-01-10 inst/doc/article.Rnw
Updates to article

@@
 \begin{figure}[ht]
-  \centerline{\includegraphics{QCL.pdf}}
-  \caption{QCL mapping: The underlying hypothesis of QCL mapping. Seen here is a network of 3 phenotypes, in 
-  which expression of P1, P2 and P3 is measured, and we want to infer the hidden relation between P1 
-  and P2. In our model a genetic pertubation is causing P1 to change in expression showing a QTL at a marker. We expect variation 
-  induced by P1 to transfer to P2, However (like in most biological systems) expression of P2 is tightly controlled, 
-  and variation seen in phenotype 1 wil not induce a QTL in phenotype 2. However when we partition individuals based 
-  on the genotype at the QTL marker of P1, we can make use of the correlation structure between P1 and P2. Allowing 
-  us to infer the relationship between phenotype 1 and 2, when we see difference in correlation between P1A and P2A 
-  and P1B and P2B. We don't observe this difference in correlation between P3 and P1, meaning there is no relation 
-  between P1 and P3}
-  \label{QCL_img}
+  \centerline{\includegraphics[height=7cm]{qcl.pdf}}
+  \vspace{-15pt}
+  \caption{{\scriptsize QCL mapping: We start by looking at a single genetic marker and dividing our populaiton into two groups.
+  We start by choosing a phenotype (red), and use its expression to detect the red module (high correlation in gray genotype). 
+  We see that something underlying our current marker in the gray genotype regulates the expression of red (but not blue)}}
+  \vspace{-20pt}
+  \label{fig:qcl}
+\end{figure}
+
+\begin{figure}[ht]
+  \centerline{\includegraphics[height=6cm]{qcl_n.pdf}}
+  \vspace{-15pt}
+  \caption{{\scriptsize Reconstructed pathway location from gene expression data: Insight into the underlying (hidden) regulatory network 
+  can be obtained when we look at the amount of expression showing the same switch. Depending on the amount of blue expressions 
+  co-affected we can guestimate the location of the red expression inside the pathway}}
+  \vspace{-20pt}
+  \label{fig:qcl_n}
 \end{figure}
@@
 @ 
-  \caption{QCL scores obtained during permutations (A. Thaliana) additional information obtained by QCL mapping, 
-  which can be used to reconstruct the underlying genetic architecture.\label{fig:histPermAra}}
+  \caption{{\scriptsize QCL scores obtained during permutations (A. Thaliana) additional information obtained by QCL mapping, 
+  which can be used to reconstruct the underlying genetic architecture. }}
+  \label{fig:histPermAra}
 \end{figure}
@@
 @
-  \caption{QCL profile (A. Thaliana, X3.Hydroxypropyl) transformed to LOD scores compared to the QTL profile of the same phenotype, We see an 
-  additional significant peek for this phenotype (Between Marker 50 to 60) \label{fig:singletrait}}
+  \caption{{\scriptsize QCL profile (A. Thaliana, X3.Hydroxypropyl) transformed to LOD scores compared to the QTL profile of the same phenotype, We see an 
+  additional significant peek for this phenotype (Between Marker 50 to 60) }}
+  \label{fig:singletrait}
 \end{figure}
@@
 @ 
-  \caption{QCL profile (A. Thaliana, X3.Hydroxypropyl) transformed to LOD scores. Shown is the likelyhood contribution of each of 
-  the 24 metabolites to the QCL profile. \label{fig:singletraitLOD}}
+  \caption{{\scriptsize QCL profile (A. Thaliana, X3.Hydroxypropyl) transformed to LOD scores. Shown is the likelyhood contribution of each of 
+  the 24 metabolites to the QCL profile.}}
+  \label{fig:singletraitLOD}
 \end{figure}
@@
 @ 
-  \caption{Phenotypes x Markers matrix (A. Thaliana) this is 'comparable' to a QTL scan on multiple traits.\label{fig:p2mmatrix}}
+  \caption{{\scriptsize Phenotypes x Markers matrix (A. Thaliana) this is 'comparable' to a QTL scan on multiple traits.}}
+  \label{fig:p2mmatrix}
 \end{figure}
@@
 @
-  \caption{Heatmap showing QTL scores for all 24 metabolities in A. Thaliana.\label{fig:QTLmatrix}}
+  \caption{{\scriptsize Heatmap showing QTL scores for all 24 metabolities in A. Thaliana.}}
+  \label{fig:QTLmatrix}
 \end{figure}
@@
 @ 
-  \caption{Phenotypes x Phenotypes matrix (A. Thaliana) additional information obtained by QCL mapping, 
-  which can be used to reconstruct the underlying genetic architecture.\label{fig:p2pmatrix}}
-\end{figure}
-
-\begin{figure}[ht]
-  \centerline{\includegraphics{AraNet.pdf}}
-  \caption{Reconstructed network for chain elongation in Arabidopsis Thaliana}
-  \label{AraNet}
+  \caption{{\scriptsize Phenotypes x Phenotypes matrix (A. Thaliana) additional information obtained by QCL mapping, 
+  which can be used to reconstruct the underlying genetic architecture}}
+  \label{fig:p2pmatrix}
 \end{figure}

### 2012-01-30 R/qcl.plot.R
Old R updates

@@ -38,7 +38,7 @@ plot.QCLobject <- function(x, pheno.col=1:length(x), ...){
   return(image.QCLobject(x,...))
 }
 
-plot.QCLscan <- function(x, onlySignificant = TRUE, qcl.threshold =0.6, do.legend=TRUE, ...){
+plot.QCLscan <- function(x, addQTL = TRUE, onlySignificant = TRUE, qcl.threshold =0.6, do.legend=TRUE, ...){
   if(missing(x)) stop("argument 'x' is missing, with no default")
   if(!is.null(x$p)){
     mysign <- as.numeric(which(apply(abs(x$qcl),1,max) > getPermuteThresholds(x)[1]))
@@ -55,7 +55,7 @@ plot.QCLscan <- function(x, onlySignificant = TRUE, qcl.threshold =0.6, do.legen
     QCLmatrix <- matrix(x$l[mysign, ],length(mysign),ncol(x$l))
   }
   summarized <- apply(QCLmatrix,2,sum)
-  plot(c(0,ncol(x$qcl)),c(0,max(summarized)), type='n',xlab="Marker", ylab="-log10(P-value)", main=paste("Phenotype contribution to QCL of",attr(x$qcl,"name")),...)
+  plot(c(0,ncol(x$qcl)),c(0,max(c(summarized,x$qtl))), type='n',xlab="Marker", ylab="-log10(P-value)", main=paste("Phenotype contribution to QCL of",attr(x$qcl,"name")),...)
   p <- rep(0,ncol(x$qcl))
   i <- 1;
   mycolors <- topo.colors(nrow(QCLmatrix))
@@ -69,13 +69,15 @@ plot.QCLscan <- function(x, onlySignificant = TRUE, qcl.threshold =0.6, do.legen
     }
   )
   if(!is.null(x$l)){
-    abline(h=-log10(c(0.05,0.01,0.001)),col=c("red","orange","green"),lty=2)
+    n <- dim(x$qcl)[2]
+    abline(h=-log10(c(0.05/n,0.01/n,0.001/n)),col=c("red","orange","green"),lty=2)
     legend("topright",as.character(paste("QCL-FDR:",c(0.05,0.01,0.001),"%")),col=c("red","orange","green"),lty=rep(2,3),lwd=1,cex=0.7)
   }
   if(do.legend){
     legend("topleft",rownames(x$qcl)[mysign],col=mycolors,lwd=1,cex=0.7)
   }
[…]

### 2012-01-30 R/qcl.print.R

@@ -17,17 +17,23 @@ print.QCLscan <- function(x, ...){
 }
 
 getPermuteThresholds <-function(x, ..., verbose = FALSE){
-  if(!any(class(x)=="QCLpermute")) x <- x$p
-  if(!any(class(x)=="QCLpermute")) stop("No permutations found in the QCLscan object")
+  if(any(class(x)=="QCLscan")){
+    n <- dim(x$qcl)[2]
+    x <- x$p
+    if(!any(class(x)=="QCLpermute")) stop("No permutations found in the QCLscan object")
+  }else{
+    stop("No QCLscan object")
+  }
   sorted <- sort(unlist(x))
   l <- length(sorted)
   values <- NULL
   valnames <- NULL
   for(x in c(.95,.99,.999)){
-    if(1/(1-x) < length(sorted)){
-    values <- c(values,sorted[l*x])
+    if(1/((1-x)/n) < length(sorted)){
+    v <- sorted[l*(1-(1-x)/n)]
+    values <- c(values,v)
     valnames <- c(valnames,paste((1-x)*100,"%"))
-    if(verbose) cat((1-x)*100,"%\t",sorted[l*x],"\n")
+    if(verbose) cat((1-x)*100,"%\t",v,"\n")
     }else{
     values <- c(values,NaN)
     valnames <- c(valnames,paste((1-x)*100,"%"))

### 2012-01-30 R/qcl.permutation.R
Old R updates

@@ -40,7 +40,7 @@ QCLpermute.internal <- function(perm, genotypes, phenotypes, pheno.col, method =
   if(saveFiles) write.table(myperm, file=paste(directory,"/Permutation_",pheno.col,"_",perm,".txt",sep=""))
   el <- proc.time()
   if(verbose) cat("  - Permutation",perm,"took:",as.numeric(el[3]-sl[3]),"seconds.\n")
-  as.numeric(max(abs(myperm)))
+  as.numeric(apply(abs(myperm),2,max))
 }
 
 #-- QCLpermute main function --#
@@ -57,7 +57,7 @@ QCLpermute <- function(genotypes, phenotypes, pheno.col, method = c("pearson", "
     for(p in pheno.col){
       ss <- proc.time()
       cat("  - Starting permutation for phenotype",p,"\n")
-      #Generate random number from a single thread, so we don't run into concurrency issues
+      #Generate random numbers from a single thread, so we don't run into concurrency issues
       rvm <- getRVM(n.perm,nrow(genotypes))
       for(x in 1:n.perm){
         QCLperm[[idx]] <- c(QCLperm[[idx]],QCLpermute.internal(x,genotypes,phenotypes,p,method,rvm,genotype.values,directory,saveFiles,verbose))

### 2012-01-30 inst/methodPaper/document.tex
Document was not added

\documentclass{bioinfo}
\usepackage{subfigure}
\usepackage{amsmath}

\copyrightyear{2012}
\pubyear{2012}

\begin{document}
\firstpage{1}
\title[CTLmapping]{CTL mapping - The added value from segregating correlation}
\author[Arends \emph{et al.}]{
  Danny Arends\,$^{1,\footnote{to whom correspondence should be addressed}}$,
  and Ritsert C. Jansen\,$^{1}$
}
\address{
  $^{1}$Groningen Bioinformatics Centre, University of Groningen, Groningen, The Netherlands.
}
\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}
\editor{Associate Editor: XXXXXXX}
\maketitle
\begin{abstract}
  \section{Motivation:}
  Here we present the concept of Correlated Traits Locus (CTL) mapping, a novel approach 
  to detect genetic regulation of phenotypes in recombinant inbred line populations (RIL) 
  or SNP genotypes populations. CTL mapping allows association of phenotype correlation 
  differences to genetic markers, in almost the same way as QTL mapping associates genetic 
  markers with differences in expression.

  We use CTL mapping to reconstruct underlying genetic architecture, and provide insight 
  into regulation of (gene) expression. CTL information in combination with classical QTL 
  information leads to a more detailed insight into the genetic architecture underlying 
  complex traits. CTL mapping provides a missing link in regulatory network reconstruction.

  \section{Results:}
  The software is available under the terms of GNU General Public License, version 3
  (\href{http://www.r-project.org/Licenses/GPL-3}{www.r-project.org/Licenses/GPL-3}).
  \section{Availability:}
  Software, documentation and source code are available from:
  \href{http://www.mapctl.org}{www.mapctl.org}
  \section{Contact:} 
  \href{Danny.Arends@Gmail.com}{Danny.Arends@Gmail.com}
\end{abstract}

\section{Introduction}
 Genetical genomics experiments have shown that gene expression variation can be mapped to 
  genotypic variation \cite{Jansen:2001}. Quantitative Trait Locus (QTL) mapping of a 
  gene identifies regions in the genome for which different genotypes lead to differential 
  expression level of the gene. This QTL mapping can be done on all molecular levels including, 
  gene expression (eQTL), protein abundance (pQTL) and metabolites (mQTL).

  We present Correlated Traits Locus (CTL) mapping as a comparable method to QCL mapping, 
  CTL unlike QTL associates differential correlation to genetic variation, i.e. to identify regions in the 
  genome for which one genotype value leads to correlated expression between a phenotype and all 
  other phenotypes, exploiting the fact that any QTL will be the result of either direct induced 
  variation or indirect variation, which will be visible as correlation differences among 
  individuals carrying different genotypes.

  We merge information coming from CTL and QTL to reveal underlying genetic architecture, which 
  would remain hidden when using only classical QTL information.

\section{Methodology}
  Choose a phenotype to subject to CTL analysis, this can be your favourite gene, 
  protein, metabolite or classical phenotype.

  Split the population at a genetic marker in two groups (aa and bb). Calculate the correlation 
  between our phenotype (i) and all other phenotypes. Calculate the CTL matrix by taking the 
  squared difference between matrix A and B.

  Single trait CTL profiles can be obtained by transforming the CTL scores into LOD scores. 
  For this we use a look-up table of scores obtained during 15.000 permutation. We note that for each trait this 
  shows a different distribution. We assess the likelyhood and take the $-log10(P)$ of every CTL score to 
  compare QTL profiles with summarized CTL profiles. A CTL profile show the contribution of each phenotype to 
  the total likelyhood.

  \begin{figure*}[ht]
    \begin{center}
    \vspace{-0.2cm}
    \hspace{-1.3cm}
    \subfigure[CTL profiles]{
      \includegraphics[keepaspectratio,scale=.26]{img/ctl_qtl.pdf}
      \label{subfiga}
    }
    \hspace{0.3cm}
    \subfigure[Phenotype x Phenotype interaction matrix]{
      \includegraphics[keepaspectratio,scale=.65]{img/ctl_pxp.pdf}
      \label{subfigb}
    }
    \hspace{0.3cm}
    \subfigure[Reconstructed network]{
      \includegraphics[keepaspectratio,scale=.17]{img/ctl_net.pdf}
      \label{subfigc}
    }
  \end{center}
  \begin{minipage}{7in}
  \label{newplots}
  \vspace{-0.7cm}
  \caption[Plots]{
  {\emph {\bf a}}: Example CTL + QTL profiles in Red we see the QTL profile on top of the CTL profile 
  (which also shows the names of the underlying traits contributing to the CTL). {\emph {\bf b}}: 
  Phenotype x Phenotype interaction matrix (PxP matrix). Information obtained using CTL mapping, showing 
  for each phenotype the significance of the difference in correlation to all other phenotypes. 
  This matrix can be transformed into .sif file to generate a network image. {\emph {\bf c}}: Reconstructed 
  network, this shows the same information as the PxP matrix only in a more accessible way, we visualized our 
  network using cytoscape, the linewidth is the summarized interaction evidence.
 }
\end{minipage}
\vspace{-0.5cm}
\end{figure*}

\emph{ {\bf Network reconstruction}}
  To build up gene interaction networks, we take the Phenotype x Marker and Phenotype x Phentype 
  relations matrices.  From the PxM we can identify partners of our phenotype, and from the PxP we 
  identify the location of out phenotype in the pathway or main regulatory phenotypes.

\emph{ {\bf Significance testing}}
  Comparing CTL profiles to QTL profiles CTL scores need to be converted to LOD scores. Significance 
  of CTL is assessed by using permutations, in each round the link between genotype and 
  phenotype is broken, by redistributing at random genotypes amongst individuals not 
  allowing for duplicates. The CTL on these new individuals allows for estimation of 
  the underlying 0-distribution of CTL scores. When observe CTL scores in 
  real data which are higher than any CTL score obtained during permutation we 
  use a generalized Pareto distribution (GPD) to estimate the extreme tail 
  of the distribution \cite{Knijnenburg:2009}, to get likelyhood estimates for the 
  extreme scores observed.

\emph{ {\bf Materials}}
  Metabolite data from an \emph{A. Thaliana} Landsberg erecta (Ler) and Cape Verde Islands (Cvi) 
  dataset from R/qtl, 24 characterized metabolites which are the products of chain elongation. 
  There are 162 recombinant inbred lines genotyped at 117 markers on 5 chromosomes. Additional 
  details about the experimental settings are available in the original paper \cite{Keurentjes:2006,AlonsoBlanco:1998}
\vspace{-0.6cm}
\section{Results}
  CTL mapping was performed and a per trait CTL thresholds of $P < 0.001$ was determined by using 15.000 permutations.
  Information obtained by CTL mapping is shown in the Phenotype to Phenotype interaction matrix (Fig. \ref{subfigb}).
  The CTL profiles were combined with QTL information, leading to the following observations \ref{subfiga}:\\
  1) Additional CTL found, where no QTL was present. Loss of variation in the genes 
  underneath the CTL are not causing variation differences in the observed phenotype, CTL is able to 
  detect the differences in correlation.\\
  2) CTL co-localized with QTL, in this situation CTL gives a detailed view of which phenotypes 
  are contributing variation to the observed QTL.\\
  3) QTL with no CTL, we interpret this as the source of the variation. Analysis in this way of 
  all 24 Single trait CTL profiles combined with their QTL profiles allows us to reconstruct the 
  source of natural variation and follow it down the network.

\emph{ {\bf Network reconstruction}}
  Using the previously decribed method to draw the arrows between traits, we reconstruct the interaction 
  network amongst the 24 metabolities in Arabidopsis (Fig. \ref{subfigc}). We observe that our network is 
  composed of 2 parts which seem connected only by a just significant CTL from X3.Benzoyloxypropyl to 
  Quercetin.deoxyhexosyl.dihexoside.

\section{Conclusion \& Discussion}
  Microarray experiments have identified genes differentially expressed between various 
  conditions, or groups of genes coexpressed across samples. Genetical genomics screens 
  are carried out to map differential expression to genetic variation. Screening 
  the genome for genotype combinations for which there is differential expression is 
  a successful strategy when mapping Mendelian traits.

  Results on complex traits however have been less then satisfactory with only small 
  amounts of heritability explained by individuals loci. CTL mapping 'borrows'
  information from all phenotypes in the analysis, and extracts phenotype to phenotype 
   information.

  Furthermore CTLs seem to 'transfer' from phenotype to phenotypes in a stabile 
  fashion, allowing reconstructing of genetic architecture. CTL combined with QTL 
  data identify gene transcript introducing variation (by absense), while co-localizing 
  with QTL affected by upstream variation.

\paragraph{Funding\textcolon}
The Centre for BioSystems Genomics (CBSG) and the Netherlands Consortium of Systems 
Biology (NCSB), both of which are part of the Netherlands Genomics Initiative / 
Netherlands Organisation for Scientiﬁc Research [to DA]; the EU 7th Framework 
Programme under the Research Project PANACEA [222936 to R.J.].
\vspace{-0.7cm}
\bibliographystyle{natbib}
\bibliography{document}
\end{document}

### 2012-01-30 inst/methodPaper/document.tex
CTL, and phenotype -> trait

@@
   Single trait CTL profiles can be obtained by transforming the CTL scores into LOD scores. 
-  For this we use a look-up table of scores obtained during 15.000 permutation. We note that for each trait this 
-  shows a different distribution. We assess the likelyhood and take the $-log10(P)$ of every CTL score to 
-  compare QTL profiles with summarized CTL profiles. A CTL profile show the contribution of each phenotype to 
+  For this we use permutations obtained during 15.000 permutation. We note that for each trait this 
+  shows a different distribution of permutation scores. We assess the likelyhood (per trait) and 
+  take the $-log10(P)$ of every CTL score to compare QTL profiles with summarized CTL profiles. A CTL profile show the contribution of each phenotype to 
   the total likelyhood.

### 2012-01-30 inst/methodPaper/document.tex
Phenotype -> trait

@@
   Here we present the concept of Correlated Traits Locus (CTL) mapping, a novel approach 
-  to detect genetic regulation of phenotypes in recombinant inbred line populations (RIL) 
-  or SNP genotypes populations. CTL mapping allows association of phenotype correlation 
+  to detect genetic regulation of traits in recombinant inbred line populations (RIL) 
+  or SNP genotypes populations. CTL mapping allows association of trait correlation 
   differences to genetic markers, in almost the same way as QTL mapping associates genetic 
@@
   CTL unlike QTL associates differential correlation to genetic variation, i.e. to identify regions in the 
-  genome for which one genotype value leads to correlated expression between a phenotype and all 
-  other phenotypes, exploiting the fact that any QTL will be the result of either direct induced 
+  genome for which one genotype value leads to correlated expression between a trait and all 
+  other traits, exploiting the fact that any QTL will be the result of either direct induced 
   variation or indirect variation, which will be visible as correlation differences among 
@@
 \section{Methodology}
-  Choose a phenotype to subject to CTL analysis, this can be your favourite gene, 
-  protein, metabolite or classical phenotype.
+  Choose a trait to subject to CTL analysis, this can be your favourite gene, 
+  protein, metabolite or classical trait.
 
   Split the population at a genetic marker in two groups (aa and bb). Calculate the correlation 
-  between our phenotype (i) and all other phenotypes. Calculate the CTL matrix by taking the 
+  between our trait of interest and other traits. Calculate the CTL matrix by taking the 
   squared difference between matrix A and B.
@@
   shows a different distribution of permutation scores. We assess the likelyhood (per trait) and 
-  take the $-log10(P)$ of every CTL score to compare QTL profiles with summarized CTL profiles. A CTL profile show the contribution of each phenotype to 
-  the total likelyhood.
+  take the $-log10(P)$ of every CTL score to compare QTL profiles with summarized CTL profiles. 
+  A CTL profile show the contribution of each trait to the total likelyhood.
 
@@
     \hspace{0.3cm}
-    \subfigure[Phenotype x Phenotype interaction matrix]{
+    \subfigure[Traits x Traits interaction matrix]{
       \includegraphics[keepaspectratio,scale=.65]{img/ctl_pxp.pdf}
@@
   \caption[Plots]{
-  {\emph {\bf a}}: Example CTL + QTL profiles in Red we see the QTL profile on top of the CTL profile 
-  (which also shows the names of the underlying traits contributing to the CTL). {\emph {\bf b}}: 
-  Phenotype x Phenotype interaction matrix (PxP matrix). Information obtained using CTL mapping, showing 
-  for each phenotype the significance of the difference in correlation to all other phenotypes. 
-  This matrix can be transformed into .sif file to generate a network image. {\emph {\bf c}}: Reconstructed 
-  network, this shows the same information as the PxP matrix only in a more accessible way, we visualized our 
+  {\emph {\bf a}}: Example CTL + QTL profiles (red). Seen are the QTL profile on top of the CTL profile. 
+  The first peak is a CTL caused by X4.Methylsulfinylbutyl and X5.Methylthiopenty showing a significant 
+  difference in correlation, we don't observe a QTL for this location. 
+  {\emph {\bf b}}: Traits x Traits interaction matrix (TxT matrix). Information obtained using CTL mapping, 
+  showing for each trait the significance of the difference in correlation to all other traits. This 
+  matrix can be transformed into .sif file to generate a network image. 
+  {\emph {\bf c}}:  Reconstructed network, this shows the same information as the TxT matrix only in a more accessible way, we visualized our 
   network using cytoscape, the linewidth is the summarized interaction evidence.
@@
 \emph{ {\bf Network reconstruction}}
-  To build up gene interaction networks, we take the Phenotype x Marker and Phenotype x Phentype 
-  relations matrices.  From the PxM we can identify partners of our phenotype, and from the PxP we 
-  identify the location of out phenotype in the pathway or main regulatory phenotypes.
+  To build up gene interaction networks, we take the Traits x Marker and Traits x Phentype 
+  relations matrices.  From the TxM we can identify partners of our trait, and from the PxP we 
+  identify the location of out trait in the pathway or main regulatory traits.
 
@@
   of CTL is assessed by using permutations, in each round the link between genotype and 
-  phenotype is broken, by redistributing at random genotypes amongst individuals not 
+  trait is broken, by redistributing at random genotypes amongst individuals not 
   allowing for duplicates. The CTL on these new individuals allows for estimation of 
@@
   CTL mapping was performed and a per trait CTL thresholds of $P < 0.001$ was determined by using 15.000 permutations.
-  Information obtained by CTL mapping is shown in the Phenotype to Phenotype interaction matrix (Fig. \ref{subfigb}).
+  Information obtained by CTL mapping is shown in the trait to trait interaction matrix (Fig. \ref{subfigb}).
   The CTL profiles were combined with QTL information, leading to the following observations \ref{subfiga}:\\
   1) Additional CTL found, where no QTL was present. Loss of variation in the genes 
-  underneath the CTL are not causing variation differences in the observed phenotype, CTL is able to 
+  underneath the CTL are not causing variation differences in the observed trait, CTL is able to 
   detect the differences in correlation.\\
-  2) CTL co-localized with QTL, in this situation CTL gives a detailed view of which phenotypes 
+  2) CTL co-localized with QTL, in this situation CTL gives a detailed view of which traits 
   are contributing variation to the observed QTL.\\
@@
   amounts of heritability explained by individuals loci. CTL mapping 'borrows'
-  information from all phenotypes in the analysis, and extracts phenotype to phenotype 
+  information from all traits in the analysis, and extracts trait to trait 
    information.
 
-  Furthermore CTLs seem to 'transfer' from phenotype to phenotypes in a stabile 
+  Furthermore CTLs seem to 'transfer' from trait to traits in a stabile 
   fashion, allowing reconstructing of genetic architecture. CTL combined with QTL 

### 2012-01-30 R/correlations.R
Cleaning up: stuff that doesn't need to be here

@@ -1,52 +1,0 @@
-#
-# correlations.R
-#
-# copyright (c) 2011 Danny Arends and Ritsert C. Jansen
-# last modified Oct, 2011
-# first written Oct, 2011
-# 
-# R->D functions to do correlations
-#
-
-correlation <- function(x, y, verbose = FALSE){
-  if(is.matrix(x) && dim(x)[2] > 1){
-    return(matrix_correlation(x,verbose=verbose))
-  }else{
-    if(!get(".has_d", envir = .QclEnv)){
-      if(verbose) cat(.has_d_warnmsg)
-      return(cor(x,y))
-    }else{
-      if(missing(y))stop("'y' is missing");
-      if(length(x) == length(y)){
-        return(.C("correlation_v",
-          x=as.numeric(x),
-          y=as.numeric(y),
-          as.integer(length(x)),
-          cor=as.numeric(0),
-          PACKAGE="Dcode")$cor
-        )
-      }else{
-        stop("The length of 'x' is not equal to the length of 'y'")
-      }
-    }
-  }
-}
-
-matrix_correlation <- function(x, verbose = FALSE){
-  if(!get(".has_d", envir = .QclEnv)){
-    if(verbose) cat(.has_d_warnmsg)
-    return(cor(x))
-  }else{
-    res <- .C("correlation_m",
-      nrx=as.integer(nrow(x)),
-      ncx=as.integer(ncol(x)),
-      x=as.numeric(x),
-      cor=as.numeric(rep(1,ncol(x)*ncol(x))),
-      PACKAGE="Dcode"
-    )
-    mcor = matrix(res$cor,ncol(x),ncol(x))
-    return(mcor)
-  }
-}
-
-# end of correlations.R

### 2012-01-30 tests/test_correlation.R

@@ -1,49 +1,0 @@
-require(qcl)
-
-depth <- 3
-variables <- c(10,50,100,250,500)
-observations <- c(100,200,300,400,500)
-
-cat("Reporting:", length(variables),"times\n")
-baseline <- vector("list",length(variables))
-results <- vector("list",length(variables))
-cnt <- 1
-for(myvar in variables){
-  for(myobs in observations){
-    matri <- matrix(runif(myvar*myobs),myvar,myobs)
-    
-    measurement <- NULL
-    for(x in 1:depth){ 
-      s <- proc.time()
-      res1 <- cor(matri); 
-      e <- proc.time()
-      measurement <- c(measurement,e[3]-s[3])
-    }
-    baseline[[cnt]] <- rbind(baseline[[cnt]],measurement)
-
-    measurement <- NULL
-    for(x in 1:depth){ 
-      s <- proc.time()
-      res2 <- correlation(matri); 
-      e <- proc.time()
-      measurement <- c(measurement,e[3]-s[3])
-    }
-    if(!(sum(res1-res2) < 1e-10)){
-      stop("Correlations deviate too much")
-    }else{
-      cat("Tested:",myobs,":",myvar,"\n")
-    }
-    results[[cnt]] <- rbind(results[[cnt]],measurement)
-  }
-  rownames(results[[cnt]]) <- observations
-  colnames(results[[cnt]]) <- 1:depth
-  rownames(baseline[[cnt]]) <- observations
-  colnames(baseline[[cnt]]) <- 1:depth
-  cat("Done:",cnt,"\n")
-  cnt <- cnt+1
-}
-names(results) <- variables
-names(baseline) <- variables
-
-boxplot(results,col=c("red"))
-boxplot(baseline,col=c("green"),add=T)

### 2012-01-30 R/qcl.kegg.R
Cleaning up: stuff that doesn't need to be here

@@ -1,68 +1,0 @@
-#
-# qcl.kegg.R
-#
-# copyright (c) 2010 Danny Arends
-# last modified Jan, 2012
-# first written Oct, 2011
-# 
-# Download gene annotation from KEGG for QCL analysis
-#
-
-KEGG.pathway.parser <- function(HTMLdata, verbose = FALSE){
-  possible <- regexpr("<nobr><a href=\"/kegg-bin/show_pathway(.+?)</tr>", HTMLdata,useBytes = TRUE)
-  mydata <- NULL
-  while(possible > 0){
-    ppathway <- substr(HTMLdata, possible[[1]], possible[[1]]+attr(possible,"match.length")-11)
-    start <- regexpr("<td align=\"left\">",ppathway)[[1]]+17
-    ppathway <- substr(ppathway,start,nchar(ppathway))
-    if(verbose) cat(" - Possible pathway: ",ppathway,"\n")
-    mydata <- rbind(mydata,ppathway)
-    HTMLdata <- substr(HTMLdata,possible[[1]]+attr(possible,"match.length")-1,nchar(HTMLdata))
-    possible <- regexpr("<nobr><a href=\"/kegg-bin/show_pathway(.+?)</tr>", HTMLdata,useBytes = TRUE)
-  }
-  invisible(mydata)
-}
-
-download.KEGG.annotation <- function(genenames, species = "sce", directory = "tmp_kegg", force.download=FALSE, verbose = FALSE){
-  if(get(".has_rcurl", envir = .QclEnv)){
-    require("RCurl")
-    if(missing(genenames)) stop("No gene names to get annotation for supplied")
-    cnt <- 1
-    p_cnt <- 1
-    if(!file.exists(directory)) dir.create(directory)
-    cat("",file="node.kegg.attributes")
-    annotations <- vector("list",length(genenames))
-    for(gene_name in genenames){
-      file_out_name <- paste(directory,"/kegg_",gene_name,".txt",sep="")
-      if(file.exists(file_out_name) && !force.download){
-        HTMLdata <- paste(scan(file_out_name,what="character",quiet=TRUE),collapse=" ")
-        if(verbose) cat("Loaded ",cnt,":",gene_name," from ",file_out_name,"\n",sep="")
-      }else{
-        url <- paste("http://www.kegg.jp/entry/",species,"/",gene_name,sep="")
-        HTMLdata <- RCurl::getURL(url)
[…]

### 2012-01-30 R/qcl.print.R

@@ -12,7 +12,7 @@ print.QCLscan <- function(x, ...){
   cat("QCLscan summary",attr(x$qcl,"name"),"\n\n")
   cat("- Number of background phenotypes",dim(x$qcl)[1],"\n")
   cat("- Number of markers",dim(x$qcl)[2],"\n")
-  cat("- Number of permutations",length(unlist(x$p)),"\n")
+  cat("- Number of permutations",length(unlist(x$p))/dim(x$qcl)[2],"\n")
   getPermuteThresholds(x,..., verbose = TRUE)
 }
 
@@ -22,7 +22,10 @@ getPermuteThresholds <-function(x, ..., verbose = FALSE){
     x <- x$p
     if(!any(class(x)=="QCLpermute")) stop("No permutations found in the QCLscan object")
   }else{
-    stop("No QCLscan object")
+    n <- 1
+    if(!any(class(x)=="QCLpermute")){
+      stop("No QCLscan object")
+    }
   }
   sorted <- sort(unlist(x))
   l <- length(sorted)

### 2012-01-30 R/zzz.R

@@ -12,40 +12,27 @@
 .QclEnv <- new.env()
 
 get_QclEnv <- function(){.QclEnv}
-has_d <- function(){ get(".has_d", envir = .QclEnv) }
 has_snow <- function(){ get(".has_snow", envir = .QclEnv) }
-has_rcurl <- function(){ get(".has_rcurl", envir = .QclEnv) }
 has_rqtl <- function(){ get(".has_rqtl", envir = .QclEnv) }
 
-.has_d_warnmsg       <- "- D not available, using standard R/C/C++ functionality\n"
 .has_snow_warnmsg    <- "- Package: SNOW not found, multi CPU support is not available/disabled\n"
-.has_rcurl_warnmsg   <- "- Package: RCURL not found, automatic KEGG pathway annotations not available/disabled\n"
 .has_rqtl_warnmsg    <- "- Package: qtl not found, R/qtl functionality not available/disabled\n"
 
 #Package loading
 .onAttach <- function(lib, pkg){
   packageStartupMessage("- Loading package qcl\n", appendLF = FALSE)
-  .has_d <- TRUE
   .has_snow <- FALSE
-  .has_rcurl <- FALSE
   .has_rqtl <- FALSE
-  library.dynam("qcl", pkg, lib)
-  tryCatch(
-    library.dynam("Dcode", pkg, lib),
-    error = function(e){
-     .has_d <<- FALSE
-   })
   .has_snow  <- ("snow" %in% installed.packages(lib.loc=lib)[,1])
-  .has_rcurl <- ("RCurl" %in% installed.packages(lib.loc=lib)[,1])
   .has_rqtl  <- ("qtl" %in% installed.packages(lib.loc=lib)[,1])
-  assign(".has_d",     .has_d,     envir = .QclEnv)
   assign(".has_snow",  .has_snow,  envir = .QclEnv)
-  assign(".has_rcurl", .has_rcurl, envir = .QclEnv)
   assign(".has_rqtl",  .has_rqtl,  envir = .QclEnv)
-  if(!get(".has_d", envir = .QclEnv)) packageStartupMessage(.has_d_warnmsg, appendLF = FALSE)
-  if(!get(".has_snow", envir = .QclEnv)) packageStartupMessage(.has_snow_warnmsg, appendLF = FALSE)
[…]

### 2012-01-30 src/install.libs.R

@@ -1,23 +1,0 @@
-doDMDcompilation <- function(command){
-  if(system(command)==0){
-    cat("** DMD compiler found\n")
-  }else{
-    cat("** WARNING: No DMD compiler found\n")
-  }
-}
-
-Dfiles <- paste(Sys.glob(c("D/*.d","D/dll/*","D/libs/*")),collapse=" ")
-if(WINDOWS){
-  doDMDcompilation(paste("dmd -ofDcode.dll -L/IMPLIB  -O -inline -release ",Dfiles," phobos.lib -IC:/D/dmd2/windows/bin",sep=""))
-}else{
-  doDMDcompilation(paste("dmd -ofDcode.so ",Dfiles," phobos.lib",sep=""))
-}
-files <- Sys.glob(paste("*", SHLIB_EXT, sep=''))
-libarch <- if (nzchar(R_ARCH)) paste('libs', R_ARCH, sep='') else 'libs'
-dest <- file.path(R_PACKAGE_DIR, libarch)
-dir.create(dest, recursive = TRUE, showWarnings = FALSE)
-file.copy(files, dest, overwrite = TRUE)
-file.remove(Sys.glob(paste("*", SHLIB_EXT, sep='')))
-file.remove(Sys.glob("*.o"))
-file.remove(Sys.glob("*.lib"))
-file.remove(Sys.glob("*.obj"))

### 2012-01-30 src/qcl.cpp

@@ -1,7 +1,0 @@
-/**********************************************************************
- *
- * QCL.cpp
- *
- **********************************************************************/
-
-#include "qcl.h"

### 2012-01-30 src/qcl.h

@@ -1,12 +1,0 @@
-/**********************************************************************
- *
- * QCL.h
- *
- **********************************************************************/
-
-#ifndef __QCL_H
-  #define __QCL_H
-
-#include <R.h>
-
-#endif // QCL_H

### 2012-01-30 inst/methodPaper/document.tex
No idealized cut-off, use the permutations

@@
     \subfigure[CTL profiles]{
-      \includegraphics[keepaspectratio,scale=.26]{img/ctl_qtl.pdf}
+      \includegraphics[keepaspectratio,scale=.3]{img/ctl_qtl.pdf}
       \label{subfiga}
@@
     \subfigure[Traits x Traits interaction matrix]{
-      \includegraphics[keepaspectratio,scale=.65]{img/ctl_pxp.pdf}
+      \includegraphics[keepaspectratio,scale=.3]{img/ctl_pxp.pdf}
       \label{subfigb}
@@
     \subfigure[Reconstructed network]{
-      \includegraphics[keepaspectratio,scale=.17]{img/ctl_net.pdf}
+      \includegraphics[keepaspectratio,scale=.2]{img/ctl_net.pdf}
       \label{subfigc}

### 2012-01-30 R/qcl.image.R
No idealized cut-off, use the permutations

@@ -16,7 +16,7 @@ image.QCLobject <- function(x, against = c("markers","phenotypes"), onlySignific
     if(verbose) cat("Processing:",p,"from QCL to LOD\n")
     lod <- QCLtoLODvector(x[[p]], against)
     if(onlySignificant){
-      if(max(lod) > -log10(significance)){
+      if(max(lod) > getPermuteThresholds(x[[p]])[1]){
         mymatrix <- rbind(mymatrix,lod)
         mynames <- c(mynames,attr(x[[p]]$qcl,"name"))  
       }
@@ -33,8 +33,8 @@ image.QCLobject <- function(x, against = c("markers","phenotypes"), onlySignific
 internal.image <- function(mymatrix, colorrange, mainlabel, do.grid, grid.col){
   if(!is.null(mymatrix)){ 
     image(1:ncol(mymatrix),1:nrow(mymatrix),t(mymatrix), main=mainlabel, yaxt="n", 
-          xaxt="n", ylab="", xlab="",col=colorrange, cex.main=0.7, 
-          breaks = c(0,1,2,3,4,5,6,7,8,9,10,100))
+          xaxt="n", ylab="", xlab="",col=c("white","gray","darkgray","black"), cex.main=0.7, 
+          breaks = c(0,5,10,50,100))
     axis(2,rownames(mymatrix),at=1:nrow(mymatrix),las=2,cex.axis=0.5)
     axis(1,colnames(mymatrix),at=1:ncol(mymatrix),las=2,cex.axis=0.5)
     if(do.grid){

### 2012-01-30 inst/methodPaper/document.tex
Updates to the text

@@
   \section{Motivation:}
-  Here we present the concept of Correlated Traits Locus (CTL) mapping, a novel approach 
-  to detect genetic regulation of traits in recombinant inbred line populations (RIL) 
-  or SNP genotypes populations. CTL mapping allows association of trait correlation 
-  differences to genetic markers, in almost the same way as QTL mapping associates genetic 
-  markers with differences in expression.
+  Here we present Correlated Traits Locus (CTL) mapping, a novel approach 
+  to detect genetic regulation of traits in natural and inbred populations. 
+  CTL mapping allows association of trait correlation differences to genetic 
+  markers, in almost the same way as QTL mapping associates genetic  markers with differences in expression.
 
@@
 
-  We present Correlated Traits Locus (CTL) mapping as a comparable method to QCL mapping, 
-  CTL unlike QTL associates differential correlation to genetic variation, i.e. to identify regions in the 
-  genome for which one genotype value leads to correlated expression between a trait and all 
-  other traits, exploiting the fact that any QTL will be the result of either direct induced 
-  variation or indirect variation, which will be visible as correlation differences among 
-  individuals carrying different genotypes.
+  We present Correlated Traits Locus (CTL) mapping as a comparable method to QCL mapping.
+  CTL unlike QTL links differences in correlation to genetic variation, i.e. to identify regions in the 
+  genome for which one genotype value leads to correlated expression between a trait and  
+  other traits, while the other genotype doesn't show this correlation. CTL exploits the fact that 
+  any QTL will be the result of either direct induced variation or indirect variation. The latter is 
+  visible as correlation differences among individuals carrying different genotypes.
 
@@
 
-  Split the population at a genetic marker in two groups (aa and bb). Calculate the correlation 
-  between our trait of interest and other traits. Calculate the CTL matrix by taking the 
-  squared difference between matrix A and B.
-
-  Single trait CTL profiles can be obtained by transforming the CTL scores into LOD scores. 
-  For this we use permutations obtained during 15.000 permutation. We note that for each trait this 
-  shows a different distribution of permutation scores. We assess the likelyhood (per trait) and 
-  take the $-log10(P)$ of every CTL score to compare QTL profiles with summarized CTL profiles. 
-  A CTL profile show the contribution of each trait to the total likelyhood.
+  Split the population at a genetic marker in two groups (A and B). Calculate the correlation 
+  between our trait of interest and other traits. Then calculate the CTL score by taking the 
+  squared difference between A and B.
 
@@
   \caption[Plots]{
-  {\emph {\bf a}}: Example CTL + QTL profiles (red). Seen are the QTL profile on top of the CTL profile. 
-  The first peak is a CTL caused by X4.Methylsulfinylbutyl and X5.Methylthiopenty showing a significant 
-  difference in correlation, we don't observe a QTL for this location. 
+  {\emph {\bf a}}: Example CTL + QTL profiles. Seen are the QTL profile on top of the CTL profile. 
+  The first peak is a CTL, we don't observe a QTL for this location. The second peak is a QTL peek, with no underlying CTL. 
   {\emph {\bf b}}: Traits x Traits interaction matrix (TxT matrix). Information obtained using CTL mapping, 
@@
 
-\emph{ {\bf Network reconstruction}}
-  To build up gene interaction networks, we take the Traits x Marker and Traits x Phentype 
-  relations matrices.  From the TxM we can identify partners of our trait, and from the PxP we 
-  identify the location of out trait in the pathway or main regulatory traits.
-
 \emph{ {\bf Significance testing}}
@@
   trait is broken, by redistributing at random genotypes amongst individuals not 
-  allowing for duplicates. The CTL on these new individuals allows for estimation of 
-  the underlying 0-distribution of CTL scores. When observe CTL scores in 
-  real data which are higher than any CTL score obtained during permutation we 
-  use a generalized Pareto distribution (GPD) to estimate the extreme tail 
-  of the distribution \cite{Knijnenburg:2009}, to get likelyhood estimates for the 
+  allowing for duplicates. When we observe CTL scores in real data which are higher than any 
+  CTL score obtained during permutation a generalized Pareto distribution (GPD) is used to 
+  estimate the extreme tail of the distribution \cite{Knijnenburg:2009}, to get likelihood estimates for the 
   extreme scores observed.
@@
 \section{Results}
-  CTL mapping was performed and a per trait CTL thresholds of $P < 0.001$ was determined by using 15.000 permutations.
+  CTL mapping was performed and a per trait CTL thresholds of $P < 0.001$ was 
+  determined by using 15,000 permutations.
   Information obtained by CTL mapping is shown in the trait to trait interaction matrix (Fig. \ref{subfigb}).
   The CTL profiles were combined with QTL information, leading to the following observations \ref{subfiga}:\\
-  1) Additional CTL found, where no QTL was present. Loss of variation in the genes 
+  1) Additional CTL found, where no QTL was present (Fig \ref{subfiga} - Peak 1). Loss of variation in the genes 
   underneath the CTL are not causing variation differences in the observed trait, CTL is able to 
@@
   2) CTL co-localized with QTL, in this situation CTL gives a detailed view of which traits 
-  are contributing variation to the observed QTL.\\
-  3) QTL with no CTL, we interpret this as the source of the variation. Analysis in this way of 
+  are contributing variation to the observed QTL  (Fig \ref{subfiga} - Peak 3).\\
+  3) QTL with no CTL  (Fig \ref{subfiga} - Peak 2), we interpret this as the source of the variation. Analysis in this way of 
   all 24 Single trait CTL profiles combined with their QTL profiles allows us to reconstruct the 

### 2012-01-30 inst/methodPaper/document.tex
Updates to the text

@@
   determined by using 15,000 permutations.
-  Information obtained by CTL mapping is shown in the trait to trait interaction matrix (Fig. \ref{subfigb}).
-  The CTL profiles were combined with QTL information, leading to the following observations \ref{subfiga}:\\
+  When combining CTL and QTL information, we observe:\\
   1) Additional CTL found, where no QTL was present (Fig \ref{subfiga} - Peak 1). Loss of variation in the genes 
@@
 
+  Information obtained by CTL mapping is shown in the trait to trait interaction matrix (Fig. \ref{subfigb}).
+
 \emph{ {\bf Network reconstruction}}
-  Using the previously decribed method to draw the arrows between traits, we reconstruct the interaction 
+  Using the information from CTL and QTL we are able to draw arrows between traits, reconstruct the interaction 
   network amongst the 24 metabolities in Arabidopsis (Fig. \ref{subfigc}). We observe that our network is 

### 2012-01-31 R/qcl.image.R
Code changes (add a plot on Ritserts advice)

@@ -33,8 +33,8 @@ image.QCLobject <- function(x, against = c("markers","phenotypes"), onlySignific
 internal.image <- function(mymatrix, colorrange, mainlabel, do.grid, grid.col){
   if(!is.null(mymatrix)){ 
     image(1:ncol(mymatrix),1:nrow(mymatrix),t(mymatrix), main=mainlabel, yaxt="n", 
-          xaxt="n", ylab="", xlab="",col=c("white","gray","darkgray","black"), cex.main=0.7, 
-          breaks = c(0,5,10,50,100))
+          xaxt="n", ylab="", xlab="",col=c("white",gray.colors(4)[4:1]), cex.main=0.7, 
+          breaks = c(0,2,4,8,10,100))
     axis(2,rownames(mymatrix),at=1:nrow(mymatrix),las=2,cex.axis=0.5)
     axis(1,colnames(mymatrix),at=1:ncol(mymatrix),las=2,cex.axis=0.5)
     if(do.grid){

### 2012-01-31 R/qcl.plot.R

@@ -38,6 +38,48 @@ plot.QCLobject <- function(x, pheno.col=1:length(x), ...){
   return(image.QCLobject(x,...))
 }
 
+plot.QCLscan2 <- function(x, addQTL = TRUE, onlySignificant = TRUE, qcl.threshold =0.6, do.legend=TRUE, ...){
+  if(missing(x)) stop("argument 'x' is missing, with no default")
+  if(!is.null(x$p)){
+    mysign <- as.numeric(which(apply(abs(x$qcl),1,max) > getPermuteThresholds(x)[1]))
+  }else{
+    mysign <- as.numeric(which(apply(abs(x$qcl),1,max) > qcl.threshold))
+  }
+  if(length(mysign) ==0){
+    mysign <- 1:nrow(x$qcl)
+    do.legend=FALSE
+  }
+  if(is.null(x$l)){
+    QCLmatrix <- matrix(abs(x$qcl[mysign, ]),length(mysign),ncol(x$qcl))
+  }else{
+    QCLmatrix <- matrix(x$l[mysign, ],length(mysign),ncol(x$l))
+  }
+  summarized <- apply(QCLmatrix,2,sum)
+  plot(c(0,ncol(x$qcl)),c(0,max(c(summarized,x$qtl))), type='n',xlab="Marker", ylab="-log10(P-value)", main=paste("Phenotype contribution to QCL of",attr(x$qcl,"name")),...)
+  p <- rep(0,ncol(x$qcl))
+  i <- 1;
+  mycolors <- rep("black",nrow(QCLmatrix))
+  apply(QCLmatrix,1,
+    function(d){
+      points(d,type='b',col=mycolors[i],pch=i,lwd=2)
+      p <<- p + d
+      i <<- i + 1
+    }
+  )
+  #if(!is.null(x$l)){
+  #  n <- dim(x$qcl)[2]
+  #  abline(h=-log10(c(0.05/n,0.01/n,0.001/n)),col=c("red","orange","green"),lty=2)
+  #  legend("topright",as.character(paste("QCL-FDR:",c(0.05,0.01,0.001),"%")),col=c("red","orange","green"),lty=rep(2,3),lwd=1,cex=1)
+  #}
+  if(do.legend){
+    legend("topleft",c("QTL",paste("CTL",rownames(x$qcl)[mysign])),col=c("red",mycolors),lwd=2,pch=c(NA,1:length(mycolors)),cex=1.2)
+  }
[…]

### 2012-01-31 R/qcl.statistics.R

@@ -115,10 +115,10 @@ toLod <- function(QCLscan, onlySignificant = TRUE, verbose = FALSE){
 QCLtoLODvector <- function(QCLscan, against = c("markers","phenotypes")){
   if(!is.null(QCLscan$l)){
     if(against[1]=="markers")return(apply(QCLscan$l,2,sum))
-    if(against[1]=="phenotypes")return(apply(QCLscan$l,1,sum))
+    if(against[1]=="phenotypes")return(apply(QCLscan$l,1,max))
   }else{
     if(against[1]=="markers")return(apply(abs(QCLscan$qcl),2,sum))
-    if(against[1]=="phenotypes")return(apply(abs(QCLscan$qcl),1,sum))  
+    if(against[1]=="phenotypes")return(apply(abs(QCLscan$qcl),1,max))  
   }
 }
 

### 2012-01-31 R/qcl.profiles.R
Code changes (add a plot on Ritserts advice)

@@ -17,7 +17,7 @@ QCLprofiles <- function(QCLobject, significance = 0.05, against = c("markers","p
     if(verbose) cat("Processing:",p,"from QCL to LOD\n")
     lod <- QCLtoLODvector(QCLobject[[p]], against)
     cat(length(lod),"\n")
-    if(max(lod) > -log10(significance)){
+    if(max(lod) > getPermuteThresholds(QCLobject[[p]])[1]){
       mymatrix <- rbind(mymatrix,lod)
       mynames <- c(mynames,attr(QCLobject[[p]]$qcl,"name"))  
     }

### 2012-01-31 inst/methodPaper/document.tex
Paper updates
Code for paper figures

@@
 \address{
-  $^{1}$Groningen Bioinformatics Centre, University of Groningen, Groningen, The Netherlands.
+  $^{1}$Groningen Bio-informatics Centre, University of Groningen, Groningen, The Netherlands.
 }
 \history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}
-\editor{Associate Editor: XXXXXXX}
+\editor{Associate Editor: XXXXX}
 \maketitle
@@
   \section{Availability:}
-  Software, documentation and source code are available from:
-  \href{http://www.mapctl.org}{www.mapctl.org}
+  Software, documentation and source code are available at request.
   \section{Contact:} 
@@
   Genetical genomics experiments have shown that gene expression variation can be mapped to 
-  genotypic variation \cite{Jansen:2001}. Quantitative Trait Locus (QTL) mapping of a 
+  genetic variation \cite{Jansen:2001}. Quantitative Trait Locus (QTL) mapping of a 
   gene identifies regions in the genome for which different genotypes lead to differential 
@@
 
-  We present Correlated Traits Locus (CTL) mapping as a comparable method to QCL mapping.
+  We present Correlated Traits Locus (CTL) mapping as a comparable method to QTL mapping.
   CTL unlike QTL links differences in correlation to genetic variation, i.e. to identify regions in the 
@@
     \subfigure[Reconstructed network]{
-      \includegraphics[keepaspectratio,scale=.2]{img/ctl_net.pdf}
+      \includegraphics[keepaspectratio,scale=.3]{img/ctl_net.pdf}
       \label{subfigc}
@@
   {\emph {\bf c}}:  Reconstructed network, this shows the same information as the TxT matrix only in a more accessible way, we visualized our 
-  network using cytoscape, the linewidth is the summarized interaction evidence.
+  network using cytoscape, the width of the line is the maximum interaction evidence.
  }
@@
   When combining CTL and QTL information, we observe:\\
-  1) Additional CTL found, where no QTL was present (Fig \ref{subfiga} - Peak 1). Loss of variation in the genes 
+  1) Additional CTL found, where no QTL was present (Fig \ref{subfiga} - P1). Loss of variation in the genes 
   underneath the CTL are not causing variation differences in the observed trait, CTL is able to 
   detect the differences in correlation.\\
-  2) CTL co-localized with QTL, in this situation CTL gives a detailed view of which traits 
-  are contributing variation to the observed QTL  (Fig \ref{subfiga} - Peak 3).\\
-  3) QTL with no CTL  (Fig \ref{subfiga} - Peak 2), we interpret this as the source of the variation. Analysis in this way of 
+  2) QTL with no CTL (Fig \ref{subfiga} - P2), we interpret this as the source of the variation. Analysis in this way of 
   all 24 Single trait CTL profiles combined with their QTL profiles allows us to reconstruct the 
   source of natural variation and follow it down the network.
+  3) CTL co-localized with QTL, in this situation CTL gives a detailed view of which traits 
+  are contributing variation to the observed QTL  (Fig \ref{subfiga} - P3).\\
 

### 2012-01-31 inst/methodPaper/plots.R

@@ -1,0 +1,24 @@
+#plots for the paper
+library(qcl)
+qcl_result <- QCLscan(ath.metab$genotypes, ath.metab$phenotypes, n.perm=200)
+setwd("E:\\github\\Rpackages\\QCLmapping\\inst\\methodPaper\\img")
+
+
+pdf("ctl_qtl.pdf",width=10,height=7)
+  plot.QCLscan2(qcl_result[[3]])
+  text(54,10,"P1")
+  arrows(54,9,y1=7,col="red",lwd=2)
+  text(73,10,"P2")
+  arrows(73,9,y1=7,col="red",lwd=2)
+  text(93,10,"P3")
+  arrows(93,9,97,y1=7,col="red",lwd=2)
+dev.off()
+
+pdf("ctl_pxp.pdf",width=10,height=7)
+  op <- par(mar=c(7, 8, 4, 2) + 0.1)
+  image(qcl_result,against="phenotypes")
+dev.off()
+
+QCLnetwork(qcl_result)
+
+#Open cytoscape and create the last plot from network.sif

### 2012-02-06 inst/doc/article.Rnw
Fixed R CMD build and check

@@
 
-\begin{figure}[ht]
-  \centerline{\includegraphics[height=7cm]{qcl.pdf}}
-  \vspace{-15pt}
-  \caption{{\scriptsize QCL mapping: We start by looking at a single genetic marker and dividing our populaiton into two groups.
-  We start by choosing a phenotype (red), and use its expression to detect the red module (high correlation in gray genotype). 
-  We see that something underlying our current marker in the gray genotype regulates the expression of red (but not blue)}}
-  \vspace{-20pt}
-  \label{fig:qcl}
-\end{figure}
+%\begin{figure}[ht]
+%  \centerline{\includegraphics[height=7cm]{qcl.pdf}}
+%  \vspace{-15pt}
+%  \caption{{\scriptsize QCL mapping: We start by looking at a single genetic marker and dividing our populaiton into two groups.
+%  We start by choosing a phenotype (red), and use its expression to detect the red module (high correlation in gray genotype). 
+%  We see that something underlying our current marker in the gray genotype regulates the expression of red (but not blue)}}
+%  \vspace{-20pt}
+%  \label{fig:qcl}
+%\end{figure}
 
-\begin{figure}[ht]
-  \centerline{\includegraphics[height=6cm]{qcl_n.pdf}}
-  \vspace{-15pt}
-  \caption{{\scriptsize Reconstructed pathway location from gene expression data: Insight into the underlying (hidden) regulatory network 
-  can be obtained when we look at the amount of expression showing the same switch. Depending on the amount of blue expressions 
-  co-affected we can guestimate the location of the red expression inside the pathway}}
-  \vspace{-20pt}
-  \label{fig:qcl_n}
-\end{figure}
+%\begin{figure}[ht]
+%  \centerline{\includegraphics[height=6cm]{qcl_n.pdf}}
+%  \vspace{-15pt}
+%  \caption{{\scriptsize Reconstructed pathway location from gene expression data: Insight into the underlying (hidden) regulatory network 
+%  can be obtained when we look at the amount of expression showing the same switch. Depending on the amount of blue expressions 
+%  co-affected we can guestimate the location of the red expression inside the pathway}}
+%  \vspace{-20pt}
+%  \label{fig:qcl_n}
+%\end{figure}
 

### 2012-02-07 R/ctl.hist.R
R and tests to CTL

@@ -1,0 +1,24 @@
+#
+# ctl.hist.R
+#
+# copyright (c) 2010-2011 Danny Arends and Ritsert C. Jansen
+# last modified Jan, 2012
+# first written Dec, 2011
+# 
+# Histogram routines for CTL analysis
+#
+
+hist.CTLobject <- function(x, pheno.col=1, ...){
+  if(missing(x)) stop("argument 'x' which expects a 'CTLobject' object is missing, with no default")
+  namez <- NULL
+  plot(c(0,1.0),c(0,length(unlist(x[[pheno.col[1]]]$p))/5),type='n',main="CTL scores permutations",ylab="Frequency",xlab="Difference in correlation^2")
+  for(pheno in pheno.col){
+    if(is.null(x[[pheno]]$p)) stop(paste("Permutations not found for pheno.col=",pheno))
+    sorted <- sort(unlist(x[[pheno]]$p))
+    hist(sorted, breaks=seq(0,1.0,0.01), add=TRUE, col=pheno, ...)
+    namez <- c(namez,attr(x[[pheno]]$ctl,"name"))
+  }
+  legend("topright", namez, col=pheno.col,lwd=6)
+}
+
+# end of ctl.hist.R

### 2012-02-07 R/ctl.image.R

@@ -1,32 +1,32 @@
 #
-# qcl.image.R
+# ctl.image.R
 #
 # copyright (c) 2010 Danny Arends and Ritsert C. Jansen
 # last modified Jan, 2012
 # first written Nov, 2010
 # 
-# Image plot routines for QCL analysis
+# Image plot routines for CTL analysis
 #
 
-image.QCLobject <- function(x, against = c("markers","phenotypes"), onlySignificant = FALSE, significance = 0.05, do.grid=TRUE, grid.col = "black", verbose = FALSE, ...){
+image.CTLobject <- function(x, against = c("markers","phenotypes"), onlySignificant = FALSE, significance = 0.05, do.grid=TRUE, grid.col = "black", verbose = FALSE, ...){
   colorrange <- c("white",gray.colors(10)[10:1])
   mymatrix <- NULL
   mynames <- NULL
   for(p in 1:length(x)){
-    if(verbose) cat("Processing:",p,"from QCL to LOD\n")
-    lod <- QCLtoLODvector(x[[p]], against)
+    if(verbose) cat("Processing:",p,"from CTL to LOD\n")
+    lod <- CTLtoLODvector(x[[p]], against)
     if(onlySignificant){
       if(max(lod) > getPermuteThresholds(x[[p]])[1]){
         mymatrix <- rbind(mymatrix,lod)
-        mynames <- c(mynames,attr(x[[p]]$qcl,"name"))  
+        mynames <- c(mynames,attr(x[[p]]$ctl,"name"))  
       }
     }else{
       mymatrix <- rbind(mymatrix,lod)
-      mynames <- c(mynames,attr(x[[p]]$qcl,"name"))
+      mynames <- c(mynames,attr(x[[p]]$ctl,"name"))
     }
   }
   rownames(mymatrix) <- mynames
-  mainlabel <- paste("QCL phenotypes vs",against[1],"at P-value <",significance)
+  mainlabel <- paste("CTL phenotypes vs",against[1],"at P-value <",significance)
   internal.image(mymatrix, colorrange, mainlabel,do.grid, grid.col)
 }
 
@@ -51,16 +51,16 @@ QTLimage <- function(x, onlySignificant = FALSE, significance = 0.05, do.grid=TR
   mymatrix <- NULL
   mynames <- NULL
   for(p in 1:length(x)){
-    if(verbose) cat("Processing QTL",p,"from QCLobject\n")
[…]

### 2012-02-07 R/ctl.permutation.R

@@ -1,58 +1,58 @@
 #
-# qcl.permutations.R
+# ctl.permutations.R
 #
 # copyright (c) 2010 Danny Arends, Bruno Tesson and Ritsert C. Jansen
 # last modified Jan, 2012
 # first written Nov, 2010
 # 
-# R functions to do permutation on QCL mapping
+# R functions to do permutation on CTL mapping
 # Example data C. Elegans and available at request ( Danny.Arends@gmail.com )
 #
 
-#-- QCLpermuteMC function --#
-QCLpermuteMC <- function(genotypes, phenotypes, pheno.col, method = c("pearson", "kendall", "spearman"), n.perm=100, n.cores=2, genotype.values=c(1,2), directory="permutations", saveFiles = FALSE, verbose=FALSE, ...){
+#-- CTLpermuteMC function --#
+CTLpermuteMC <- function(genotypes, phenotypes, pheno.col, method = c("pearson", "kendall", "spearman"), n.perm=100, n.cores=2, genotype.values=c(1,2), directory="permutations", saveFiles = FALSE, verbose=FALSE, ...){
   if(!has_snow()) stop("SNOW is not installed (or not yet loaded)")
   if(!file.exists(directory)) dir.create(directory)
   if(missing(pheno.col)) pheno.col <- 1:ncol(phenotypes)
-  QCLperm <- vector("list",length(pheno.col))
+  CTLperm <- vector("list",length(pheno.col))
   idx <- 1
   cl <- snow::makeCluster(rep("localhost",n.cores))
   for(p in pheno.col){
     ss <- proc.time()
     cat("  - Starting multi core permutation for phenotype",p,"\n")
     rvm <- getRVM(n.perm,nrow(genotypes))
-    QCLperm[[idx]] <- unlist(snow::parLapply(cl,as.list(1:n.perm),get("QCLpermute.internal"), genotypes, phenotypes, p, method, rvm, genotype.values, directory, saveFiles, verbose,...))
+    CTLperm[[idx]] <- unlist(snow::parLapply(cl,as.list(1:n.perm),get("CTLpermute.internal"), genotypes, phenotypes, p, method, rvm, genotype.values, directory, saveFiles, verbose,...))
     idx <- idx+1
     el <- proc.time()
     cat("  -",n.perm,"permutations took:",as.numeric(el[3]-ss[3]),"seconds.\n")
   }
-  class(QCLperm) <- c(class(QCLperm),"QCLpermute")
[…]

### 2012-02-07 R/ctl.plot.R

@@ -1,0 +1,135 @@
+#
+# ctl.plot.R
+#
+# copyright (c) 2010 Danny Arends and Bruno Tesson
+# last modified Jan, 2012
+# first written Nov, 2010
+# 
+# Plotting routines for CTL analysis
+#
+
+CTLasLOD <- function(CTLscan, QTLscores, main, do.legend=TRUE){
+  if(missing(CTLscan)) stop("argument 'CTLscan' is missing, with no default")
+  if(missing(main)){
+    main <- paste("Comparison CTL:QTL of",attr(CTLscan$ctl,"name"))
+  }
+  op <- par(mfrow=c(2,1))
+  plot.CTLscan(CTLscan,do.legend=do.legend)
+  CTLscores <- CTLtoLODvector(CTLscan)
+  if(!missing(QTLscores)){
+    QTLscores <- as.numeric(unlist(QTLscores))
+    plot(c(0,length(CTLscores)),c(0,max(c(CTLscores,QTLscores))),type='n', main=main, ylab="LOD",xlab="Marker")
+    points(CTLscores,type='l',col="black",lwd=3)
+    points(QTLscores,type='l',col="red",lwd=2,lty=1)
+    if(do.legend) legend("topleft",c("CTL","QTL"),col=c("black","red"),lty=c(1,1),lwd=c(3,2))
+  }else{
+    plot(c(0,length(CTLscores)),c(0,max(c(CTLscores))),type='n', main=main, ylab="LOD",xlab="Marker")
+    points(CTLscores,type='l',col="black",lwd=3)
+  }
+}
+
+plot.CTLobject <- function(x, pheno.col=1:length(x), ...){
+  if(length(x) == 1){
+    return(plot.CTLscan(x[[1]],...))
+  }
+  if(length(pheno.col) == 1){
+    return(plot.CTLscan(x[[pheno.col]],...))
+  }
+  return(image.CTLobject(x,...))
+}
+
+plot.CTLscan2 <- function(x, addQTL = TRUE, onlySignificant = TRUE, significance = 0.05, do.legend=TRUE, ...){
+  if(missing(x)) stop("argument 'x' is missing, with no default")
+  if(!is.null(x$p) && !is.nan(getPermuteThresholds(x,significance)[1])){
[…]

### 2012-02-07 R/ctl.scan.R

@@ -1,15 +1,15 @@
 #
-# qcl.R
+# ctl.scan.R
 #
 # copyright (c) 2010 Danny Arends and Ritsert C. Jansen
-# last modified Jan, 2012
+# last modified Feb, 2012
 # first written Jan, 2011
 # 
-# R functions to do QCL mapping
+# R functions to do CTL mapping
 #
 
-#-- QCLscan main function --#
-QCLscan <- function(genotypes, phenotypes, pheno.col = 1:ncol(phenotypes), method = c("pearson", "kendall", "spearman"), n.perm=0, n.cores=2, genotype.values=c(1,2), directory="permutations", saveFiles = FALSE, verbose = FALSE){
+#-- CTLscan main function --#
+CTLscan <- function(genotypes, phenotypes, pheno.col = 1:ncol(phenotypes), method = c("pearson", "kendall", "spearman"), n.perm=0, n.cores=2, genotype.values=c(1,2), directory="permutations", saveFiles = FALSE, verbose = FALSE){
   if(missing(genotypes)) stop("argument 'genotypes' is missing, with no default")
   if(missing(phenotypes)) stop("argument 'phenotypes' is missing, with no default")
   
@@ -20,15 +20,15 @@ QCLscan <- function(genotypes, phenotypes, pheno.col = 1:ncol(phenotypes), metho
   results <- vector("list",length(pheno.col))
   idx <- 1
   for(p in pheno.col){
-    cat("Stage 1: Scanning QCL\n")
-    results[[idx]]$qcl <- QCLmapping(genotypes, phenotypes, p, method=method, genotype.values, verbose)
+    cat("Stage 1: Scanning CTL\n")
+    results[[idx]]$ctl <- CTLmapping(genotypes, phenotypes, p, method=method, genotype.values, verbose)
     
     cat("Stage 2: Scanning QTL\n")
     results[[idx]]$qtl <- QTLscan(genotypes, phenotypes, p, verbose)
     
     if(n.perm > 0){
       cat("Stage 3: Permutation\n")
-      results[[idx]]$p <- QCLpermute(genotypes, phenotypes, p, method=method, n.perm, n.cores, genotype.values, directory, saveFiles, verbose)
+      results[[idx]]$p <- CTLpermute(genotypes, phenotypes, p, method=method, n.perm, n.cores, genotype.values, directory, saveFiles, verbose)
       
[…]

### 2012-02-07 R/qcl.hist.R

@@ -1,24 +1,0 @@
-#
-# qcl.hist.R
-#
-# copyright (c) 2010-2011 Danny Arends and Ritsert C. Jansen
-# last modified Jan, 2012
-# first written Dec, 2011
-# 
-# Histogram routines for QCL analysis
-#
-
-hist.QCLobject <- function(x, pheno.col=1, ...){
-  if(missing(x)) stop("argument 'x' which expects a 'QCLobject' object is missing, with no default")
-  namez <- NULL
-  plot(c(0,1.0),c(0,length(unlist(x[[pheno.col[1]]]$p))/5),type='n',main="QCL scores permutations",ylab="Frequency",xlab="Difference in correlation^2")
-  for(pheno in pheno.col){
-    if(is.null(x[[pheno]]$p)) stop(paste("Permutations not found for pheno.col=",pheno))
-    sorted <- sort(unlist(x[[pheno]]$p))
-    hist(sorted, breaks=seq(0,1.0,0.01), add=TRUE, col=pheno, ...)
-    namez <- c(namez,attr(x[[pheno]]$qcl,"name"))
-  }
-  legend("topright", namez, col=pheno.col,lwd=6)
-}
-
-# end of qcl.hist.R

### 2012-02-07 R/qcl.plot.R

@@ -1,135 +1,0 @@
-#
-# qcl.plot.R
-#
-# copyright (c) 2010 Danny Arends and Bruno Tesson
-# last modified Jan, 2012
-# first written Nov, 2010
-# 
-# Plotting routines for QCL analysis
-#
-
-QCLasLOD <- function(QCLscan, QTLscores, main, do.legend=TRUE){
-  if(missing(QCLscan)) stop("argument 'QCLscan' is missing, with no default")
-  if(missing(main)){
-    main <- paste("Comparison QCL:QTL of",attr(QCLscan$qcl,"name"))
-  }
-  op <- par(mfrow=c(2,1))
-  plot.QCLscan(QCLscan,do.legend=do.legend)
-  QCLscores <- QCLtoLODvector(QCLscan)
-  if(!missing(QTLscores)){
-    QTLscores <- as.numeric(unlist(QTLscores))
-    plot(c(0,length(QCLscores)),c(0,max(c(QCLscores,QTLscores))),type='n', main=main, ylab="LOD",xlab="Marker")
-    points(QCLscores,type='l',col="black",lwd=3)
-    points(QTLscores,type='l',col="red",lwd=2,lty=1)
-    if(do.legend) legend("topleft",c("QCL","QTL"),col=c("black","red"),lty=c(1,1),lwd=c(3,2))
-  }else{
-    plot(c(0,length(QCLscores)),c(0,max(c(QCLscores))),type='n', main=main, ylab="LOD",xlab="Marker")
-    points(QCLscores,type='l',col="black",lwd=3)
-  }
-}
-
-plot.QCLobject <- function(x, pheno.col=1:length(x), ...){
-  if(length(x) == 1){
-    return(plot.QCLscan(x[[1]],...))
-  }
-  if(length(pheno.col) == 1){
-    return(plot.QCLscan(x[[pheno.col]],...))
-  }
-  return(image.QCLobject(x,...))
-}
-
-plot.QCLscan2 <- function(x, addQTL = TRUE, onlySignificant = TRUE, significance = 0.05, do.legend=TRUE, ...){
-  if(missing(x)) stop("argument 'x' is missing, with no default")
-  if(!is.null(x$p) && !is.nan(getPermuteThresholds(x,significance)[1])){
[…]

### 2012-02-07 tests/test_ctl_long.R

@@ -1,0 +1,17 @@
+library(ctl)
+data(ath.metabolites)
+ctl_result <- CTLscan(ath.metab$genotypes, ath.metab$phenotypes, pheno.col=1:3, n.perm=20)
+
+jpeg("pxmmatrix.jpg",w=1024,h=768)
+image(ctl_result,grid.col="white")
+dev.off()
+
+jpeg("pxpmatrix.jpg",w=1024,h=768)
+image(ctl_result,grid.col="white",against="phenotypes")
+dev.off()
+
+for(x in 1:3){
+  jpeg(paste("phenotype",x,".jpg",sep=""),w=1024,h=768)
+  plot(ctl_result[[1]])  
+  dev.off()
+}

### 2012-02-07 tests/test_qcl_long.R

@@ -1,17 +1,0 @@
-library(qcl)
-data(ath.metabolites)
-qcl_result <- QCLscan(ath.metab$genotypes, ath.metab$phenotypes, pheno.col=1:3, n.perm=20)
-
-jpeg("pxmmatrix.jpg",w=1024,h=768)
-image(qcl_result,grid.col="white")
-dev.off()
-
-jpeg("pxpmatrix.jpg",w=1024,h=768)
-image(qcl_result,grid.col="white",against="phenotypes")
-dev.off()
-
-for(x in 1:3){
-  jpeg(paste("phenotype",x,".jpg",sep=""),w=1024,h=768)
-  plot(qcl_result[[1]])  
-  dev.off()
-}

### 2012-02-07 R/ctl.network.R
R and tests to CTL

@@ -1,0 +1,87 @@
+#
+# ctl.network.R
+#
+# copyright (c) 2010 Danny Arends and Bruno Tesson
+# last modified Jan, 2012
+# first written Oct, 2011
+# 
+# Network routines for CTL analysis
+#
+
+CTLnetwork <- function(CTLscan, ctl.threshold=0.45, verbose = FALSE){
+  cat("",file="network.sif")
+  cnt <- 0
+  edge_count <- NULL
+  node_names <- NULL
+  for(CTL in CTLscan){
+    for(x in 1:ncol(CTL$ctl)){
+      if(is.null(CTL$p)) stop("No permutations found, need permutations to determine likelihood\n")
+      for(id in which(abs(CTL$ctl[,x]) > getPermuteThresholds(CTL)[1])){
+        edge_name <- paste(attr(CTL$ctl,"name"),"CTL",colnames(CTL$ctl)[x],sep="\t")
+        if(edge_name %in% edge_count[,1]){
+          edgeid <- which(edge_count[,1] %in% edge_name)
+          if(verbose) cat("Duplicate edge",edgeid,":", edge_name,"\n")
+          edge_count[edgeid,2] <- as.numeric(edge_count[edgeid,2])+abs(CTL$l[id,x])
+          edge_count[edgeid,3] <- as.numeric(edge_count[edgeid,3])+1
+        }else{
+          edge_count <- rbind(edge_count,c(edge_name, CTL$l[id,x], 1))
+        }
+        node_names <- c(node_names,attr(CTL$ctl,"name"),rownames(CTL$ctl)[id])
+        cat(attr(CTL$ctl,"name"),"CTL",rownames(CTL$ctl)[id],CTL$ctl[id,x],colnames(CTL$ctl)[x],"\n",file="network.sif",append=TRUE)
+        cnt <- cnt+1
+      }
+    }
+    if(verbose)cat("Analysed phenotype:",attr(CTL,"name"),"\n")
+  }
+  cat("Wrote",cnt,"edges to network.sif\n")
+  if(!is.null(edge_count)){
+    cat("",file="edge_summary.sif")
+    for(x in 1:nrow(edge_count)){
+      cat(edge_count[x,1],"\t",round(as.numeric(edge_count[x,2]),d=1),"\t",edge_count[x,3],"\n",file="edge_summary.sif",append=TRUE,sep="")
+    }
+    cat("Wrote",nrow(edge_count),"unique edges to edge_summary.sif\n")
+  }
[…]

### 2012-02-07 R/ctl.print.R

@@ -1,30 +1,30 @@
 #
-# qcl.print.R
+# ctl.print.R
 #
 # copyright (c) 2010-2011 Danny Arends and Ritsert C. Jansen
 # last modified Jan, 2012
 # first written Oct, 2011
 # 
-# Print routines for QCL analysis
+# Print routines for CTL analysis
 #
 
-print.QCLscan <- function(x, ...){
-  cat("QCLscan summary",attr(x$qcl,"name"),"\n\n")
-  cat("- Number of background phenotypes",dim(x$qcl)[1],"\n")
-  cat("- Number of markers",dim(x$qcl)[2],"\n")
-  cat("- Number of permutations",length(unlist(x$p))/dim(x$qcl)[2],"\n")
+print.CTLscan <- function(x, ...){
+  cat("CTLscan summary",attr(x$ctl,"name"),"\n\n")
+  cat("- Number of background phenotypes",dim(x$ctl)[1],"\n")
+  cat("- Number of markers",dim(x$ctl)[2],"\n")
+  cat("- Number of permutations",length(unlist(x$p))/dim(x$ctl)[2],"\n")
   getPermuteThresholds(x,..., verbose = TRUE)
 }
 
 getPermuteThresholds <-function(x, significance = c(.05,.01,.001), ..., verbose = FALSE){
-  if(any(class(x)=="QCLscan")){
-    n <- dim(x$qcl)[2]
+  if(any(class(x)=="CTLscan")){
+    n <- dim(x$ctl)[2]
     x <- x$p
-    if(!any(class(x)=="QCLpermute")) stop("No permutations found in the QCLscan object")
+    if(!any(class(x)=="CTLpermute")) stop("No permutations found in the CTLscan object")
   }else{
     n <- 1
-    if(!any(class(x)=="QCLpermute")){
-      stop("No QCLscan object")
+    if(!any(class(x)=="CTLpermute")){
+      stop("No CTLscan object")
     }
   }
   sorted <- sort(unlist(x))
@@ -47,8 +47,8 @@ getPermuteThresholds <-function(x, significance = c(.05,.01,.001), ..., verbose 
   values
 }
 
-print.QCLpermute <- function(x, ...){
+print.CTLpermute <- function(x, ...){
   getPermuteThresholds(x, ..., verbose=TRUE)
 }
 
-# end of qcl.print.R
+# end of ctl.print.R

### 2012-02-07 R/ctl.profiles.R

@@ -1,24 +1,24 @@
 #
-# qcl.profiles.R
+# ctl.profiles.R
 #
 # copyright (c) 2011 Danny Arends and Ritsert C. Jansen
-# last modified Jan, 2012
+# last modified Feb, 2012
 # first written Oct, 2010
 # 
 #
 
-#Create the 2 possible QCL matrices: 1) phenotypes versus markers (PxM) and 2) Phenotypes versus phenotypes (PxP)
-QCLprofiles <- function(QCLobject, against = c("markers","phenotypes"), significance=0.05, verbose=FALSE, warn = TRUE){
+#Create the 2 possible CTL matrices: 1) phenotypes versus markers (PxM) and 2) Phenotypes versus phenotypes (PxP)
+CTLprofiles <- function(CTLobject, against = c("markers","phenotypes"), significance=0.05, verbose=FALSE, warn = TRUE){
   mymatrix <- NULL
   mynames <- NULL
   warn <- warn
   notice <- TRUE
-  for(p in 1:length(QCLobject)){
-    lod <- QCLtoLODvector(QCLobject[[p]], against)
+  for(p in 1:length(CTLobject)){
+    lod <- CTLtoLODvector(CTLobject[[p]], against)
     threshold <- -log10(significance)
-    if(!is.nan(getPermuteThresholds(QCLobject[[p]],significance)[1])){
+    if(!is.nan(getPermuteThresholds(CTLobject[[p]],significance)[1])){
       if(notice) cat("  - [Notice] Permutation available parameter 'significance' is ignored\n")
-      threshold <- getPermuteThresholds(QCLobject[[p]],significance)[1]
+      threshold <- getPermuteThresholds(CTLobject[[p]],significance)[1]
       notice <- FALSE
     }else{
       if(warn) cat("  - [Warning] Too few permutations, unable to find significant\n")
@@ -26,7 +26,7 @@ QCLprofiles <- function(QCLobject, against = c("markers","phenotypes"), signific
     }
     if(max(lod) > threshold){
       mymatrix <- rbind(mymatrix,lod)
-      mynames <- c(mynames,attr(QCLobject[[p]]$qcl,"name"))  
+      mynames <- c(mynames,attr(CTLobject[[p]]$ctl,"name"))  
     }
   }
   rownames(mymatrix) <- mynames
@@ -39,13 +39,13 @@ QCLprofiles <- function(QCLobject, against = c("markers","phenotypes"), signific
 }
 
[…]

### 2012-02-07 R/ctl.statistics.R

@@ -1,50 +1,50 @@
 #
-# qcl.statistics.R
+# ctl.statistics.R
 #
 # copyright (c) 2010 Danny Arends, Bruno Tesson and Ritsert C. Jansen
 # last modified Jan, 2012
 # first written Nov, 2011
 # 
-# R functions to do transform QCL mapping scores to Pvalues and LOD
+# R functions to do transform CTL mapping scores to Pvalues and LOD
 # Example data C. Elegans and available at request ( Danny.Arends@gmail.com )
 #
 
-QCLtoP <- function(QCLscan, onlySignificant = TRUE, verbose = TRUE){
-  if(missing(QCLscan)) stop("argument 'QCLscan' is missing, with no default")
-  permvalues <- sort(unlist(QCLscan$p))
+CTLtoP <- function(CTLscan, onlySignificant = TRUE, verbose = TRUE){
+  if(missing(CTLscan)) stop("argument 'CTLscan' is missing, with no default")
+  permvalues <- sort(unlist(CTLscan$p))
   l <- length(permvalues)
   if(onlySignificant){
-    mysignificant <- as.numeric(which(apply(abs(QCLscan$qcl),1,max) > getPermuteThresholds(QCLscan$p)[1]))
+    mysignificant <- as.numeric(which(apply(abs(CTLscan$ctl),1,max) > getPermuteThresholds(CTLscan$p)[1]))
     if(length(mysignificant) > 1){
-      scaled <- abs(QCLscan$qcl[mysignificant, ])
-      rnames <- rownames(QCLscan$qcl)[mysignificant]
+      scaled <- abs(CTLscan$ctl[mysignificant, ])
+      rnames <- rownames(CTLscan$ctl)[mysignificant]
     }else{
-      scaled <- abs(QCLscan$qcl)
-      rnames <- rownames(QCLscan$qcl)
+      scaled <- abs(CTLscan$ctl)
+      rnames <- rownames(CTLscan$ctl)
     }
   }else{
-    scaled <- abs(QCLscan$qcl)
-    rnames <- rownames(QCLscan$qcl)
+    scaled <- abs(CTLscan$ctl)
+    rnames <- rownames(CTLscan$ctl)
   }
-  result <- apply(scaled, 2, function(x){QCLtoPvalue.internal(x, permvalues, l)})
+  result <- apply(scaled, 2, function(x){CTLtoPvalue.internal(x, permvalues, l)})
   rownames(result) <- rnames
   result
 }
 
[…]

### 2012-02-07 R/ctl.utils.R

@@ -1,8 +1,8 @@
 #
-# qcl.utils.R
+# ctl.utils.R
 #
 # copyright (c) 2010 Danny Arends and Ritsert C. Jansen
-# last modified Jan, 2012
+# last modified Feb, 2012
 # first written nov, 2010
 # 
 # check.genotypes, getRVM, lodscorestoscanone, getCorrelatedPhenotypes, gcLoop
@@ -34,7 +34,7 @@ getRVM <- function(n.perms, n.rows){
 
 #Change any list of lodscores into a scanone object (only pre-req: length(lodscores)==sum(nmar(cross))
 lodscorestoscanone <- function(cross,lodscores,traitnames = NULL){
-  if(get(".has_rqtl", envir = .QclEnv)){
+  if(has_rqtl()){
     require(qtl)
     mymap <- qtl::pull.map(cross)
     n <- unlist(lapply(FUN=names,mymap))
@@ -82,4 +82,4 @@ gcLoop <- function(verbose = FALSE){
   }
 }
 
-# end of qcl.utils.R
+# end of ctl.utils.R

### 2012-02-07 R/qcl.network.R

@@ -1,124 +1,0 @@
-#
-# qcl.network.R
-#
-# copyright (c) 2010 Danny Arends and Bruno Tesson
-# last modified Jan, 2012
-# first written Oct, 2011
-# 
-# Network routines for QCL analysis
-#
-
-QCLnetwork <- function(QCLscan, qcl.threshold=0.45, verbose = FALSE){
-  cat("",file="network.sif")
-  cnt <- 0
-  edge_count <- NULL
-  node_names <- NULL
-  for(QCL in QCLscan){
-    for(x in 1:ncol(QCL$qcl)){
-      if(is.null(QCL$p)) stop("No permutations found, need permutations to determine likelihood\n")
-      for(id in which(abs(QCL$qcl[,x]) > getPermuteThresholds(QCL)[1])){
-        edge_name <- paste(attr(QCL$qcl,"name"),"QCL",colnames(QCL$qcl)[x],sep="\t")
-        if(edge_name %in% edge_count[,1]){
-          edgeid <- which(edge_count[,1] %in% edge_name)
-          if(verbose) cat("Duplicate edge",edgeid,":", edge_name,"\n")
-          edge_count[edgeid,2] <- as.numeric(edge_count[edgeid,2])+abs(QCL$l[id,x])
-          edge_count[edgeid,3] <- as.numeric(edge_count[edgeid,3])+1
-        }else{
-          edge_count <- rbind(edge_count,c(edge_name, QCL$l[id,x], 1))
-        }
-        node_names <- c(node_names,attr(QCL$qcl,"name"),rownames(QCL$qcl)[id])
-        cat(attr(QCL$qcl,"name"),"QCL",rownames(QCL$qcl)[id],QCL$qcl[id,x],colnames(QCL$qcl)[x],"\n",file="network.sif",append=TRUE)
-        cnt <- cnt+1
-      }
-    }
-    if(verbose)cat("Analysed phenotype:",attr(QCL,"name"),"\n")
-  }
-  cat("Wrote",cnt,"edges to network.sif\n")
-  if(!is.null(edge_count)){
-    cat("",file="edge_summary.sif")
-    for(x in 1:nrow(edge_count)){
-      cat(edge_count[x,1],"\t",round(as.numeric(edge_count[x,2]),d=1),"\t",edge_count[x,3],"\n",file="edge_summary.sif",append=TRUE,sep="")
-    }
-    cat("Wrote",nrow(edge_count),"unique edges to edge_summary.sif\n")
[…]

### 2012-02-07 R/zzz.R

@@ -1,36 +1,36 @@
 #
-# qcl.significance.R
+# zzz.R
 #
 # copyright (c) 2010 Danny Arends and Bruno Tesson
 # last modified feb, 2011
 # first written nov, 2010
 # 
-# .First.lib is run when the package is loaded with library(qtl)
+# .First.lib is run when the package is loaded with library(ctl)
 #
 
 #Global package variables
-.QclEnv <- new.env()
+.CtlEnv <- new.env()
 
-get_QclEnv <- function(){.QclEnv}
-has_snow <- function(){ get(".has_snow", envir = .QclEnv) }
-has_rqtl <- function(){ get(".has_rqtl", envir = .QclEnv) }
+get_CtlEnv <- function(){.CtlEnv}
+has_snow <- function(){ get(".has_snow", envir = .CtlEnv) }
+has_rqtl <- function(){ get(".has_rqtl", envir = .CtlEnv) }
 
 .has_snow_warnmsg    <- "- Package: SNOW not found, multi CPU support is not available/disabled\n"
 .has_rqtl_warnmsg    <- "- Package: qtl not found, R/qtl functionality not available/disabled\n"
 
 #Package loading
 .onAttach <- function(lib, pkg){
-  packageStartupMessage("- Loading package qcl\n", appendLF = FALSE)
+  packageStartupMessage("- Loading package ctl\n", appendLF = FALSE)
   .has_snow <- FALSE
   .has_rqtl <- FALSE
   .has_snow  <- ("snow" %in% installed.packages(lib.loc=lib)[,1])
   .has_rqtl  <- ("qtl" %in% installed.packages(lib.loc=lib)[,1])
-  assign(".has_snow",  .has_snow,  envir = .QclEnv)
-  assign(".has_rqtl",  .has_rqtl,  envir = .QclEnv)
-  if(!get(".has_snow", envir = .QclEnv)){
+  assign(".has_snow",  .has_snow,  envir = .CtlEnv)
+  assign(".has_rqtl",  .has_rqtl,  envir = .CtlEnv)
+  if(!get(".has_snow", envir = .CtlEnv)){
     packageStartupMessage(.has_snow_warnmsg, appendLF = FALSE)
   }
-  if(!get(".has_rqtl", envir = .QclEnv)){
+  if(!get(".has_rqtl", envir = .CtlEnv)){
     packageStartupMessage(.has_rqtl_warnmsg, appendLF = FALSE)
   }
 }

### 2012-02-07 tests/test_ctl.R

@@ -1,4 +1,4 @@
-library(qcl)
+library(ctl)
 
 if(has_rqtl()){
   require(qtl)
@@ -6,11 +6,11 @@ if(has_rqtl()){
   cross <- read.cross("csvr",file="brem_cross.csvr",geno=c("AA","AB"))
   cross <- convert2riself(cross)
 
-  QCL <- QCLscan.cross(cross,1, verbose=T)
-  image(QCL)
-  plot(QCL)
+  CTL <- CTLscan.cross(cross,1, verbose=T)
+  image(CTL)
+  plot(CTL)
 
-  nodes <- QCLnetwork(QCL,0.55)
+  nodes <- CTLnetwork(CTL,0.55)
 
   write.nodeAttributeFile(nodes,probeannot)
   genes <- nodesToGenes(nodes,probeannot)

### 2012-02-07 inst/doc/article.Rnw
CTL in inst/docs

@@
 % Danny Arends
-% The qcl mapping routine
+% The ctl mapping routine
 %\VignetteIndexEntry{Article}
@@
 
-\title{QCL: Added value of differential correlation}
+\title{CTL: Added value of differential correlation}
 
@@
 <<echo=FALSE,results=hide>>=
-  require(qcl)
+  require(ctl)
   set.seed(200)
@@
 \begin{sciabstract}
-  In this paper we present Quanitative Correlation Locus (QCL) mapping, a novel approach 
+  In this paper we present Quanitative Correlation Locus (CTL) mapping, a novel approach 
   to detect genetic regulation of phenotypes in recombinant inbred line populations (RIL) 
-  or SNP genotypes populations. QCL mapping allows association of phenotype correlation 
+  or SNP genotypes populations. CTL mapping allows association of phenotype correlation 
   differences to genetic markers, in almost the same way as QTL mapping associates genetic 
   markers with differences in expression.
-  We use QCL mapping to reconstruct underlying genetic architecture, and provide insight 
-  into regulation of (gene) expression. QCL information in combination with classical QTL 
+  We use CTL mapping to reconstruct underlying genetic architecture, and provide insight 
+  into regulation of (gene) expression. CTL information in combination with classical QTL 
   information leads to a more detailed insight into the genetic architecture underlying 
@@
   inference. The importance of identifying differences in correlation has been increasingly 
-  recognized and this led to the development of this algorithm to perform QCL analysis in 
+  recognized and this led to the development of this algorithm to perform CTL analysis in 
   experimental crosses.
@@
   depends (but is not dependant) on A, a phenomena generally refered to as phenotype buffering. 
-  This is what QCL mapping exploits, the hypothesis is that although B doesn't show a QTL, 
+  This is what CTL mapping exploits, the hypothesis is that although B doesn't show a QTL, 
   it will become uncorrelated when we partition phenotype B into the groups based on the genotype at 
@@
 
-  QCL thus maps not differential co-expression but differential correlation to genetic 
+  CTL thus maps not differential co-expression but differential correlation to genetic 
   variation, i.e. to identify regions in the genome for which one genotype value leads to 
@@
   which will be visible as correlation differences among individuals carrying different genotypes. 
-  We merge information coming from QCL and QTL to reveal underlying genetic architecture, 
+  We merge information coming from CTL and QTL to reveal underlying genetic architecture, 
   which would remain hidden when using only classical QTL information.
 
-  We urge all QTL researchers to combine the new QCL method with existing QTL analysis to expose 
+  We urge all QTL researchers to combine the new CTL method with existing QTL analysis to expose 
   transcriptional networks involved in the regulation of complex phenotypes. For this we 
-  provide an R package with a reference implementation of the QCL algorithm.
+  provide an R package with a reference implementation of the CTL algorithm.
 \bigskip
@@
 
-\emph{\sffamily QCL mapping}\\
-  As mentioned in the introduction QCL mapping uses buffering to detect interaction partners. 
+\emph{\sffamily CTL mapping}\\
+  As mentioned in the introduction CTL mapping uses buffering to detect interaction partners. 
   When we split the population based on genotype at a QTL marker, only related phenotypes 
@@
   caused by regulatory phenotypes because variation of these regulators causes many 'downstream' 
-  phenotypes to show differential correlation with the inducing phenotype, because QCL relies on 
-  correlation changes as opposed to differences in phenotype mean QCL is unaffected by (p2).
+  phenotypes to show differential correlation with the inducing phenotype, because CTL relies on 
+  correlation changes as opposed to differences in phenotype mean CTL is unaffected by (p2).
 
 \emph{\sffamily Choosing a start/seed probe (Optional)}\\
-  Choose a phenotype to subject to QCL analysis, this can be your favourite gene, 
+  Choose a phenotype to subject to CTL analysis, this can be your favourite gene, 
   protein, metabolite or classical phenotype.
 
-\emph{\sffamily Single phenotype QCL mapping}\\
+\emph{\sffamily Single phenotype CTL mapping}\\
   Using genotype data we can split individuals at each marker in two groups (A or B).
@@
   phenotypes in groups A and B. Collect all the result vectors into two matrices A 
-  and B, and calculate the QCL matrix by taking the squared difference between matrix A 
+  and B, and calculate the CTL matrix by taking the squared difference between matrix A 
   and B.
@@
 \emph{\sffamily Significance testing}\\
-  To compare to QTL mapping we need to convert the QCLscores to LOD scores. Significance 
-  of QCL were assessed by using permutations, in each round the link between genotype and 
+  To compare to QTL mapping we need to convert the CTLscores to LOD scores. Significance 
+  of CTL were assessed by using permutations, in each round the link between genotype and 
   phenotype is broken, by redistributing at random genotypes amongst individuals not 
-  allowing for duplicates. The QCL on these new individuals allows for estimation of 
-  the underlying null-distribution of QCL scores. We frequently observe QCL scores in 
-  real data which is higher than any QCL score obtained during permutation. When this 
+  allowing for duplicates. The CTL on these new individuals allows for estimation of 
+  the underlying null-distribution of CTL scores. We frequently observe CTL scores in 
+  real data which is higher than any CTL score obtained during permutation. When this 
   happens we use a generalized Pareto distribution (GPD) to estimate the extreme tail 
@@
 
-\emph{\sffamily Creating single trait QCL profiles}\\
-  After permutation single trait QCL profiles can be obtained by transforming the QCL scores 
+\emph{\sffamily Creating single trait CTL profiles}\\
+  After permutation single trait CTL profiles can be obtained by transforming the CTL scores 
   into LOD scores. For this we use a look-up table of scores obtained during permutation, we 
-  assess the likelyhood and take the $-log10(P.value)$ of every QCL score see Fig. 
-  \ref{fig:singletraitLOD}. This is done 1) to compare QTL profiles with summarized QCL profiles 
+  assess the likelyhood and take the $-log10(P.value)$ of every CTL score see Fig. 
+  \ref{fig:singletraitLOD}. This is done 1) to compare QTL profiles with summarized CTL profiles 
   (an example in Fig. \ref{fig:singletrait}), and 2) Show the contribution of each phenotype 
@@
 
-\emph{\sffamily Visualizing the Phenotypes x Phenotypes QCL relations}\\
-  In Fig. \ref{fig:singletraitLOD} we see the additional information obtained by QCL mapping summarized 
+\emph{\sffamily Visualizing the Phenotypes x Phenotypes CTL relations}\\
+  In Fig. \ref{fig:singletraitLOD} we see the additional information obtained by CTL mapping summarized 
   graphically.  We can also calculate the full Phenotypes x Phenotypes matrix. This shows the summed 
@@
 \emph{\sffamily Visualizing the Phenotypes x Marker relations}\\
-  When plotting multiple QCL profiles as a heatmap, QCLs are seen mapping to the 
+  When plotting multiple CTL profiles as a heatmap, CTLs are seen mapping to the 
   same genetic location. At these locations, regulation is taking place. We expect 
-  to find one (or more) of the genes to show a cis-QTL, this is because a QCL can 
+  to find one (or more) of the genes to show a cis-QTL, this is because a CTL can 
   only exist if there is a QTL inducing it.
@@
 
-  We show two examples in which we reconstruct the genetic architecture using QCL mapping. 
+  We show two examples in which we reconstruct the genetic architecture using CTL mapping. 
   Both datasets are available inside the R package, as well as the code to analyse these 
@@
   \subsection{Example 1. A. Thaliana}
-  QCL mapping was performed on 24 characterized metabolities measured in Arabidopsis 
-  Thaliana. A significance threshold for QCL mapping was determined for each trait 
+  CTL mapping was performed on 24 characterized metabolities measured in Arabidopsis 
+  Thaliana. A significance threshold for CTL mapping was determined for each trait 
   by using 1000 permutations, the outcomes can be seen in Fig \ref{fig:histPermAra}. 
-  We used a very stringent QCL threshold at P < 0.001 to use as 'significant' QCL 
+  We used a very stringent CTL threshold at P < 0.001 to use as 'significant' CTL 
   score for further analysis.
@@
 
-  \emph{\sffamily Single trait QCL}\\
-  The 24 QCL profiles were compared to the QTL profiles previously determined. We observed 3 cases:\\
-  1) Additional QCL found, where no QTL was present as seen in the QCL profile X3.Hydroxypropyl 
+  \emph{\sffamily Single trait CTL}\\
+  The 24 CTL profiles were compared to the QTL profiles previously determined. We observed 3 cases:\\
+  1) Additional CTL found, where no QTL was present as seen in the CTL profile X3.Hydroxypropyl 
   (Fig. \ref{fig:singletraitLOD}), these are the cases in which the loss of variation in the genes 
-  underneath the QCL is not causing variation differences in the observed phenotype, but still 
+  underneath the CTL is not causing variation differences in the observed phenotype, but still 
   we're able to detect the correlation differences.\\
-  2) QCL co-localized with QTL, in this situation QCL gives a detailed view of which phenotypes 
+  2) CTL co-localized with QTL, in this situation CTL gives a detailed view of which phenotypes 
   are underneath its peek, we interpret this as variation transfered from the phenotypes underneath 
-  the QCL to the phenotype investigated.\\
-  3) QTL with no QCL, we interpret this as the source of the variation. Analysis in this way of 
-  all 24 Single trait QCL profiles combined with their QTL profiles allows us to reconstruct the 
+  the CTL to the phenotype investigated.\\
+  3) QTL with no CTL, we interpret this as the source of the variation. Analysis in this way of 
+  all 24 Single trait CTL profiles combined with their QTL profiles allows us to reconstruct the 
   source of natural variation and follow it down the network.
 
-  \emph{\sffamily QCL combined with QTL}\\
-  Plots of the results of the QCL mapping in Arabidopsis can be found in 
+  \emph{\sffamily CTL combined with QTL}\\
+  Plots of the results of the CTL mapping in Arabidopsis can be found in 
   Figure XX Phenotype to Marker interaction, which is comparable to the QTL mapping 
-  (Fig. XX). The additional information obtained by QCL mapping is shown in a heatmap 
+  (Fig. XX). The additional information obtained by CTL mapping is shown in a heatmap 
   of the Phenotype to Phenotype interaction matrix (figure XX) and the same information 
-  but now represented as an interaction network is seen in figure XX and in the QCL plot
+  but now represented as an interaction network is seen in figure XX and in the CTL plot
 
@@
   Using the derived data we can begin to reconstruct the network. We'll start with a metabolite 
-  which shows mappable variation changes, but no QCL colocatizing (Inducer).
+  which shows mappable variation changes, but no CTL colocatizing (Inducer).
 
   \subsection{Example 2. S. Cerevisiae}
-  QCL mapping is computationaly expensive (We need 1000+ permutations for each phenotype), 
+  CTL mapping is computationaly expensive (We need 1000+ permutations for each phenotype), 
   and for the S. Cerevisiae we pre-selected a background of 300 probes which showed the 
   highest variation, and had been previously been identified as having QTLs. Significance 
-  thresholds for QCL mapping were determined for each trait by using 1000 permutations 
+  thresholds for CTL mapping were determined for each trait by using 1000 permutations 
   ({\emph Data not shown}).
 
-  \emph{\sffamily Single trait QCL}\\
+  \emph{\sffamily Single trait CTL}\\
 
@@
   However results on complex traits have been less then satisfactory with only small 
-  amounts of heritability explained by individuals loci. QCL mapping however 'borrows' 
+  amounts of heritability explained by individuals loci. CTL mapping however 'borrows' 
   information from each phenotype in the analysis set, and enables extraction of 
   phenotype to marker to phenotype networks, the flow of variation in the network. 
-  Furthermore QCLs seem to 'transfer' from phenotype to phenotypes in a stabile 
+  Furthermore CTLs seem to 'transfer' from phenotype to phenotypes in a stabile 
   fashion, which gives greater detail and confidence when reconstructing genetic 
-  architecture. Also QCL allows us to better identify gene transcript introducing 
+  architecture. Also CTL allows us to better identify gene transcript introducing 
   variation, by their absense under QTL peeks which introduce variation, compared 
@@
 %\begin{figure}[ht]
-%  \centerline{\includegraphics[height=7cm]{qcl.pdf}}
+%  \centerline{\includegraphics[height=7cm]{CTL.pdf}}
 %  \vspace{-15pt}
-%  \caption{{\scriptsize QCL mapping: We start by looking at a single genetic marker and dividing our populaiton into two groups.
+%  \caption{{\scriptsize CTL mapping: We start by looking at a single genetic marker and dividing our populaiton into two groups.
 %  We start by choosing a phenotype (red), and use its expression to detect the red module (high correlation in gray genotype). 
@@
 %  \vspace{-20pt}
-%  \label{fig:qcl}
+%  \label{fig:CTL}
 %\end{figure}
@@
 %\begin{figure}[ht]
-%  \centerline{\includegraphics[height=6cm]{qcl_n.pdf}}
+%  \centerline{\includegraphics[height=6cm]{CTL_n.pdf}}
 %  \vspace{-15pt}
@@
 %  \vspace{-20pt}
-%  \label{fig:qcl_n}
+%  \label{fig:CTL_n}
 %\end{figure}
@@
 @ 
-  \caption{{\scriptsize QCL scores obtained during permutations (A. Thaliana) additional information obtained by QCL mapping, 
+  \caption{{\scriptsize CTL scores obtained during permutations (A. Thaliana) additional information obtained by CTL mapping, 
   which can be used to reconstruct the underlying genetic architecture. }}
@@
 @
-  \caption{{\scriptsize QCL profile (A. Thaliana, X3.Hydroxypropyl) transformed to LOD scores compared to the QTL profile of the same phenotype, We see an 
+  \caption{{\scriptsize CTL profile (A. Thaliana, X3.Hydroxypropyl) transformed to LOD scores compared to the QTL profile of the same phenotype, We see an 
   additional significant peek for this phenotype (Between Marker 50 to 60) }}
@@
 op <- par(cex=0.5)
-QCLasLOD(ath.result[[1]], ath.result[[1]]$qtl)
+CTLasLOD(ath.result[[1]], ath.result[[1]]$qtl)
 op <- par(cex=1)
 @ 
-  \caption{{\scriptsize QCL profile (A. Thaliana, X3.Hydroxypropyl) transformed to LOD scores. Shown is the likelyhood contribution of each of 
-  the 24 metabolites to the QCL profile.}}
+  \caption{{\scriptsize CTL profile (A. Thaliana, X3.Hydroxypropyl) transformed to LOD scores. Shown is the likelyhood contribution of each of 
+  the 24 metabolites to the CTL profile.}}
   \label{fig:singletraitLOD}
@@
 @ 
-  \caption{{\scriptsize Phenotypes x Phenotypes matrix (A. Thaliana) additional information obtained by QCL mapping, 
+  \caption{{\scriptsize Phenotypes x Phenotypes matrix (A. Thaliana) additional information obtained by CTL mapping, 
   which can be used to reconstruct the underlying genetic architecture}}

### 2012-02-09 inst/methodPaper/generate-pdf.sh
Generate PDF script

@@ -1,0 +1,13 @@
+#! /bin/sh
+#
+# Create PDF - run this script several times to get the citations
+# correct.
+
+bibtex document
+# perl nat2jour.pl -maxauth 2 BioGem
+latex document.tex
+if [ "$1" = "--pdf" ] ; then
+  dvipdf document.dvi document.pdf
+else
+  echo "Skipped generating PDF!"
+fi

### 2012-02-09 inst/methodPaper/document.tex
Use EPS instead of PDF - PDF is not platform supported by latex

@@
     \subfigure[CTL profiles]{
-      \includegraphics[keepaspectratio,scale=.3]{img/ctl_qtl.pdf}
+      \includegraphics[keepaspectratio,scale=.3]{img/ctl_qtl.eps}
       \label{subfiga}
@@
     \subfigure[Traits x Traits interaction matrix]{
-      \includegraphics[keepaspectratio,scale=.3]{img/ctl_pxp.pdf}
+      \includegraphics[keepaspectratio,scale=.3]{img/ctl_pxp.eps}
       \label{subfigb}
@@
     \subfigure[Reconstructed network]{
-      \includegraphics[keepaspectratio,scale=.3]{img/ctl_net.pdf}
+      \includegraphics[keepaspectratio,scale=.3]{img/ctl_net.eps}
       \label{subfigc}

### 2012-02-09 inst/methodPaper/document.tex
Document

@@
 \usepackage{amsmath}
+
+\usepackage{color,soul}  % for markup
+\setul{1ex}{0.8ex}
+\definecolor{orange}{rgb}{1,0.5,0}
+\newcommand{\marked}[1]{\hl{#1}}
+\newcommand{\comment}[1]{\hl{\bf FIXME! #1}}
 
@@
 \title[CTLmapping]{CTL mapping - The added value from segregating correlation}
+
 \author[Arends \emph{et al.}]{
   Danny Arends\,$^{1,\footnote{to whom correspondence should be addressed}}$,
+  Pjotr Prins\,$^{1,2}$
   and Ritsert C. Jansen\,$^{1}$
@@
 \address{
-  $^{1}$Groningen Bio-informatics Centre, University of Groningen, Groningen, The Netherlands.
+  $^{1}$Groningen \hl{Bio-informatics} Centre, University of Groningen, Groningen, The Netherlands.
+  $^{2}$Laboratory of Nematology, Wageningen University, The Netherlands.
 }
@@
 \maketitle
+
+\comment{Title is too cute, and looks too much like an earlier paper;
+also should be more descriptive}
+
+\comment{3x Groningen in address. And does that dash belong there?}
+
 \begin{abstract}
-  \section{Motivation:}
-  Here we present Correlated Traits Locus (CTL) mapping, a novel approach 
-  to detect genetic regulation of traits in natural and inbred populations. 
-  CTL mapping allows association of trait correlation differences to genetic 
-  markers, in almost the same way as QTL mapping associates genetic  markers with differences in expression.
+  \section{Summary:}
 
-  We use CTL mapping to reconstruct underlying genetic architecture, and provide insight 
-  into regulation of (gene) expression. CTL information in combination with classical QTL 
-  information leads to a more detailed insight into the genetic architecture underlying 
-  complex traits. CTL mapping provides a missing link in regulatory network reconstruction.
+  % Summary moet 1 of 2 zinnen zijn! Zoiets beter?
 
-  \section{Results:}
+  Mapping of correlated traits loci (CTL) to the genome potentially
+  allows the detection of genetic regulation of traits by tracking
+  correlation differences at genetic marker locations. CTL can be
+  combined with quantitative trait loci (QTL) in regulatory network
+  reconstruction.
+
+  %  , in almost the same way as QTL mapping associates genetic
+  %  markers with differences in expression.  <- als je dit zegt is
+  %  het niet zo nieuw meer ;)
+
+  % We use CTL mapping to reconstruct underlying genetic architecture, and provide insight 
+  % into regulation of (gene) expression. CTL information in combination with classical QTL 
+  % information leads to a more detailed insight into the genetic architecture underlying 
+  % complex traits. CTL mapping
+
+  \section{Availability:}
+
   The software is available under the terms of GNU General Public License, version 3
@@
   \section{Contact:} 
-  \href{Danny.Arends@Gmail.com}{Danny.Arends@Gmail.com}
+  \href{Danny.Arends\hl{@Gmail.com}}{Danny.Arends@\hl{Gmail.com}}
 \end{abstract}
+
+\comment{Use university E-mail}
+
+\comment{License: BSD license is preferred these days}
+
+\comment{Availability: Bioinformatics Journal wants open source software only,
+availability should reflect that - i.e. a github.org repo. They won't
+like your option ;)}
 
@@
   {\emph {\bf a}}: Example CTL + QTL profiles. Seen are the QTL profile on top of the CTL profile. 
-  The first peak is a CTL, we don't observe a QTL for this location. The second peak is a QTL peek, with no underlying CTL. 
+  The first peak is a CTL, we don't observe a QTL for this location.
+  The second peak is a QTL peek, with no underlying CTL. \hl{You
+  should explain P1,P2 and P3 in the caption too! A figure should be
+  read on its own} 
   {\emph {\bf b}}: Traits x Traits interaction matrix (TxT matrix). Information obtained using CTL mapping, 
@@
 \vspace{-0.6cm}
-\section{Results}
+\section{FEATURES}
   CTL mapping was performed and a per trait CTL thresholds of $P < 0.001$ was 
@@
 
-\section{Conclusion \& Discussion}
+\section{Conclusion}
   Microarray experiments have identified genes differentially expressed between various 

### 2012-02-19 Rctl/inst/doc/manual.Rnw
Moved the R page into Rctl folder

--- 2011-10-28 inst/doc/manual.Rnw
@@
 % Danny Arends
-% The qcl mapping routine
+% The ctl mapping routine
+%\VignetteIndexEntry{Manual}
 %
@@
 \vspace*{1in}
-\textbf{\Large QCL mapping in R }
+\textbf{\Large CTL mapping in R }
 
@@
 Groningen Bioinformatics centre \& GCC \\
-Technical Report \# 214 \\[24pt]
+revision \# 1 \\[24pt]
 XX Okt 2011}
@@
 \textbf{\sffamily Abstract}: Introduction into quanitative correlation 
-locus (QCL) mapping using the R package qcl
+locus (CTL) mapping using the R package CTL
 \vfill
@@
 
-\textbf{\sffamily QCLscan and QCLscan.network}
+\textbf{\sffamily CTLscan and CTLscan.network}
 \nopagebreak
 
-Explains how to map QCL using the QCLscan and QCLscan.network functions.
+Explains how to map CTL using the CTLscan and CTLscan.network functions.
 
 \bigskip
-\textbf{\sffamily Significance using QCLpermute}
+\textbf{\sffamily Significance using CTLpermute}
 \nopagebreak
 
-Explains how to determine QCL significance using the QCLpermute function.
+Explains how to determine CTL significance using the CTLpermute function.
 
 \bigskip
-\textbf{\sffamily Advanced plots and QCL networks}
+\textbf{\sffamily Advanced plots and CTL networks}
 \nopagebreak
@@
 {\small
-\noindent \textbf{\sffamily Package qcl function overview} \\
+\noindent \textbf{\sffamily Package CTL function overview} \\
 \noindent \begin{tabular}{ll} 
 \hspace*{25mm} & \hspace*{103mm} \\ \hline
-QCLscan & Genome wide scan for QCL \\
-QCLscan.network & Scan for QCL using start gene(s) \\
-QCLpermute & Significance thresholds for QCL by permutation \\
-plot.QCL & Plot QCL profiles at increasing cut-offs \\
-image.QCL & Image a multiple phenotype scan \\
+CTLscan & Genome wide scan for CTL \\
+CTLscan.network & Scan for CTL using start gene(s) \\
+CTLpermute & Significance thresholds for CTL by permutation \\
+plot & Plot CTL profiles at increasing cut-offs \\
+image & Image a multiple phenotype scan \\
 \hline

### 2012-05-09 Rctl/R/ctl.network.R
[Fixes] Ath result dataset

@@ -8,9 +8,9 @@
 # Network routines for CTL analysis
 #
 
-write.marker.annotation <- function(CTLscan,chr.edges){
+write.marker.attributes <- function(QTLscan, chr.edges){
   chr   <- 1
-  namez <- colnames(attr(CTLscan,"qtl"))
+  namez <- colnames(QTLscan)
   cat("", file="node_attributes.txt")
   #Connections between genetic markers
   for(m1 in 1:(length(namez)-1)){
@@ -23,14 +23,14 @@ write.marker.annotation <- function(CTLscan,chr.edges){
   }
   #Annotate the last marker on the last chromosome
   if(!missing(chr.edges)) cat(paste(namez[length(namez)],"\tCHR",chr,"\n",sep=""),file="node_attributes.txt",append=TRUE)
-  if(!missing(chr.edges)) cat("Wrote annotation for",length(namez),"markers on",chr,"chromosomes\n")
+  if(!missing(chr.edges)) cat("Wrote attributes for",length(namez),"markers on",chr,"chromosomes\n")
 }
 
 QTLnetwork <- function(CTLscan, lod.cutoff = 5, chr.edges, filename){
   idx   <- 1
   edges <- 0
   chr   <- 1
-  write.marker.annotation(CTLscan, chr.edges)
+  write.marker.attributes(attr(CTLscan,"qtl"), chr.edges)
   if(missing(filename)){
     filename <- "network_qtl.sif"
     cat("",file=filename)
@@ -50,7 +50,7 @@ QTLnetwork <- function(CTLscan, lod.cutoff = 5, chr.edges, filename){
   }
   for(x in apply(attr(CTLscan,"qtl"),1,function(x){which(x > lod.cutoff)})){
     for(marker in names(x)){
-      cat(paste(name(CTLscan[[idx]]),"\tQTL\t",marker,"\t",round(attr(CTLscan,"qtl")[idx,marker],digits=1),"\n",sep=""),file=filename,append=TRUE)
+      cat(paste(ctl.name(CTLscan[[idx]]),"\tQTL\t",marker,"\t",round(attr(CTLscan,"qtl")[idx,marker],digits=1),"\n",sep=""),file=filename,append=TRUE)
       edges <- edges+1
     }
     idx <- idx + 1
@@ -68,7 +68,7 @@ CTLnetwork <- function(CTLscan, lod.cutoff = 5, chr.edges, add.qtl = TRUE, verbo
[…]

### 2012-05-09 Rctl/R/helper.functions.R

@@ -12,7 +12,11 @@
 # - Chromosome edge locations from mapfile
 # - Get the top-correlated metabolites
 
-ctl.name <- function(CTLobject){ return(attr(CTLobject$ctl,"name")); }
+ctl.name      <- function(CTLscan){ return(attr(CTLscan$ctl,"name")); }
+ctl.names     <- function(CTLobject){ unlist(lapply(CTLobject,function(x){return(attr(x$ctl,"name"));})) }
+ctl.ctlmatrix <- function(CTLscan){ return(CTLscan$ctl); }
+ctl.lodmatrix <- function(CTLscan){ return(CTLscan$l); }
+
 remove.diag <- function(x){ return(x*lower.tri(x) + x*upper.tri(x)); }
 up <- function(){abs(seq(-2,-0,0.1))/2} 
 dw <- function(){seq(0.1,2,0.1)/2}

### 2012-05-09 Rctl/R/ctl.image.R
[Rctl] Updating docs to match code

@@ -8,23 +8,9 @@
 # Image plot routines for CTL analysis
 #
 
-image.CTLobject <- function(x, against = c("markers","phenotypes"), onlySignificant = FALSE, significance = 0.05, col=c("white",gray.colors(10)[10:1]), do.grid=TRUE, grid.col = "black", verbose = FALSE, add=FALSE, ...){
+image.CTLobject <- function(x, against = c("markers","phenotypes"), significance = 0.05, col=whiteblack, do.grid=TRUE, grid.col = "black", verbose = FALSE, add=FALSE, ...){
   colorrange <- col
-  mymatrix <- NULL
-  mynames <- NULL
-  for(p in 1:length(x)){
-    if(onlySignificant){
-      maxes <- apply(abs(x[[p]]$ctl),1,max)
-      if(max(maxes) > getPermuteThresholds(x[[p]], significance)[1]){
-        mymatrix <- rbind(mymatrix,apply(x[[p]]$l,2,sum))
-        mynames <- c(mynames,attr(x[[p]]$ctl,"name"))  
-      }
-    }else{
-      mymatrix <- rbind(mymatrix,apply(x[[p]]$l,2,sum))
-      mynames <- c(mynames,attr(x[[p]]$ctl,"name"))
-    }
-  }
-  rownames(mymatrix) <- mynames
+  mymatrix <- CTLprofiles(x, against, significance)
   mainlabel <- paste("CTL phenotypes vs",against[1],"at P-value <",significance)
   internal.image(mymatrix, colorrange, mainlabel,do.grid, grid.col)
 }
@@ -59,14 +45,10 @@ internal.image <- function(mymatrix, colorrange, mainlabel, do.grid, grid.col, a
   invisible(mymatrix)
 }
 
-QTLimage <- function(x, onlySignificant = FALSE, significance = 0.05, do.grid=TRUE, grid.col = "black", verbose = FALSE, ...){
+QTLimage <- function(x, do.grid=TRUE, grid.col = "black", verbose = FALSE, ...){
   colorrange <- c("white",gray.colors(10)[10:1])
   mymatrix <- attr(x,"qtl")
-  if(onlySignificant){
-    mainlabel <- paste("QTL heatmap at P-value <", significance)
-  }else{
-    mainlabel <- paste("QTL heatmap")
-  }
+  mainlabel <- paste("QTL heatmap")
   internal.image(mymatrix, colorrange, mainlabel,do.grid, grid.col)
 }
 

### 2012-05-09 Rctl/R/ctl.network.R

@@ -26,16 +26,16 @@ write.marker.attributes <- function(QTLscan, chr.edges){
   if(!missing(chr.edges)) cat("Wrote attributes for",length(namez),"markers on",chr,"chromosomes\n")
 }
 
-QTLnetwork <- function(CTLscan, lod.cutoff = 5, chr.edges, filename){
+QTLnetwork <- function(CTLobject, lod.threshold = 5, chr.edges, filename){
   idx   <- 1
   edges <- 0
   chr   <- 1
-  write.marker.attributes(attr(CTLscan,"qtl"), chr.edges)
+  write.marker.attributes(attr(CTLobject,"qtl"), chr.edges)
   if(missing(filename)){
     filename <- "network_qtl.sif"
     cat("",file=filename)
   }
-  namez <- colnames(attr(CTLscan,"qtl"))
+  namez <- colnames(attr(CTLobject,"qtl"))
   #Connections between genetic markers
   for(m1 in 1:(length(namez)-1)){
     if(missing(chr.edges)){
@@ -48,9 +48,9 @@ QTLnetwork <- function(CTLscan, lod.cutoff = 5, chr.edges, filename){
       }
     }
   }
-  for(x in apply(attr(CTLscan,"qtl"),1,function(x){which(x > lod.cutoff)})){
+  for(x in apply(attr(CTLobject,"qtl"),1,function(x){which(x > lod.threshold)})){
     for(marker in names(x)){
-      cat(paste(ctl.name(CTLscan[[idx]]),"\tQTL\t",marker,"\t",round(attr(CTLscan,"qtl")[idx,marker],digits=1),"\n",sep=""),file=filename,append=TRUE)
+      cat(paste(ctl.name(CTLobject[[idx]]),"\tQTL\t",marker,"\t",round(attr(CTLobject,"qtl")[idx,marker],digits=1),"\n",sep=""),file=filename,append=TRUE)
       edges <- edges+1
     }
     idx <- idx + 1
@@ -58,17 +58,17 @@ QTLnetwork <- function(CTLscan, lod.cutoff = 5, chr.edges, filename){
   cat("Wrote",edges,"edges to",filename,"\n")
 }
 
-CTLnetwork <- function(CTLscan, lod.cutoff = 5, chr.edges, add.qtl = TRUE, verbose = FALSE){
+CTLnetwork <- function(CTLobject, lod.threshold = 5, chr.edges, add.qtl = TRUE, verbose = FALSE){
[…]

### 2012-05-09 Rctl/R/helper.functions.R

@@ -21,6 +21,7 @@ remove.diag <- function(x){ return(x*lower.tri(x) + x*upper.tri(x)); }
 up <- function(){abs(seq(-2,-0,0.1))/2} 
 dw <- function(){seq(0.1,2,0.1)/2}
 redblue <- function(){c(rgb(up,0,0), rgb(0,0,dw))}
+whiteblack <- function(){c("white",gray.colors(10)[10:1])}
 
 get.chr.edges <- function(map_info){
   unlist(lapply(unique(map_info[,1]),function(x){max(which(map_info[,1]==x));}))

### 2012-05-09 Rctl/inst/doc/manual.Rnw
[User Manual] Updates

@@
 
-\textbf{\sffamily Abstract}: Introduction into quanitative correlation 
-locus (CTL) mapping using the R package CTL
+\textbf{\sffamily Abstract}: Introduction into quanitative correlation locus (CTL) mapping using 
+the R package 'ctl'
 \vfill
@@
 
-Loading your data
+\emph{Loading your Excel / CSV data}
+Use the example excel file and simply replace the example data by your own then: 
+  $$File->Save As->TODO$$
+
+After you have stored the data as plain text open the files in a text editor (NOT WORD, for 
+windows Notepad++ is a good option) and check to see if your data is stored correctly, if 
+not reformat your data in both the genotype.csv and the phenotype.csv files to match the 
+following:
+
+<<echo=TRUE,eval=FALSE>>=
+setwd("<PATH_TO_DATA>")
+genotypes <- read.csv("genotypes.csv",row.names=1,col.names=TRUE)
+phenotypes <- read.csv("phenotypes.csv",row.names=1,col.names=TRUE)
+ctls <- CTLscan(genotypes, phenotypes)
+@
+
+\emph{Use an R/qtl formatted dataset}
+Provided is the main interface functions to R/qtl: $CTLscan.cross()$ this functions accepts and 
+R/qtl formatted cross object as input and will scan CTL, perform permutations and transform the
+detected differential correlation to LOD score matrices.
+As an example we show the code to scan the multitrait dataset provided in R/qtl:
+
+<<echo=TRUE,eval=FALSE>>=
+require(qtl)        #load the R/qtl package
+data(multitrait)    #load the dataset
+ctls <- CTLscan.cross(multitrait, pheno.col=1:2)
+@
 
@@
 
-\bigskip
-\textbf{\sffamily Significance using CTLpermute}
-\nopagebreak
+\emph{CTLscan options}
+The genotype matrix and the $geno.enc$ param, the phenotype matrix, the important $n.perms$ param, 
+which should be set to as many as possible, especially when publishing any results, but for a quick 
+scan 100 is ussualy sufficient.
 
-Explains how to determine CTL significance using the CTLpermute function.
+When experimental setups are more complex a user can supply QTL results via the $have.qtl$ option.
+
+<<echo=TRUE,eval=FALSE>>=
+require(qtl)        #load the R/qtl package
+data(multitrait)    #load the dataset
+qtls <- scanone(multitrait, pheno.col=1:24)[,-c(1,2)]
+ctls <- CTLscan.cross(multitrait, pheno.col=1:2, have.qtl=qtls)
+@
+
+, when no qtls are supplied the internal QTL mapping routine is used, which can deal with just a 
+single condition per individual. An environmental condition can be specified by the $conditions$ 
+param and are also passed to the internal QTL mapping algorithm.
+
+<<echo=TRUE,eval=FALSE>>=
+#Specify that individuals 1:81 were in condition A / e.g. treatment
+#Specify that individuals 82:162 were in condition B / e.g. control
+conditions <- c(rep(1,81),rep(2,81))
+ctls <- CTLscan(ath.metab$genotypes, ath.metab$phenotypes, conditions=)
+@
 
@@
 
+\bigskip
+\textbf{\sffamily Big data and Dctl}
+\nopagebreak
+
+Explains how to use the D version to map big datsets
+
+\bigskip
+\textbf{\sffamily Significance using CTLpermute}
+\nopagebreak
+
+Explains how to determine CTL significance using the CTLpermute function.
+
 \newpage
@@
 \hspace*{25mm} & \hspace*{103mm} \\ \hline
-CTLscan & Genome wide scan for CTL \\
-CTLscan.network & Scan for CTL using start gene(s) \\
-CTLpermute & Significance thresholds for CTL by permutation \\
+CTLscan & Main function for Genome wide scan for Correlated Trait Loci \\
+CTLscan.network & Scan only a subset of traits for CTL using start gene(s) \\
+CTLpermute & Significance thresholds for CTL by permutation\\
 plot & Plot CTL profiles at increasing cut-offs \\

### 2012-05-09 Rctl/inst/doc/manual.Rnw
[Rctl] More working on the manual

@@
 
-\textbf{\sffamily Abstract}: Introduction into quanitative correlation locus (CTL) mapping using 
-the R package 'ctl'
+\textbf{\sffamily Abstract}: Correlated Trait Loci (CTL) mapping using the R package 'ctl'
 \vfill
@@
 
-\textbf{\sffamily Introduction}
-Some introduction text
+\textbf{\sffamily Some words in advance}
+\nopagebreak
+
+Welcome reader to this tutorial, you'll be thrown into the world of network reconstruction 
+using differential correlations. This methodology CTL mapping is not the easiest but the 
+results can be of great help unraveling trait to trait interactions in all types of 
+eukaryotes. In this tutorial I'll try to use as few difficult concepts as possible. But
+some always $sneak$ in.\\
+\\
+Enjoy this tutorial, and good luck hunting that heritability.\\
+Danny Arends\\
+\\
+PS. Use a text-editor\\
+PPS. Use a repository/backup system
+
+\bigskip
+
+\textbf{\sffamily Background \& Methodology}
+\nopagebreak
+
+Some introduction on the methodology, skip this chapter first (Do come back later) if you 
+have a dataset and just want to use the software.
 
@@
 
-\emph{Loading your Excel / CSV data}
+\emph{Loading your Excel / CSV data}\\
 Use the example excel file and simply replace the example data by your own then: 
@@
 genotypes <- read.csv("genotypes.csv",row.names=1,col.names=TRUE)
-phenotypes <- read.csv("phenotypes.csv",row.names=1,col.names=TRUE)
-ctls <- CTLscan(genotypes, phenotypes)
+traits <- read.csv("traits.csv",row.names=1,col.names=TRUE)
+genotypes[1:5,1:10] #Show the first 5 individuals, 10 markers
+traits[1:5,1:10] #Show the first 5 individuals, 10 traits
 @
 
-\emph{Use an R/qtl formatted dataset}
+\emph{Use an R/qtl formatted dataset}\\
 Provided is the main interface functions to R/qtl: $CTLscan.cross()$ this functions accepts and 

### 2012-05-09 Rctl/inst/doc/manual.Rnw
[Rctl] More working on the manual

@@
 \\
-PS. Use a text-editor\\
-PPS. Use a repository/backup system
+PS. Use a good text editor\\
+PPS. Use a repository / backup system
 
@@
 
-Some introduction on the methodology, skip this chapter first (Do come back later) if you 
+Some introduction on the methodology, skip this chapter first (DO come back later) if you 
 have a dataset and just want to use the software.
@@
 <<echo=TRUE,eval=FALSE>>=
-require(qtl)        #load the R/qtl package
-data(multitrait)    #load the dataset
-ctls <- CTLscan.cross(multitrait, pheno.col=1:2)
+require(qtl)        #Loads the R/qtl package
+data(multitrait)    #Loads the dataset
+multitrait          #Print basic dataset information
 @
@@
 Explains how to map CTL using the CTLscan and CTLscan.network functions.
+<<echo=TRUE,eval=FALSE>>=
+require(ctl)        #Loads the R/ctl package
+data(ath.metab)
+geno <- ath.metab$genotypes #Short name
+traits <- ath.metab$phenotypes #Short name
+
+#Scan all phenotypes for CTLs, using the default options
+#NOTE: Could take some time
+ctls <- CTLscan(geno,traits)
+@
 
@@
 conditions <- c(rep(1,81),rep(2,81))
-ctls <- CTLscan(ath.metab$genotypes, ath.metab$phenotypes, conditions=)
+ctls <- CTLscan(geno, traits, conditions=conditions)
 @

### 2012-05-10 Rctl/inst/doc/manual.Rnw
[Rctl] Manual updates

@@
 not reformat your data in both the genotype.csv and the phenotype.csv files to match the 
-following:
+format below. {\color{red}\emph{\bf{NOTE: Individuals must match between the genotype and 
+the phenotype file...}}} The phenotypes.csv file containing the genotype matrix is stored 
+individuals x genetic marker:
 
+\begin{tabular}{lrrrrr}
+      & Marker1 & Marker2 & Marker3 & ... & MarkerN \\
+Ind1  & 1 & 1 & 2 & ... & 2 \\
+Ind2  & 1 & 1 & 2 & ... & 2 \\
+Ind3  & 1 & 1 & 1 & ... & 2 \\
+...   & ... & ... & ... & ... & ... \\
+IndN  & 1 & 1 & 1 & ... & 2 \\
+\end{tabular}
+
+The phenotypes.csv file containing individuals x trait measurements:
+
+\begin{tabular}{lrrrrr}
+      & Trait1 & Trait2 & Trait3 & ... & TraitN \\
+Ind1  & 5.8    & 11.0   & 9.0    & ... & 1.6 \\
+Ind2  & 6.3    & 12.2   & NA     & ... & 1.3 \\
+Ind3  & 5.1    & 11.1   & 12.3   & ... & 2.0 \\
+...   & ...    & ...    & ...    & ... & ... \\
+IndN  & 9.8    & 15.8   & 23.0   & ... & 3.4 \\
+\end{tabular}
+
+After verifying your files are formatted correctly (this is the source of most errors).
+Start R and load in the data, check that the output looks like below:
 <<echo=TRUE,eval=FALSE>>=
@@
 
+\emph{Adding genetic map information}
+This step is optional, but plots and generated networks look much nicer when chromosome locations
+of genetic markers are supplied. The 'mapinfo' object is matrix with 3 columns: "Chr" - the 
+chromosome number, "cM" - the location of the marker in centiMorgans and the 3rd column "Mbp" - The 
+location of the marker in Mega basepairs. The structure of the $mapinfo.csv$ file:
+
+\begin{tabular}{lrrr}
+      & Chr    & cM     & Mbp \\
+Ind1  & 1    & 1.0   & 0.01   \\
+Ind2  & 1    & 1.2   & 0.34   \\
+Ind3  & 1    & 3.1   & 1.3    \\
+...   & ...  & ...   & ...    \\
+IndN  & 24   & 120.8 & 20.3   \\
+\end{tabular}
+
+We load the genetic map data into R by using the following command:
+
+<<echo=TRUE,eval=FALSE>>=
+mapinfo <- read.csv("mapinfo.csv",row.names=1,col.names=TRUE)
+traits[1:5,1:3] #Show the first 5 marker records
+@
+After this we can
 \bigskip
@@
 
-Some more advanced topics
+So now that we have our ctls calculated we can visualize the results in multiple ways.
+The easiest way is first to plot the two overview CTL heatmaps. The first heatmap shows 
+the summarized LOD scores for the Trait X Marker CTLs. This plot normally looks very 
+similar to a QTL heatmap:
+<<echo=TRUE,eval=FALSE>>=
+image(ctls,against="markers")
+@
+The second heatmap shows the summarized LOD scores for the Trait * Trait CTLs, this 
+shows the summarized amount of evidence we find for each Trait * Trait interaction:
+<<echo=TRUE,eval=FALSE>>=
+image(ctls,against="phenotypes")
+@
+After the overview plots it is time to go more in depth, and look at some individual CTL 
+profiles the CTLobject 'ctls' contains multiple CTLscans. To print a summary of the first 
+scan use:
+<<echo=TRUE,eval=FALSE>>=
+ctls[[1]]
+@
+To plot other summaries just replace the 1 by the trait number. We can also use the plot 
+function on individual CTLscans and shows a colorful and detailed view of the 
+(selected) Trait x Marker X Traits interactions detected in the dataset.
+<<echo=TRUE,eval=FALSE>>=
+plot(ctls[[1]])
+@
 

### 2012-05-10 Rctl/inst/doc/manual.Rnw
[Rctl] Manual updates, add bigdata and ctlscan

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\documentclass[12pt]{article}
+\documentclass[10pt]{article}
 
@@
 
-\bigskip
+\newpage
 
@@
 not reformat your data in both the genotype.csv and the phenotype.csv files to match the 
-format below. {\color{red}\emph{\bf{NOTE: Individuals must match between the genotype and 
+format below. {\color{red} \emph{\bf{NOTE: Individuals must match between the genotype and 
 the phenotype file...}}} The phenotypes.csv file containing the genotype matrix is stored 
@@
       & Marker1 & Marker2 & Marker3 & ... & MarkerN \\
-Ind1  & 1 & 1 & 2 & ... & 2 \\
-Ind2  & 1 & 1 & 2 & ... & 2 \\
-Ind3  & 1 & 1 & 1 & ... & 2 \\
+Ind1  & 1 & 1 & 1 & ... & 2 \\
+Ind2  & 1 & 2 & 2 & ... & 1 \\
+Ind3  & 2 & 2 & 1 & ... & 1 \\
 ...   & ... & ... & ... & ... & ... \\
@@
 
-\emph{Adding genetic map information}
+\emph{Adding genetic map information}\\
 This step is optional, but plots and generated networks look much nicer when chromosome locations
@@
 @
-After this we can
+Finally... after all this we can start scanning QTL and CTL...
 \bigskip
 
-\textbf{\sffamily CTLscan and CTLscan.network}
+\textbf{\sffamily Scanning for CTL / $CTLscan$}
 \nopagebreak
 
-Explains how to map CTL using the CTLscan and CTLscan.network functions.
+We start explaining how to map CTL using the basic options of the $CTLscan$ function. This function 
+scan CTL and produces an output "CTLobject".
 <<echo=TRUE,eval=FALSE>>=
 require(ctl)                     #Loads the R/ctl package
-data(ath.metab)
+data(ath.metab)                  #Loads the example dataset
 geno <- ath.metab$genotypes      #Short name
@@
 
-#Scan all phenotypes for CTLs, using the default options
-#NOTE: Could take some time
+#Scan all phenotypes for CTLs, using the default options (Could take some time)
 ctls <- CTLscan(geno,traits)
 @
+If you have loaded an R/qtl "cross" object, use the $CTLscan.cross$ interface, this will automagically 
+extract the cross, check if the type of "cross" is supported, and deal with the $geno.enc$ parameter:
+<<echo=TRUE,eval=FALSE>>=
+library(qtl)
+data(multitrait)
+ctls <- CTLscan.cross(multitrait)
+@
 
-\emph{CTLscan options}
-The genotype matrix and the $geno.enc$ param, the phenotype matrix, the important $n.perms$ param, 
-which should be set to as many as possible, especially when publishing any results, but for a quick 
-scan 100 is ussualy sufficient.
+Now you might feel tempted to just dive into the results, But... first lets take a look at some 
+options that might apply to your experimental setup like treatments applied or growth conditions, 
+and which (if set incorrectly) could seriously reduce mapping power. Also you might want to add 
+your own/previous QTL results.
+
+\emph{CTLscan options}\\
+The genotype matrix and the $geno.enc$ parameter, the phenotype matrix, the important $n.perms$ 
+parameter, which should be set to as many as possible, especially when publishing any results, 
+but for a quick scan 100 is ussualy sufficient.
 
 When experimental setups are more complex a user can supply QTL results via the $have.qtl$ option.
+An example is shown below in which we use QTL results obtained by using single marker mapping via 
+$scanone$ and the new Multiple QTL mapping routine $mqmscan$ in R/qtl:
 
@@
 data(multitrait)    #load the dataset
-qtls <- scanone(multitrait, pheno.col=1:24)[,-c(1,2)]
-ctls <- CTLscan.cross(multitrait, pheno.col=1:2, have.qtl=qtls)
+oneqtls   <- scanone(multitrait, pheno.col=1:24)[,-c(1,2)]
+ctls_one  <- CTLscan.cross(multitrait, pheno.col=1:2, have.qtl=qtls)
 @
 
-, when no qtls are supplied the internal QTL mapping routine is used, which can deal with just a 
-single condition per individual. An environmental condition can be specified by the $conditions$ 
-param and are also passed to the internal QTL mapping algorithm.
+And a more advanced example where the MQM scan uses 30 genetic markers as cofactors to first build 
+up a genetic model underlying the QTL and locus to locus interactions:
+
+<<echo=TRUE,eval=FALSE>>=
+cofactors <- mqmautocofactors(multitrait, 30)
+mqm_qtls  <- mqmscanone(multitrait, cofactors=cofactors, pheno.col=1:24)[,-c(1,2)]
+ctls_mqm  <- CTLscan.cross(multitrait, pheno.col=1:2, have.qtl=qtls)
+@
+
+CTLs can be condition dependant just like QTL, so if one has multiple conditions or treatments. we 
+need to acomodate that in the underlying models. If no QTLs are supplied the internal QTL mapping 
+routine is used, which can deal with just a single condition per individual. An environmental 
+condition can be specified by the $conditions$ parameter:
 
@@
 \bigskip
-\textbf{\sffamily Advanced plots and CTL networks}
+\textbf{\sffamily RESULTS - Plots and CTL networks}
 \nopagebreak
@@
 
-Explains how to use the D version to map big datsets
+Big datsets require special care, R isn't the best platform to deal with large scale genomics data 
+from micro- or tilling arrays (Just to mention RNAseq and full DNA strand sequencing). Thus we also
+provide a commandline version of out CTL mapping tool: mapctl.exe, for more information on all the 
+command line parameters can be found at http://www.mapctl.org/mapctl.html
+The mapctl commandline tool uses the same input files as the R version, and supports qtab the new 
+genotype and phenotype encoding scheme used by qtlHD.
+It provides only scanning and permutation features, the output is analyzed in R as outlined above. 
+To load the results from a mapctl scan into R by specifying the input files and the output directory 
+created by the tool using:
+
+<<echo=TRUE,eval=FALSE>>=
+ctls <- read.dctl("genotypes.csv","phenotypes.csv",results="<PATH>/output/")
+@
 

### 2012-05-10 Rctl/inst/doc/manual.Rnw
[Rctl] More manual updates

@@
 {\color{blue} \tt \bfseries University of Groningen \\
-Groningen Bioinformatics centre \& GCC \\
-revision \# 1 \\[24pt]
+Groningen Bioinformatics Centre \& GCC \\
+Revision \# 1 \\[24pt]
 XX Okt 2011}
@@
 \vfill
-
-\noindent \emph{email}: \verb|Danny.Arends@Gmail.com|
+\noindent \verb|Software:      http://www.mapctl.org/download.html|\\
+\noindent \verb|Manual:        http://www.mapctl.org/manual.html|\\
+\noindent \verb|Source-code:   http://www.github.com/DannyArends/ctl|\\
+\noindent \verb|E-mail:        Danny.Arends@Gmail.com|
 \newpage
@@
 
-\newpage
+\bigskip
 
@@
 not reformat your data in both the genotype.csv and the phenotype.csv files to match the 
-format below. {\color{red} \emph{\bf{NOTE: Individuals must match between the genotype and 
-the phenotype file...}}} The phenotypes.csv file containing the genotype matrix is stored 
-individuals x genetic marker:
+format below.
+{\color{red} \emph{\bf{ \\\\NOTE: Individual order must match between the genotype and the phenotype file... \\\\}}} 
+The phenotypes.csv file containing the genotype matrix is stored individuals x genetic marker:
 
@@
 detected differential correlation to LOD score matrices.
-As an example we show the code to scan the multitrait dataset provided in R/qtl:
+As an example we show the code to load the internal multitrait dataset provided in R/qtl, load
+your own by using the $read.cross()$ function:
 
@@
 multitrait          #Print basic dataset information
+?read.cross         #List of formats supported by R/qtl
 @
@@
 chromosome number, "cM" - the location of the marker in centiMorgans and the 3rd column "Mbp" - The 
-location of the marker in Mega basepairs. The structure of the $mapinfo.csv$ file:
+location of the marker in Mega basepairs.
+{\color{red} \emph{\bf{ \\\\NOTE: Marker order must match with the genotype file... \\\\}}} 
+The structure of the $mapinfo.csv$ file:
 
@@
 mapinfo <- read.csv("mapinfo.csv",row.names=1,col.names=TRUE)
-traits[1:5,1:3]     #Show the first 5 marker records
+mapinfo[1:5,1:3]     #Show the first 5 marker records
 @
@@
 
-We start explaining how to map CTL using the basic options of the $CTLscan$ function. This function 
-scan CTL and produces an output "CTLobject".
+We start explaining how to map CTL using the basic options of the $CTLscan$ function. 
+This function scan CTL and produces an output "CTLobject".
 <<echo=TRUE,eval=FALSE>>=
@@
 @
-If you have loaded an R/qtl "cross" object, use the $CTLscan.cross$ interface, this will automagically 
-extract the cross, check if the type of "cross" is supported, and deal with the $geno.enc$ parameter:
+If you have loaded an R/qtl "cross" object, use the $CTLscan.cross$ interface, this will 
+automagically extract the cross, check if the type of "cross" is supported, and deal 
+with the $geno.enc$ parameter:
 <<echo=TRUE,eval=FALSE>>=
@@
 parameter, which should be set to as many as possible, especially when publishing any results, 
-but for a quick scan 100 is ussualy sufficient.
+but for a quick scan 100 is ussualy sufficient:
+<<echo=TRUE,eval=FALSE>>=
+ctls_quick_scan      <- CTLscan(geno,traits,n.perm=100)
+ctls_for_publication <- CTLscan(geno,traits,n.perm=15000)
+@
 
@@
 
+\emph{CTLnetwork and cytoscape}\\
+The CTLnetwork function outputs CTL and (optional) QTL data from the CTLobject into a network 
+format, if mapinfo is available it is used. The network format is relatively easy and can also
+be parsed by many other programs. As an example we use cytoscape, but first lets generate some 
+network input files:
+
+<<echo=TRUE,eval=FALSE>>=
+ctl_network <- CTLnetwork(ctls, lod.threshold = 5, mapinfo)
+@
+
+The CTL edges in the network\_full.sif file are annotated as:
+
+\begin{tabular}{lrrrrr}
+Trait & CTL & Trait & LOD Score\\
+\end{tabular}
+
+The CTL edges in the network\_summary.sif file look like this:
+
+\begin{tabular}{lrrrrr}
+Trait & CTL & Trait & avgLOD Score & sumLOD Score & nTraits\\
+\end{tabular}
+
+Additional QTLs are added to both files in the folowing structure:
+
+\begin{tabular}{lrrr}
+Trait & QTL & Marker & LOD Score\\
+\end{tabular}
+
+This allows the first additional infromation column (Nr 4) to act as an distance measurement, 
+the closer two traits or two markers are together in the network the stronger a detected QTL 
+or a CTL.
+
 \bigskip
@@
 from micro- or tilling arrays (Just to mention RNAseq and full DNA strand sequencing). Thus we also
-provide a commandline version of out CTL mapping tool: mapctl.exe, for more information on all the 
-command line parameters can be found at http://www.mapctl.org/mapctl.html
+provide a commandline version of our CTL mapping routine: mapctl.exe, for more information on all 
+the command line parameters can be found at http://www.mapctl.org/mapctl.html
 The mapctl commandline tool uses the same input files as the R version, and supports qtab the new 
-genotype and phenotype encoding scheme used by qtlHD.
-It provides only scanning and permutation features, the output is analyzed in R as outlined above. 
-To load the results from a mapctl scan into R by specifying the input files and the output directory 
-created by the tool using:
+genotype and phenotype encoding scheme used by qtlHD. It provides only QTL and CTL scanning and 
+permutation features, the output is analyzed in R as outlined above. To load the results from a 
+mapctl scan into R by specifying the input files and the output directory created by the tool 
+using:
 
@@
 Explains how to determine CTL significance using the CTLpermute function.
+
+\bigskip
+\textbf{\sffamily Some famous last words}
+\nopagebreak
 

### 2012-05-10 Rctl/inst/doc/manual.Rnw
[Rctl] More manual updates

@@
 
-\textbf{\sffamily Abstract}: Correlated Trait Loci (CTL) mapping using the R package 'ctl'
+\textbf{\sffamily Abstract}: Tutorial for the Correlated Trait Loci (CTL) mapping to reconstruct
+genetic regulatory networks using the R package 'ctl' and the mapctl commandline tool. This 
+tutorial is targetted at people with a basic understanding of genetics and would like to analyse
+there inbred cross or natural population. Main focus will be explaining the functionality of the 
+software, and a how-to on loading datasets via Excel, tab delimited files (.csv) and R/qtl. In 
+the methodology and background only a short introduction in CTL mapping theory is given.
+
 \vfill
@@
 \emph{CTLscan options}\\
-The genotype matrix and the $geno.enc$ parameter, the phenotype matrix, the important $n.perms$ 
-parameter, which should be set to as many as possible, especially when publishing any results, 
-but for a quick scan 100 is ussualy sufficient:
+The genotype matrix and the $geno.enc$ parameter, if data is of a different encoding then the 
+standard 1 / 2 it can be specified by using the $geno.enc$ parameter.
+
+<<echo=TRUE,eval=FALSE>>=
+#Load a dataset with "AA" and "AB" genotypes
+ctls <- CTLscan(geno,traits,geno.enc=c("AA","AB"))
+@
+
+The important $n.perms$ parameter, which should be set to as many as possible, especially 
+when publishing any results, but for a quick the default of 100 is sufficient:
 <<echo=TRUE,eval=FALSE>>=
@@
 ctls_for_publication <- CTLscan(geno,traits,n.perm=15000)
+@
+
+CTLs could be condition dependant just like QTL, so if one has multiple conditions or 
+treatments. we need to accomodate that in the underlying models. 
+{\color{red} \emph{\bf{ \\\\NOTE: Current version doesn't handle conditions (only the QTL are compensated). \\\\}}} 
+If no QTLs are supplied the internal QTL mapping routine is used, which can deal with just a single condition per 
+individual. An environmental condition can be specified by the $conditions$ parameter:
+
+<<echo=TRUE,eval=FALSE>>=
+#Specify that individuals 1:51 were in condition A / e.g. treatment A
+#Specify that individuals 52:100 were in condition B / e.g. treatment B 
+#Specify that individuals 101:162 were in condition C / e.g. control
+conditions <- c(rep(1,51),rep(2,49),rep(2,62))
+ctls <- CTLscan(geno, traits, conditions=conditions)
 @
@@
 cofactors <- mqmautocofactors(multitrait, 30)
-mqm_qtls  <- mqmscanone(multitrait, cofactors=cofactors, pheno.col=1:24)[,-c(1,2)]
+mqm_qtls  <- mqmscan(multitrait, cofactors=cofactors, pheno.col=1:24)[,-c(1,2)]
 ctls_mqm  <- CTLscan.cross(multitrait, pheno.col=1:2, have.qtl=qtls)
-@
-
-CTLs can be condition dependant just like QTL, so if one has multiple conditions or treatments. we 
-need to acomodate that in the underlying models. If no QTLs are supplied the internal QTL mapping 
-routine is used, which can deal with just a single condition per individual. An environmental 
-condition can be specified by the $conditions$ parameter:
-
-<<echo=TRUE,eval=FALSE>>=
-#Specify that individuals 1:81 were in condition A / e.g. treatment
-#Specify that individuals 82:162 were in condition B / e.g. control
-conditions <- c(rep(1,81),rep(2,81))
-ctls <- CTLscan(geno, traits, conditions=conditions)
 @

### 2012-05-10 Rctl/inst/doc/manual.Rnw
[Rctl] More manual updates

@@
 
-\textbf{\sffamily Abstract}: Tutorial for the Correlated Trait Loci (CTL) mapping to reconstruct
-genetic regulatory networks using the R package 'ctl' and the mapctl commandline tool. This 
-tutorial is targetted at people with a basic understanding of genetics and would like to analyse
-there inbred cross or natural population. Main focus will be explaining the functionality of the 
-software, and a how-to on loading datasets via Excel, tab delimited files (.csv) and R/qtl. In 
-the methodology and background only a short introduction in CTL mapping theory is given.
+\textbf{\sffamily Abstract}: Tutorial for the Correlated Trait Loci (CTL) mapping package to 
+reconstruct genetic regulatory networks using the R package 'ctl' and the mapctl commandline 
+tool. 
+This tutorial is targetted at people with a basic understanding of genetics and would like to 
+analyse there inbred cross or natural population. Main focus will be explaining the functionality 
+of the software, and a how-to on loading datasets via Excel, tab delimited files (.csv) and 
+R/qtl. In the methodology and background only a short introduction in CTL mapping theory is 
+given.
 

### 2012-05-10 Rctl/inst/doc/manual.Rnw
[Rctl] Manual summary

@@
 This tutorial is targetted at people with a basic understanding of genetics and would like to 
-analyse there inbred cross or natural population. Main focus will be explaining the functionality 
-of the software, and a how-to on loading datasets via Excel, tab delimited files (.csv) and 
-R/qtl. In the methodology and background only a short introduction in CTL mapping theory is 
-given.
+analyse there inbred cross (RIL, DH, BC) or any population using SNP markers. Main focus will 
+be explaining the functionality of the software, and a how-to on using datasets currently in 
+Excel, tab delimited files (.csv) or use $read.cross$ from R/qtl. In the methodology and 
+background only a short introduction in CTL mapping theory is given.
 

### 2012-05-11 Rctl/inst/doc/manual.Rnw
[Rctl] manual updates

@@
 {\color{red} \emph{\bf{ \\\\NOTE: Individual order must match between the genotype and the phenotype file... \\\\}}} 
-The phenotypes.csv file containing the genotype matrix is stored individuals x genetic marker:
+The genotypes.csv file containing the genotype matrix is stored individuals x genetic marker:
 
@@
 #Specify that individuals 101:162 were in condition C / e.g. control
-conditions <- c(rep(1,51),rep(2,49),rep(2,62))
+conditions <- c(rep(1,51),rep(2,49),rep(3,62))
 ctls <- CTLscan(geno, traits, conditions=conditions)

### 2012-05-11 Rctl/R/ctl.network.R
[Rctl] Modified the code to use mapinfo structure

@@ -8,62 +8,81 @@
 # Network routines for CTL analysis
 #
 
-write.marker.attributes <- function(QTLscan, chr.edges){
-  chr   <- 1
-  namez <- colnames(QTLscan)
-  cat("", file="node_attributes.txt")
-  #Connections between genetic markers
-  for(m1 in 1:(length(namez)-1)){
-    if(!missing(chr.edges)){
-      cat(paste(namez[m1],"\tCHR",chr,"\n",sep=""),file="node_attributes.txt",append=TRUE)
+write.marker.attributes <- function(QTLscan, mapinfo){
+  if(missing(QTLscan)) stop("argument 'CTLobject' is missing, with no default")
+  if(!missing(chr.edges)){
+    chr          <- 1
+    chr.edges    <- get.chr.edges(mapinfo)+.5
+    markernames  <- colnames(QTLscan)
+  
+    cat("", file="node_attributes.txt")
+    #Connections between genetic markers
+    for(m1 in 1:(length(markernames)-1)){
+      cat(paste(markernames[m1],"\tCHR",chr,"\n",sep=""),file="node_attributes.txt",append=TRUE)
       if((((2*m1+1)/2) %in% chr.edges)){
         chr <- chr + 1
       }
     }
+    #Annotate the last marker on the last chromosome
+    cat(paste(markernames[length(markernames)],"\tCHR",chr,"\n",sep=""),file="node_attributes.txt",append=TRUE)
+    cat("Wrote attributes for",length(namez),"markers on",chr,"chromosomes\n")
+  }else{
+    cat("[WARNING] No nodeattributes.txt produced, Please supply a mapinfo file\n")
   }
-  #Annotate the last marker on the last chromosome
-  if(!missing(chr.edges)) cat(paste(namez[length(namez)],"\tCHR",chr,"\n",sep=""),file="node_attributes.txt",append=TRUE)
-  if(!missing(chr.edges)) cat("Wrote attributes for",length(namez),"markers on",chr,"chromosomes\n")
 }
 
-QTLnetwork <- function(CTLobject, lod.threshold = 5, chr.edges, filename){
-  idx   <- 1
-  edges <- 0
-  chr   <- 1
-  write.marker.attributes(attr(CTLobject,"qtl"), chr.edges)
-  if(missing(filename)){
-    filename <- "network_qtl.sif"
[…]

### 2012-05-11 Rctl/R/helper.functions.R

@@ -12,10 +12,13 @@
 # - Chromosome edge locations from mapfile
 # - Get the top-correlated metabolites
 
-ctl.name      <- function(CTLscan){ return(attr(CTLscan$ctl,"name")); }
-ctl.names     <- function(CTLobject){ unlist(lapply(CTLobject,function(x){return(attr(x$ctl,"name"));})) }
-ctl.ctlmatrix <- function(CTLscan){ return(CTLscan$ctl); }
-ctl.lodmatrix <- function(CTLscan){ return(CTLscan$l); }
+ctl.names      <- function(CTLobject){ unlist(lapply(CTLobject,function(x){return(attr(x$ctl,"name"));})) }
+ctl.qtlmatrix  <- function(CTLobject){ return(attr(CTLobject,"qtl")); }
+
+ctl.name       <- function(CTLscan){ return(attr(CTLscan$ctl,"name")); }
+ctl.ctlmatrix  <- function(CTLscan){ return(CTLscan$ctl); }
+ctl.qtlprofile <- function(CTLscan){ return(CTLscan$qtl); }
+ctl.lodmatrix  <- function(CTLscan){ return(CTLscan$l); }
 
 remove.diag <- function(x){ return(x*lower.tri(x) + x*upper.tri(x)); }
 up <- function(){abs(seq(-2,-0,0.1))/2} 
@@ -23,8 +26,8 @@ dw <- function(){seq(0.1,2,0.1)/2}
 redblue <- function(){c(rgb(up,0,0), rgb(0,0,dw))}
 whiteblack <- function(){c("white",gray.colors(10)[10:1])}
 
-get.chr.edges <- function(map_info){
-  unlist(lapply(unique(map_info[,1]),function(x){max(which(map_info[,1]==x));}))
+get.chr.edges <- function(mapinfo){
+  unlist(lapply(unique(mapinfo[,1]),function(x){max(which(mapinfo[,1]==x));}))
 }
 
 top.correlated <- function(x){

### 2012-05-11 Rctl/R/ctl.network.R
[Rctl] Network output changes

@@ -25,7 +25,7 @@ write.marker.attributes <- function(QTLscan, mapinfo){
     }
     #Annotate the last marker on the last chromosome
     cat(paste(markernames[length(markernames)],"\tCHR",chr,"\n",sep=""),file="node_attributes.txt",append=TRUE)
-    cat("Wrote attributes for",length(namez),"markers on",chr,"chromosomes\n")
+    cat("Wrote attributes for",length(markernames),"markers on",chr,"chromosomes\n")
   }else{
     cat("[WARNING] No nodeattributes.txt produced, Please supply a mapinfo file\n")
   }
@@ -33,24 +33,28 @@ write.marker.attributes <- function(QTLscan, mapinfo){
 
 QTLnetwork <- function(CTLobject, mapinfo, lod.threshold = 5, filename){
   if(missing(CTLobject)) stop("argument 'CTLobject' is missing, with no default")
-  if(missing(filename))  cat("[WARNING] Using default filename:","network_qtl.sif","\n")
+  if(!missing(mapinfo)){
+    chr.edges   <- get.chr.edges(mapinfo)+.5
+    write.marker.attributes(qtls, mapinfo)
+  }
+  if(missing(filename)){ 
+    filename <- "network_qtl.sif" 
+    cat("[WARNING] Using default filename:",filename,"\n")
+  }
   idx         <- 1
   edges       <- 0
   ncons       <- 0
   chr         <- 1
-  chr.edges   <- get.chr.edges(mapinfo)+.5
   qtls        <- ctl.qtlmatrix(CTLobject)
   markernames <- colnames(qtls)
   traitnames  <- rownames(qtls)
 
-  write.marker.attributes(qtls, mapinfo)
-
-  if(missing(filename)) filename <- "network_qtl.sif"
+  if(missing(filename)) 
   cat("",file=filename)
   #Print the CHR interconnections between genetic markers
   
   for(m1 in 1:(length(markernames)-1)){
-    if(missing(chr.edges)){
+    if(missing(mapinfo)){
       cat(paste(markernames[m1],"\tCHR\t",markernames[m1+1],"\t",1,"\n",sep=""),file=filename,append=TRUE)
       ncons <- ncons+1
     }else{

### 2012-05-11 Rctl/inst/doc/SweaveIt.R
[inst] Removed 4 files that were not supposed to be there

@@ -1,3 +1,0 @@
-Stangle(file="article.Rnw")
-Sweave(file="article.Rnw",pdf=FALSE)
-q("no")

### 2012-10-14 Rctl/R/ctl.scan.R
Reintroduced the have.qtls parameter

@@ -9,7 +9,7 @@
 #
 
 #-- CTLscan main function --#
-CTLscan <- function(genotypes, phenotypes, pheno.col = 1:ncol(phenotypes), n.perms=100, conditions = NULL, n.cores = 1, geno.enc = c(1,2), verbose = FALSE){
+CTLscan <- function(genotypes, phenotypes, pheno.col = 1:ncol(phenotypes), n.perms=100, conditions = NULL, have.qtls = NULL, n.cores = 1, geno.enc = c(1,2), verbose = FALSE){
   if(missing(genotypes)) stop("argument 'genotypes' is missing, with no default")
   if(missing(phenotypes)) stop("argument 'phenotypes' is missing, with no default")
   st <- proc.time()
@@ -27,7 +27,9 @@ CTLscan <- function(genotypes, phenotypes, pheno.col = 1:ncol(phenotypes), n.per
   }
   idx <- 1
   for(phe in pheno.col){
-    results[[idx]] <- CTLmapping(genotypes, phenotypes, pheno.col=phe, n.perms = n.perms, geno.enc=geno.enc, verbose=verbose)
+    qtl <- NULL
+    if(!is.null(have.qtls) && idx < ncol(have.qtls)) qtl <- have.qtls[,idx]
+    results[[idx]] <- CTLmapping(genotypes, phenotypes, pheno.col=phe, n.perms = n.perms, has.qtl = qtl, geno.enc=geno.enc, verbose=verbose)
     attr(results[[idx]],"name") <- colnames(phenotypes)[phe]
     class(results[[idx]]) <- c(class(results[[idx]]),"CTLscan")
     idx <- idx + 1
@@ -37,15 +39,19 @@ CTLscan <- function(genotypes, phenotypes, pheno.col = 1:ncol(phenotypes), n.per
   results
 }
 
-CTLmapping <- function(genotypes, phenotypes, pheno.col=1, n.perms=100, geno.enc=c(1,2), verbose = FALSE){
+CTLmapping <- function(genotypes, phenotypes, pheno.col=1, n.perms=100, has.qtl = NULL, geno.enc=c(1,2), verbose = FALSE){
   if(missing(genotypes)) stop("argument 'genotypes' is missing, with no default")
   if(missing(phenotypes)) stop("argument 'phenotypes' is missing, with no default")
   n.ind = nrow(genotypes); n.mar = ncol(genotypes); n.phe = ncol(phenotypes)
[…]

### 2012-10-14 Rctl/R/ctl.image.R
More manuals updated

@@ -8,14 +8,14 @@
 # Image plot routines for CTL analysis
 #
 
-image.CTLobject <- function(x, against = c("markers","phenotypes"), significance = 0.05, col=whiteblack, do.grid=TRUE, grid.col = "black", verbose = FALSE, add=FALSE, breaks = c(0, 1, 2, 3, 10, 10000), marker_info, ...){
+image.CTLobject <- function(x, against = c("markers","phenotypes"), significance = 0.05, col=whiteblack, do.grid=TRUE, grid.col = "white", verbose = FALSE, add=FALSE, breaks = c(0, 1, 2, 3, 10, 10000), marker_info, ...){
   colorrange <- col
   mymatrix <- CTLprofiles(x, against, significance)
   mainlabel <- paste("CTL phenotypes vs",against[1],"at P-value <",significance)
   internal.image(mymatrix, colorrange, mainlabel,do.grid, grid.col, breaks=breaks, marker_info=marker_info)
 }
 
-qtlimage <- function(CTLscan, do.grid = TRUE, grid.col = "black", verbose = FALSE){
+qtlimage <- function(CTLscan, do.grid = TRUE, grid.col = "white", verbose = FALSE){
   colorrange <- c("white",gray.colors(10)[10:1])
   mymatrix <- NULL
   mynames <- NULL
@@ -61,13 +61,6 @@ addChromosomeLines <- function(marker_info, col='black'){
     axis(1,at=(chr_start[x]+chr_ends[x])/2,paste("Chr",unique(marker_info[,1])[x]),cex.axis=0.3)
   }
   box()
-}
-
-QTLimage <- function(x, do.grid=TRUE, grid.col = "black", verbose = FALSE, ...){
-  colorrange <- c("white",gray.colors(10)[10:1])
-  mymatrix <- attr(x,"qtl")
-  mainlabel <- paste("QTL heatmap")
-  internal.image(mymatrix, colorrange, mainlabel,do.grid, grid.col)
 }
 
 # end of ctl.image.R

### 2012-10-14 Rctl/R/ctl.plot.R
Add the G. Churchill Bay-0 x Sha dataset with 9 metabolites as .RData

@@ -117,6 +117,7 @@ plot.CTLscan <- function(x, addQTL = TRUE, onlySignificant = TRUE, significance 
   }
   CTLmatrix <- matrix(x$ctl[,mysign],nrow(x$ctl),length(mysign))
   summarized <- apply(CTLmatrix,1,sum)
+  x$qtl[is.infinite(x$qtl)] <- max(x$qtl[is.finite(x$qtl)])
   plot(c(0,nrow(x$ctl)),c(min(c(summarized,x$qtl)),max(c(5,summarized,x$qtl))), type='n',xlab="Marker", ylab="-log10(P-value)", main=paste("Phenotype contribution to CTL of",ctl.name(x)), ...)
   p <- rep(0,nrow(x$ctl))
   i <- 1;

### 2012-10-14 Rctl/R/ctl.load.R
Minor updates to headers in R and manuals

@@ -1,7 +1,7 @@
 #
 # ctl.load.R
 #
-# copyright (c) 2012 Danny Arends, Bruno Tesson and Ritsert C. Jansen
+# copyright (c) 2010-2012 - GBIC, Danny Arends, Bruno Tesson and Ritsert C. Jansen
 # last modified Jul, 2012
 # first written Jun, 2012
 # 

### 2012-10-14 Rctl/R/ctl.network.R

@@ -1,19 +1,19 @@
 #
 # ctl.network.R
 #
-# copyright (c) 2010 Danny Arends and Bruno Tesson
-# last modified Jan, 2012
+# copyright (c) 2010-2012 - GBIC, Danny Arends, Bruno Tesson and Ritsert C. Jansen
+# last modified Oct, 2012
 # first written Oct, 2011
 # 
 # Network routines for CTL analysis
 #
 
-write.marker.attributes <- function(QTLscan, mapinfo){
-  if(missing(QTLscan)) stop("argument 'CTLobject' is missing, with no default")
+write.marker.attributes <- function(CTLobject, mapinfo){
+  if(missing(CTLobject)) stop("argument 'CTLobject' is missing, with no default")
   if(!missing(mapinfo)){
     chr.edges    <- get.chr.edges(mapinfo)+.5
     chr          <- 1
-    markernames  <- colnames(QTLscan)
+    markernames  <- colnames(CTLobject)
   
     cat("", file="node_attributes.txt")
     #Connections between genetic markers

### 2012-10-14 Rctl/R/ctl.plot.R

@@ -1,8 +1,8 @@
 #
 # ctl.plot.R
 #
-# copyright (c) 2010 Danny Arends and Bruno Tesson
-# last modified Jan, 2012
+# copyright (c) 2010-2012 - GBIC, Danny Arends, Bruno Tesson and Ritsert C. Jansen
+# last modified Oct, 2012
 # first written Nov, 2010
 # 
 # Plotting routines for CTL analysis

### 2012-10-14 Rctl/R/ctl.print.R

@@ -1,8 +1,8 @@
 #
 # ctl.print.R
 #
-# copyright (c) 2010-2011 Danny Arends and Ritsert C. Jansen
-# last modified Jan, 2012
+# copyright (c) 2010-2012 - GBIC, Danny Arends, Bruno Tesson and Ritsert C. Jansen
+# last modified Oct, 2012
 # first written Oct, 2011
 # 
 # Print routines for CTL analysis
@@ -47,6 +47,7 @@ CTLsignificant <- function(CTLobject, significance = 0.05, what = c("names","ids
 }
 
 print.CTLscan <- function(x, ...){
+  if(missing(x)) stop("argument 'x' is missing, with no default")
   cat("CTLscan summary",attr(x$ctl,"name"),"\n\n")
   cat("- Number of background phenotypes",dim(x$ctl)[1],"\n")
   cat("- Number of markers",dim(x$ctl)[2],"\n")

### 2012-10-14 Rctl/R/ctl.profiles.R

@@ -1,14 +1,14 @@
 #
 # ctl.profiles.R
 #
-# copyright (c) 2011 Danny Arends and Ritsert C. Jansen
-# last modified Feb, 2012
+# copyright (c) 2010-2012 - GBIC, Danny Arends, Bruno Tesson and Ritsert C. Jansen
+# last modified Oct, 2012
 # first written Oct, 2010
 # 
-#
 
 #Create the 2 possible CTL matrices: 1) phenotypes versus markers (PxM) and 2) Phenotypes versus phenotypes (PxP)
 CTLprofiles <- function(CTLobject, against = c("markers","phenotypes"), significance=0.05, verbose=FALSE, warn = TRUE){
+  if(missing(CTLobject)) stop("argument 'CTLobject' is missing, with no default")
   mymatrix <- NULL
   mynames <- NULL
   warn <- warn

### 2012-10-14 Rctl/R/ctl.statistics.R

@@ -1,14 +1,13 @@
 #
 # ctl.statistics.R
 #
-# copyright (c) 2010 Danny Arends, Bruno Tesson and Ritsert C. Jansen
-# last modified Jan, 2012
+# copyright (c) 2010-2012 - GBIC, Danny Arends, Bruno Tesson and Ritsert C. Jansen
+# last modified Oct, 2012
 # first written Nov, 2011
 # 
 # R functions to do transform CTL mapping scores to Pvalues and LOD
 # Example data C. Elegans and available at request ( Danny.Arends@gmail.com )
 #
-
 
 #Use the top 10% of permutation scores and fit a GPD uppertail distribution, then use the 
 #GPD to obtain P-values for outliers, If the GPD estimates a 0 p-value we reduce our value 

### 2013-04-19 Rctl/R/ctl.load.R
Fixing issue #11, also restoring the QTAB input to D version

@@ -5,7 +5,7 @@
 # last modified Jul, 2012
 # first written Jun, 2012
 # 
-# R functions to do load CTL mapping donw with the D version
+# R functions to do load CTL mapping done with the D version
 #
 
 #-- ctl.load function --#

### 2013-04-19 Rctl/inst/doc/manual.Rnw
[Manual] Minor updates
Add Sweave.sty and minor updates to the manual
Fixing issue #13, the D/test/data files are now used in the manual also

@@
 \begin{tabular}{lrrrrr}
-      & Marker1 & Marker2 & Marker3 & ... & MarkerN \\
-Ind1  & 1 & 1 & 1 & ... & 2 \\
-Ind2  & 1 & 2 & 2 & ... & 1 \\
-Ind3  & 2 & 2 & 1 & ... & 1 \\
-...   & ... & ... & ... & ... & ... \\
-IndN  & 1 & 1 & 1 & ... & 2 \\
+     & Chromosome & Location & Ind1 & Ind2 & Ind3 & ... & IndN \\
+Marker1  & 1 & 0.0 & 1 & 1 & 1 & ... & 2 \\
+marker2  & 1 & 5.3 & 1 & 2 & 2 & ... & 1 \\
+marker3  & 1 & 7.8 & 2 & 2 & 1 & ... & 1 \\
+...   & ... & ... & ... & ... & ... & ... & ... \\
+MarkerN  & 7 & 120.1 & 1 & 1 & 1 & ... & 2 \\
 \end{tabular}
@@
 \begin{tabular}{lrrrrr}
-      & Trait1 & Trait2 & Trait3 & ... & TraitN \\
-Ind1  & 5.8    & 11.0   & 9.0    & ... & 1.6 \\
-Ind2  & 6.3    & 12.2   & NA     & ... & 1.3 \\
-Ind3  & 5.1    & 11.1   & 12.3   & ... & 2.0 \\
-...   & ...    & ...    & ...    & ... & ... \\
-IndN  & 9.8    & 15.8   & 23.0   & ... & 3.4 \\
+      & Chromosome & Location & Ind1 & Ind2 & Ind3 & ... & IndN \\
+Trait1  & ... & ... & 5.8    & 11.0   & 9.0    & ... & 1.6 \\
+Trait2  & ... & ... & 6.3    & 12.2   & NA     & ... & 1.3 \\
+Trait3  & ... & ... & 5.1    & 11.1   & 12.3   & ... & 2.0 \\
+...   & ... & ... & ...    & ...    & ...    & ... & ... \\
+Trait  & ... & ... & 9.8    & 15.8   & 23.0   & ... & 3.4 \\
 \end{tabular}
@@
 <<echo=TRUE,eval=FALSE>>=
-setwd("<PATH_TO_DATA>")
-genotypes <- read.csv("genotypes.csv",row.names=1,col.names=TRUE)
-traits <- read.csv("traits.csv",row.names=1,col.names=TRUE)
-genotypes[1:5,1:10]    #Show the first 5 individuals, 10 markers
-traits[1:5,1:10]       #Show the first 5 individuals, 10 traits
+  setwd("/Path/To/Data/")
+  genotypes <- read.csv("genotypes.csv",row.names=1, header=FALSE,sep="\t")
+  traits <- read.csv("phenotypes.csv",row.names=1, header=FALSE,sep="\t")
+  genotypes[1:5,1:10]    #Show 5 individuals, 8 markers, and genetic locations
+  traits[1:5,1:10]       #Show 5 individuals, 8 traits, and genetic locations
+  mapinfo <- genotypes[,1:2]
+  genotypes <- genotypes[,-c(1,2)]  # Remove the genetic locations
+  traits <- traits[,-c(1,2)]        # Remove the genetic locations
 @
@@
 
+\bigskip
+\textbf{\sffamily Get a list of significant CTLs}
+\nopagebreak
+
+Get significant CTL interactions by using CTLsignificant.
+
+<<echo=TRUE,eval=FALSE>>=
+sign <- CTLsignificant(ctl_scan)
+@
+
 \emph{CTLnetwork and cytoscape}\\
@@
 \bigskip
-\textbf{\sffamily Significance using CTLpermute}
-\nopagebreak
-
-Explains how to determine CTL significance using the CTLpermute function.
-
-\bigskip
 \textbf{\sffamily Some famous last words}
@@
 
-\newpage
+\bigskip
 {\small

### 2016-03-22 Rctl/R/ctl.correlation.R
Minor updates to the code to remove some dependancies not available anymore

@@ -50,8 +50,6 @@ getCorrelations.cross <- function(cross, pheno.col = 1, marker.col = 1){
 }
 
 test.getCorrelation.cross <- function(){
-  library(qtl)
-  library(ctl)
   data(hyper)
   hyper$pheno <- matrix(runif(250*10), 250, 10)  # Create 10 phenotypes as a matrix
   colnames(hyper$pheno) <- paste0("Pheno", 1:10)
@@ -61,7 +59,6 @@ test.getCorrelation.cross <- function(){
 }
 
 test.correlation <- function(){
-  library(ctl)
   correlation(1:10, 1:10)
   correlation(10:1, 1:10)
   v <- runif(10)
@@ -73,7 +70,6 @@ test.correlation <- function(){
 }
 
 test.chiSQN <- function(){
-  library(ctl)
   ngeno <- 2
   correlations <- matrix(runif(20), 10, ngeno)  # 10 markers, with 2 x correlation per genotype
   nind <- c(20, 20)                             # 20 individuals in each group

### 2016-03-22 Rctl/R/ctl.power.R

@@ -134,8 +134,6 @@ CTLanalyseRatios <- function(n = 100, perms = 100, effect = 0.4, ind = 200, rati
 }
 
 test.power.test <- function(){
-  require(ctl)
-
   sample100 <- CTLanalyseEffects(effects=c(0.02, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.75), ind=100)
   sample200 <- CTLanalyseEffects(effects=c(0.02, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.75), ind=200)
   sample400 <- CTLanalyseEffects(effects=c(0.02, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.75), ind=400)

### 2016-03-22 Rctl/R/ctl.statistics.R

@@ -46,27 +46,6 @@ CTLsignificant <- function(CTLobject, significance = 0.05, what = c("names","ids
   return(all_sign)
 }
 
-#Use the top 10% of permutation scores and fit a GPD uppertail distribution, then use the 
-#GPD to obtain P-values for outliers, If the GPD estimates a 0 p-value we reduce our value 
-#to get the minimum P-value
-extrapolateBeyondRange <- function(permvalues, value = 0.6, top = 20){
-  require(POT)
-  gpd.threshold <- permvalues[(.80*length(permvalues))]
-  mle <- fitgpd(permvalues, gpd.threshold, "mle")
-  shape <- mle$param["shape"]
-  scale <- mle$scale
-  loc <- mle$threshold[1]
-  dens <- function(x) dgpd(x, loc, scale, shape)
-  warn <- FALSE
-  prev.value <- value
-  while(as.numeric(dens(value))==0 && value > 0){
-    #cat("[FIX] Out of range and p=0:", value," ",as.numeric(dens(value)),"\n")
-    warn <- TRUE
-    value <- value - 0.0001
-  }
-  as.numeric(dens(value))
-}
-
 CTLtoLODvector <- function(CTLscan, against = c("markers","phenotypes")){
   if(missing(CTLscan)) stop("argument 'CTLscan' is missing, with no default")
   if(against[1]=="markers")return(apply(CTLscan$ctl,1,sum))
